Steady-state neuron-predominant LINE-1 encoded ORF1p protein and LINE-1 RNA increase with aging in the mouse and human brain

Version 1

Abstract

Abstract

Recent studies have established a reciprocal causal link between aging and the activation of transposable elements, characterized in particular by a de-repression of LINE-1 retrotransposons. These LINE-1 elements represent 21% of the human genome, but only a minority of these sequences retain the coding potential essential for their mobility. LINE-1 encoded proteins can induce cell toxicity implicated in aging and neurodegenerative diseases. However, our knowledge of the expression and localization of LINE-1-encoded proteins in the central nervous system is limited. Using a novel approach combining atlas-based brain mapping with deep-learning algorithms on large-scale pyramidal brain images, we unveil a heterogeneous, neuron-predominant and widespread ORF1p expression throughout the murine brain at steady-state. In aged mice, ORF1p expression increases significantly which is corroborated in human post-mortem dopaminergic neurons by an increase in young LINE-1 elements including those with open reading frames. Mass spectrometry analysis of endogenous mouse ORF1p revealed novel, neuron-specific protein interactors. These findings contribute to a comprehensive description of the dynamics of LINE-1 and ORF1p expression in the brain at steady-state and in aging and provide insights on ORF1p protein interactions in the brain.

Introduction

Only about 2 % of the human genome are DNA sequences that will be translated into protein. The remaining 98 % are comprised of introns, regulatory elements, non-coding RNA, pseudogenes and repetitive elements including transposable elements. However, some sequences in what is generally considered “non-coding genome” do in fact contain sequences which encode proteins. This is true for specific lncRNAs which can encode peptides or functional proteins 1 but also for a few copies of two transposable element families, Long INterspersed Element-1 (LINE-1) and Human Endogenous RetroViruses (HERV). Non-functional copies of retrotransposons, to which LINE-1 and HERV belong, cover about 44% 2 of the human genome as remnants of an evolutionary ancient activity. Depending on the source, about 100 3 to 146 4 full-length LINE-1 elements with two open reading frames encoding ORF1 and ORF2 are present in the Human reference genome (GRCh38 Genome Assembly) and several incomplete HERV sequences encoding either or any combination of envelope (env), gag, pro or pol 5. The LINE-1 encoded protein ORF1p, an RNA binding protein with “cis” preference 6,7, and ORF2p, an endonuclease and reverse transcriptase 8,9 are required for the mobility of LINE-1 elements. As many other transposable elements (TEs) including HERVs, LINE-1 elements are repressed by multiple cellular pathways. It was thus generally thought that TEs are repressed in somatic cells with no expression at steady-state 10–12. However, the aging process reduces the reliability of these repressive mechanisms 13. It is now, 31 years after the initial proposition of the “transposon theory of aging” by Driver and McKechnie 14, generally accepted that TE activation can be both, a cause and a consequence of aging 15,16.

Sparse data has shown that the LINE-1 encoded protein ORF1p is expressed at steady-state in the mouse ventral midbrain 17, the mouse hippocampus 18 and in some regions of the human post-mortem brain 19 and recent data informed about the presence of full-length transcripts in cancer cells, human epithelial cells and mouse hippocampal neurons 20. Repression of LINE-1 might thus be incomplete and if so, it remains unclear how cells then prevent cell toxicity associated with LINE-1 encoded protein activity. Indeed, LINE-1 encoded proteins have been demonstrated to induce genomic instability (ORF2p endonuclease-mediated 17,21–26) and inflammation (ORF2p reverse transcriptase-mediated 27–29) and these cellular activities might be causally related to organismal aging, cancer, autoimmune and neurological diseases 30. For instance, LINE-1 activation can drive neurodegeneration of mouse dopaminergic neurons 17, of drosophila neurons 31,32 and of mouse Purkinje neurons 33 which can be at least partially rescued with nucleoside analogue reverse transcriptase inhibitors (NRTIs) or other anti-LINE-1 strategies. NRTIs are currently being tested in several clinical trials designed to target either the RT of HERVs or the RT encoded by the LINE-1 ORF2 protein. It is not known today, however, to which extend LINE-1 encoded proteins are expressed at steady-state throughout the mouse and human brain, whether there is cell-type specificity and whether activation of LINE-1 encoded proteins is associated with brain aging or human neurodegeneration. Here, using a deep-learning assisted cellular detection methodology applied to pyramidal large-scale images of the mouse brain mapped to the Allen mouse brain atlas combined with post-mortem human brain imaging, co-IP mass spectrometry and transcriptomic analysis of LINE-1 expression, we describe a brain-wide map of ORF1p expression and interacting proteins at steady-state and in the context of aging. We find a heterogenous but widespread expression of ORF1p in the mouse brain with predominant expression in neurons. In aged mice, neuronal ORF1p expression increases brain-wide and in some brain regions to up to 27%. In human dopaminergic neurons, young LINE-1 transcripts and specific full-length and coding LINE-1 copies are increased in aged individuals. We further describe endogenous mouse ORF1p interacting proteins revealing known interactors and unexpected interacting proteins belonging to GO categories related to RNA metabolism, chromatin remodeling, cytoskeleton and the synapse.

Results

Widespread and heterogenous expression of the LINE-1 encoded ORF1p protein in the wildtype mouse brain

To investigate the expression pattern and intensities of endogenous LINE-1 encoded ORF1p protein throughout the entire mouse brain, we devised a deep-learning assisted cellular detection methodology applied to pyramidal large-scale images using a comprehensive workflow complemented by an approach based on confocal imaging as schematized in Figure 1A. Briefly, starting from sagittal slide scanner images of the mouse brain, we defined anatomical brain regions by mapping the Allen Brain Atlas onto the slide scanner images using Aligning Big Brains & Atlases (ABBA). We then employed a deep-learning detection method to identify all cell nuclei (Hoechst) and categorize all detected cells into neuronal cells (NeuN+) or non-neuronal cells (NeuN-) and ORF1p-expressing cells (ORF1p+) or cells that do not express ORF1p (ORF1p-). This workflow allowed us then to characterize the cell identity of ORF1p+ cells and ORF1p intensity throughout the whole brain but also in specific anatomical regions. In parallel, we completed the approach using confocal microscopy on selected anatomical regions allowing for comparison with higher resolution. Importantly, the specificity of the ORF1p antibody, a widely used, commercially available antibody 18,34–38, was confirmed by blocking the ORF1p antibody with purified mouse ORF1p protein resulting in the complete absence of immunofluorescence staining (Suppl Fig. 1A), by using an in-house antibody against mouse ORF1p17 which colocalized with the anti-ORF1p antibody used (Suppl Fig. 1B, quantified in Suppl Fig. 1C), and by immunoprecipitation and mass spectrometry used in this study (see below). Unexpectedly, we found a generalized and widespread expression of ORF1p throughout the brain of wildtype mice (Fig. 1B; Swiss OF1 mice, three months-old; whole brain except regions with particularly high cellular density (cerebellum, hippocampus, olfactory bulb) which impedes nuclei detection by deep-learning. ORF1p is detectable in all regions and subregions analyzed with heterogenous expression patterns (density and intensity) per region/subregions. The ten regions shown in Figure 1B exemplify visible different densities of ORF1p+ cells with varying levels of expression. Notably, the expression pattern of ORF1p in the hippocampus is similar to what has recently been published 18 (Fig. 1B, panel 2). Throughout the entire brain, the mean density of ORF1p+ cells per mm² was ≈ 305 ±18 (mean ±SEM), representing up to 20% of all detected cells (Fig. 1C). ORF1p+ cells in each mouse brain analyzed showed up to eight-fold disparities in intensity between low- and high-expressed cells (Fig. 1C). We then quantified nine anatomical regions according to the Allen Brain Atlas on four brains of three-month old mice (Fig. 1D) using the automated workflow (Fig. 1A) with regard to cell density (Fig. 1D), cell proportions (Fig. 1E) and fluorescent intensity of ORF1p+ cells (Fig. 1F). This approach permitted the analysis of about 10 000 ORF1p+ cells per animal highlighting the power of our large-scale analysis. Densities of ORF1p+ cells ranged from the lowest density in the hindbrain with 154 ±19 cells per mm2 (mean ±SEM) to the highest density of ORF1p+ cells in the isocortex with 451 ±44 cells per mm2 (mean ±SEM) and the thalamus with 446 ±50 cells per mm2 (mean ±SEM). The proportion of ORF1p+ cells per anatomical region fluctuated between 10% ±2.1 (ventral striatum, mean ±SEM) and 31% ±1.6 (thalamus, mean ±SEM). The dorsal striatum (“striatal dorsal” in the Allen Brain Atlas denomination) exhibited the lowest ORF1p expression intensity (658 ±3 mean ±SEM) of all regions tested, the hindbrain the highest mean intensity of ORF1p per cell (mean ±SD 1221 ±548) as illustrated in Figure 1B and quantified in Figure 1D and 1F. Interestingly, cell density did not correlate with expression levels. Dorsal and ventral striatum for instance displayed similar ORF1p intensities per cell but exhibited significant differences in ORF1p cell density and proportion. The “midbrain motor” region as defined by the Allen Brain Atlas showed an intermediate cell density (mean ±SEM 265 ±16 cells per mm2) and a rather high ORF1p expression intensity (mean ±SEM 1006 ±533). Statistical analysis comparing mean density of ORF1p+cells per mm2 or mean intensity per ORF1p+ cells among regions confirmed the heterogeneity concerning ORF1p expression throughout the mouse brain (Fig. 1D, 1F). Slide scanner and confocal images revealed an exceptionally high ORF1p expression intensity in the ventral region of the midbrain which we identified as the Substantia nigra pars compacta (SNpc). This region displayed an important density of ORF1p+ cells and a comparatively high level of ORF1p expression as illustrated by confocal imaging (Fig. 1B, panel 8), but could not be quantified independently with our brain-wide approach due to the geometrically-complex anatomy of this region and its small size (subregion-level in the Allen Brain Atlas hierarchy). Another region which could not be included in our brain-wide analysis was the cerebellum due to its extremely high density of cell nuclei. However, slide scanner and confocal imaging (Fig. 1B, panel 10) revealed that ORF1p is expressed in Purkinje cells, while not detectable in the molecular or granular layers.

Widespread and heterogenous expression of ORF1p protein in the mouse brain

(A) Schematic representation of the unbiased cell detection pipeline on large scale and confocal images. Immunofluorescent images on sagittal mouse brain slices were acquired on a digital pathology slide-scanner or on a confocal microscope (DNA stain: Hoechst, neuronal marker; NeuN, protein of interest: ORF1p). Pyramidal images aquired with the slide scanner were then aligned with the hierarchical anatomical annotation of the Allen Brain Atlas using ABBA. Once the regions defined, a deep-learning based detection of cell nuclei (Hoechst staining, Stardist) and cell cytoplasm (NeuN staining, Cellpose) was performed on each sub-region of the atlas. Objects were filtered according to the background intensity measured in each sub-region for each channel (NeuN and ORF1p). The identity and intensity measures were analyzed at the regional and whole brain level. In parallel, confocal images (multiple z-stacks) of two selected regions (frontal cortex and ventral midbrain) were also acquired and identity and intensity were quantified using Cellpose and Stardist.

(B) Widespread and heterogenous expression of the LINE-1 encoded protein ORF1p in the mouse brain. Representative image of ORF1p immunostaining (orange) of a sagittal section of the brain of a young (three months-old) mouse acquired on a slide scanner. Scale bar = 1mm. (1-10) Representative images of immunostainings showing ORF1p expression (orange) in 10 different regions of the mouse brain acquired on a confocal microscope. Nuclei are represented in blue (Hoechst), scale bar = 50µm. (1) Isocortex, (2) Hippocampus, (3) Striatum dorsal, (4) Thalamus, (5) Midbrain motor, (6) Pallidum, (7) Hypothalamus, (8) Substantia nigra pars compacta, (9) Hindbrain, (10) Cerebellum. ORF1p expression profile in the mouse brain. The entire mouse brain with the exception of the olfactory bulb and the cerebellum were analyzed according to the pipeline on large-scale images described in (A). Bar plot showing the total number of ORF1p+ cells per mm² in the mouse brain. Data is represented as mean ±SEM, n=4 mice (top). Bar plot indicating the proportion of ORF1p+ cells compared to all cells detected. Data is represented as mean ±SEM, n=4 mice, 202001 total cells analyzed (middle). Scatter plot showing the mean intensity of ORF1p per ORF1p+ cell. Data is represented as mean ±SD, n=4 mice, 40999 ORF1p+ cells analyzed (bottom).

(D-F) ORF1p expression profile (density, proportion and expression) in defined anatomical regions of the mouse brain. Nine anatomical regions as defined by the Allen Brain Atlas and mapped onto sagittal brain slices (four three-month-old Swiss/ OF1) with ABBA were analyzed using the pipeline on large scale images described in (A). (D) ORF1p+ cell density in 9 different regions. Bar plot showing the number of ORF1p+ cells per mm². Data is represented as mean ±SEM; *p<0.05; **p<0.01; adjusted p-value, one-way ANOVA followed by a Benjamin-Hochberg test (E) Proportion of ORF1p positive cells in 9 different regions. Bar plot showing the proportion of ORF1p+ cells among all cells detected per region. Data is represented as mean ±SEM. (F) Mean ORF1p expression per cell in 9 different regions. Dot plot showing the mean intensity of ORF1p signal per ORF1p+ cell in 9 different regions. Data is represented as mean ±SD. The number of analyzed cells per region is indicated in the figure. *p<0.05; **p<0.01; ***p<0.001; ****p<0.0001; adjusted p-value, nested one-way ANOVA followed by Sidak’ multiple comparison test.

(G) ORF1p expression in the frontal cortex and ventral midbrain. Confocal images with multiple z-stacks were analyzed using Cellpose and Stardist. Dot plot representing the mean expression of ORF1p per ORF1p+ cells. Individual (four three-month-old Swiss/ OF1) mice are represented each by a different color, the scattered line represents the median. ****p<0.0001, nested one-way ANOVA. Total cells analyzed = 4645.

(H-I) ORF1p expression in the frontal cortex and the ventral midbrain. (H) Western blots showing ORF1p (top) and actin expression (bottom) in four individual mice per region which were quantified in (I) using actin as a reference control. The signal intensity is plotted as the fold change of ORF1p expression in the ventral midbrain to ORF1p expression in the frontal cortex. *p<0.05; two-sided, unpaired student’s-test.

(J) ORF1p expression in three regions of the human brain. Western blot showing ORF1p expression in the cingulate gyrus (CG), frontal cortex (FC) and cerebellum (CB) of post-mortem tissues from a healthy individual. ORF1p (Top), Actin (bottom).

In order to confirm ORF1p expression by an independent method, we performed Western blot analysis on six micro-dissected regions from the mouse brain (Swiss/OF1 mouse, three-month old). As shown in Suppl Fig. 1D, ORF1p is expressed in all six regions with varying expression levels confirming the overall presence of ORF1p throughout the brain. We then chose two regions with significantly divergent ORF1p expression intensities as identified and quantified on pyramidal large-scale images: the frontal cortex (low) and the ventral midbrain (intermediate to high). We confirmed a significant higher expression of ORF1p in the ventral midbrain compared to the frontal cortex using an approach based on the unbiased, automated quantification of multiple z-stacks using a confocal microscope (Fig. 1G) and by Western-blotting on micro-dissected regions (Fig. 1H,I). In concordance with the findings stemming from the large-scale image quantification pipeline (Fig. 1F), the ventral midbrain showed ≈ 2-times higher expression of ORF1p than the frontal cortex as quantified in Figure 1G (1.8-fold) and Figure I (2.3-fold) validating our cellular detection methodology for pyramidal large-scale imaging and underscoring the heterogeneity of ORF1p expression levels in the mouse brain. To investigate intra-individual expression patterns of ORF1p in the post-mortem human brain, we analyzed three brain regions of a neurologically-healthy individual (Fig. 1J) by Western blotting. ORF1p was expressed at different levels in the cingulate gyrus, the frontal cortex and the cerebellum underscoring a widespread expression of human ORF1p across the human brain. In summary, our findings reveal the consistent presence of ORF1p expression throughout the mouse brain in all anatomical regions analyzed with high regional variability in terms of density of ORF1p+ cells per mm2 per region and ORF1p+ cell intensity per region. This finding raises several questions concerning cell-type identity of ORF1p expressing cells and potential functions or consequences of ORF1p expression in the mouse and human brain at steady-state.

ORF1p is predominantly expressed in neurons

Following our observation of a wide-spread expression of endogenous ORF1p throughout the brain, we first addressed the question of the cellular identity of ORF1p+ cells. To this end, we used the neuron-specific marker NeuN, commonly used to identify post-mitotic neurons in the central nervous system 39. This allowed us to determine the proportion of neuronal (NeuN+) or non-neuronal cells (NeuN-) expressing ORF1p (ORF1p+) or not (ORF1p-). Making use of our large-scale imaging approach (Fig. 1A), we observed drastic dissimilarities in detected cellular proportions between the white and grey brain matter. As expected, we observed only 1% of NeuN+ cells in the white matter (corpus callosum; Fig. 2A) validating both, the neuronal marker NeuN as such and the ABBA superposition of the Allen Brain Atlas onto the sagittal brain slices. In the grey matter, our approach detected 30.5% NeuN+ cells (dark red and yellow bars in Fig. 2A) which, according to the literature, should include all post-mitotic neurons with only minor exceptions 39–42 and corresponds to the reported proportion of neurons present in the mouse brain 43. The nine identified grey matter regions in Fig. 2A display the proportions of the different cell types per region. The proportion of all cells in a given region which are positive for ORF1p (dark red bars) differed between regions (lowest proportion: hindbrain: 7%; highest proportion dorsal striatum: 26.6%). In the isocortex and the midbrain motor-related regions, the majority of neurons detected express ORF1p (54% and 59% by large-scale analysis, Suppl Fig. 2A; 68.7% and 68.8% by confocal imaging (Fig. 2B, quantified in C and Suppl Fig. 2B), respectively), while in the midbrain sensory related regions the proportion dropped to 25% whereas it reached 82% in the thalamus (Suppl Fig. 2A). Altogether, nearly half of all NeuN+ cells throughout the mouse brain expressed ORF1p (mean of all regions: 48.2%; Suppl Fig. 2A). Regarding the cell identity of ORF1p+ cells brain-wide, more than 70% were identified as neuronal by the large-scale approach (Suppl Fig. 2C). This contrasted somewhat with results obtained by the second approach using confocal imaging on multiple z-stacks which indicated that 91.3% (frontal cortex) and 88.5% (ventral midbrain) of ORF1p+ cells were neuronal (Fig. 2D). This difference in percentages of ORF1p+ expressing neurons among all neurons between the large-scale image cell detection methodology and the confocal workflow is most probably due to technical limitations inherent to our large-scale pipeline. Indeed, with the latter approach, region-dependent differences in cell density and signal intensity levels might be the cause for an underestimation of the proportion of ORF1p+ cells being neuronal due to difficulties in cell detection by StarDist/Cellpose (high cell density) on a single focal plan, technical difficulties which are widely reduced by the multiple z-stack based approach when using a confocal microscope. Notably, frontal cortex and ventral midbrain present similar proportion of neurons expressing ORF1p and ORF1p although the percentage of NeuN+ cells between these two regions is significantly different (Suppl Fig. 2D). As we could not rule out that ORF1p might also be expressed in non-neuronal cells, we turned to non-neuronal markers specific for different glial cell populations using two different astrocytic markers (GFAP, Sox9), the astro- and oligodendrocytic marker S100β and the microglial marker Iba143,44 and performed co-staining with ORF1p followed by confocal imaging as illustrated in Figure 2E. We screened multiple images of frontal cortex, ventral midbrain, hippocampus and striatum and did not find a single ORF1p+ cell, which could unambiguously be defined as non-neuronal. This indicated that ORF1p is not or only very rarely expressed in non-neuronal cells. To further confirm the predominant presence of expression of ORF1p in neurons and the absence of ORF1p expression in non-neuronal cells, we used fluorescence-activated cell sorting (FACS) to isolate neurons (using a NeuN antibody) and non-neuronal cells (NeuN-) from the adult mouse brain followed by Western blotting with an antibody against ORF1p (Fig. 2F, 2G). We detected ORF1p exclusively in the neuronal population, confirming the results based on two different imaging approaches. Finally, to assess whether predominant, if not exclusive ORF1p expression in neurons is mouse brain specific or a pattern also applicable to the human brain, we investigated the identity of ORF1p expressing cells in the cingulate gyrus of a healthy human brain. Similar to what we found in the mouse brain, we observed sparse NeuN expression in the white (Suppl Fig. 2E) and extensive NeuN staining in the grey matter corresponding to the cortical layers (right, separated by a dashed white line) with ORF1p+ cells predominantly located in the grey matter (images shown in Figure 2H located in the grey matter). All cells stained by ORF1p were co-stained with NeuN indicating that ORF1p was expressed in neuronal cells in the human brain (Fig. 2H). However, due to the lower signal quality inherent to human post-mortem sections compared to mouse sections, the identity of ORF1p+ cells was estimated to be 80% neuronal by the automated image analysis pipeline, although no ORF1p+ / NeuN-cells could be clearly identified (Fig. 2I). Of all neurons identified, 37.2% were ORF1p+ (Fig. 2J), indicating that, similar to the mouse brain, only a fraction of neurons express ORF1p (Fig. 2H, right).

ORF1p is predominantly expressed in neurons in the mouse brain

(A) ORF1p expression is absent in the white matter (corpus callosum) and predominantly expressed in neurons. Proportion of ORF1p+/NeuN+, ORF1p+/NeuN-, ORF1p-/NeuN+ and ORF1p-/NeuN-cells in the white matter (corpus callosum) and the grey matter (left) and in nine different regions (right) analyzed by the cell detection pipeline on large scale images presented in Figure1A. Exact values can be found in Suppl_Table1.

(B) Representative confocal microscopy images showing ORF1p (red) and NeuN expression (green) in two different regions of the mouse brain. The bottom images show the merge of the two stainings, an overlap of both markers is represented in orange. z-projections; scalebar = 25µm.

(C) Proportion of neurons expressing ORF1p in the frontal cortex and ventral midbrain quantified on confocal images. ns: non-significant; chi-square test on the cell number of the different cell-types analyzed; n=4 mice, data is represented as mean ± SEM.

(D) Proportion of ORF1p+ cells identified as NeuN+ or NeuN- in two different regions, analyzed by confocal microscopy on multiple z-stacks. ns: non-significant; chi-square test, n=4 mice.

(E) ORF1p does not colocalize with glial or microglial cell markers. Representative confocal microscopy images showing ORF1p staining (red) and three different glial cell (GFAP, Sox9, S100β) or microglial (Iba1) markers (white). Note that Iba1 antibody (rabbit) was used with the ORF1p 09 antibody (guinea pig, in house) z-projections, scalebar = 50µm.

(F-G) Separation of neuronal and non–neuronal - cells by FACS confirms predominant neuronal expression of ORF1p. (F) Neuronal (NeuN+) and non-neuronal (NeuN-) cells isolated by fluorescent activated cell sorting (FACS). Dot plots showing autofluorescence versus an appropriate control antibody (IgG rabbit 647; left) and an antibody against NeuN (name of the AB 657, right). The P4 window represents isolated NeuN+ cells (pink) and the P5 fraction NeuN-cells (orange) containing the same number of cells as sorted in P4 for comparison, others NeuN- are represented in blue. (G) Western blot. ORF1p expression (top), in NeuN- and NeuN+ FACS-sorted cells stemming from Figure F.

(H) Representative confocal microscopy images showing ORF1p (red), NeuN (green) and Hoechst (blue) in the cingulate gyrus of the human brain. z-projection; scalebar = 25µm (left). Example of individuals neurons expressing ORF1p or not are shown on the right panel. z-projection; scalebar = 5 µm (right).

(I) Proportion of ORF1p+ cells identified as NeuN+ or NeuN- in the human cingulate gyrus, analyzed by confocal microscopy on multiple z-stacks.

(J) Proportion of neurons expressing ORF1p in the human cingulate gyrus, analyzed by confocal microscopy on multiple z-stacks.

In summary, ORF1p expression in the mouse and human brain is widely restricted to neurons of which a proportion express ORF1p. This raises the question of the function and consequences of ORF1p expression specifically in neurons but also on the dynamic regulation of this expression upon exogenous (exposome) or endogenous (aging) challenges.

ORF1p expression is increased in the aged mouse brain

ORF1p is expressed at steady-state throughout the brain, but if this expression is dynamically regulated is not known. Aging has unequivocally been linked to LINE-1 regulation 16,45 both as a trigger and as a consequence of LINE-1 activation but whether this is true for the brain has not been thoroughly investigated with ORF1p as a read-out. We therefore addressed the question whether advanced age (16-month-old mice) was paralleled by an enhanced presence of ORF1p expression in the brain compared to young, three-month old mice. In the context of aging, we globally observed a reduction in the proportion of ORF1p+/NeuN+ cells using the cell detection workflow applied to large-scale images described in Figure 1A, phenomenon mainly driven by the midbrain motor, the dorsal striatum, the pallidum and the thalamus regions (Fig. 3A, dark red bars, Suppl_Table1). The confocal approach applied to two regions, the frontal cortex and the ventral midbrain (Fig. 3B) confirmed a loss of ORF1p+/NeuN+ cells in the ventral midbrain with no change in cell proportions in the frontal cortex in accordance with the large-scale imaging approach (Fig. 3A). The predominantly neuronal identity of ORF1p+ cells, however, was unchanged (Suppl Fig. 3A) just as the proportion of neurons expressing ORF1p (Suppl Fig. 3B). We observed a significant decrease of NeuN+ cells in the aged ventral midbrain (Fig. 3B, Suppl Fig. 3C). We next analyzed ORF1p expression levels in the context of brain aging. Interestingly, the mean intensity of ORF1p expression increased significantly with age throughout the brain (13% increase brain-wide; Fig. 3C). Frequency distribution analysis unveiled a shift in ORF1p mean expression per cell in aged mice (Fig. 3D). Importantly, the Hoechst mean intensity within nuclei of ORF1p+ cells, serving as an internal control, showed no significant change (Fig. 3E). Among nine analyzed regions, five demonstrated a general increase in ORF1p mean intensity per cell in aged mice (p≤0.05), a change independent from inter-individual variations in both young and aged mice (Fig. 3F). An increase of ORF1p expression was also observed in three others regions albeit not significant. The only exception was the isocortex which remained unchanged with aging. The general increase of ORF1p expression (fold change intensity) in the whole brain, reaching nearly a 30% increase in some regions, is represented on the heatmap in Figure 1G. These results were confirmed by the confocal imaging approach: ORF1p expression in the frontal cortex remained unchanged and the ventral midbrain region increased significantly in aged mice as quantified in Figure 3H and shown with a representative image in Figure 3I. Overall, these results highlight an age-dependent increase in ORF1p expression in neurons throughout the brain with some regions showing an increase of up to 27 % in ORF1p intensities.

ORF1p expression is increased throughout the whole mouse brain in the context of aging

(A) Proportion of ORF1p+/NeuN+, ORF1p+/NeuN+, ORF1p+/NeuN-, ORF1p-/NeuN+ and ORF1p-/NeuN-cell-type in the whole brain (left) and the different analyzed regions (right) of young (three-month aged, n=4) and aged (16-month-old, n=4) mice using the cell detection pipeline on large scale images presented in Figure 1A, data is represented as mean ± SEM. Exact values can be found in Suppl_Table1.

(B) Proportion of ORF1p+/NeuN+, ORF1p+/NeuN+, ORF1p+/NeuN-, ORF1p-/NeuN+ and ORF1p-/NeuN-cell-type in two different regions of young and aged mice, analyzed on multiple z-stack confocal images. ns: non-significant; *p<0.05 calculated using two-way ANOVA with sidak’s multiple comparisons test on the cell number of the different cell-types analyzed; data is represented as mean ± SEM.

(C) ORF1p mean expression per ORF1p+ cell in the brain analyzed on large-scale images. Dot plot showing the ORF1p mean expression per ORF1p+ cell in young (n=4) and aged (n=4) mice in the whole brain (except cerebellum and olfactory bulb). 74985 total cells were analyzed; * p<0.05, two-way ANOVA with sidak’s multiple comparisons; data is represented as mean ± SEM.

(D) Frequency distribution of ORF1p mean intensity in ORF1p+ cells. ***p<0.001, Kolmogorov-Smirnov test.

(E) Frequency distribution of Hoechst mean intensity in the nuclei of OrF1p+ cells. ns: non-significant, Kolmogorov-Smirnov test.

(F) Mean ORF1p expression per ORF1p+ cell in nine different anatomical regions. Dot plot showing the ORF1p mean expression per ORF1p positive cell (n=74985). Adjusted p-value are represented, two-tailed nested t-test followed by a Benjamin, Krieger and Yukutieli test; n=4 young and n=4 aged mice per region, data is represented as mean ± SEM.

(G) Color-coded representation of fold-changes of ORF1p expression with aging. Represented is the fold-change in percent (aged vs young) of the “mean of the mean” ORF1p expression per ORF1p+ cell quantified mapped onto the nine different regions analyzed as shown in (F).

(H) ORF1p expression is increased in the ventral midbrain of aged mice. Dot plot representing ORF1p expression in two different regions of young and aged mice analyzed on confocal images with multiple z-stacks; total cells analyzed = 8381 ns: non-significant *p<0.05, two-tailed one-way ANOVA; dashed lines represent the medians.

(I) Representative confocal microscopy acquisition showing increased ORF1p expression (red) in the ventral midbrain region of aged mice (one z plan is shown). Cell nuclei are shown in blue (Hoechst staining). Scalebar = 50µm.

Coding LINE-1 transcripts are increased in aged human dopaminergic neurons

Following the observation of increased ORF1p expression in the aged mouse brain, among which the ventral midbrain, and given the age-related susceptibility of dopaminergic neurons in the SNpc to cell death and to degeneration in PD 46, we turned to a RNA-seq dataset of laser-captured micro-dissected post-mortem human dopaminergic neurons of brain-healthy individuals 47, in order to interrogate full-length LINE-1 mRNA expression profiles as a function of age. To avoid read-length bias to which TE analysis is particularly sensitive, we analyzed only the data derived from 50bp paired-end reads of linearly amplified total RNA as this dataset represented all age categories (n=41; with ages ranging from 38 to 97; mean age: 79.88 (SD ±12.07); n=6 ≤65y; n=35 >65y; mean PMI: 7.07 (SD ±7.84), mean RIN: 7.09 (±0.94)). As age-related dysregulation of TEs might not be linear, we considered individuals with ages-at-death younger or equal to 65 years as “young” (n=6, 38-65 years, mean age 57.5 years (SD ±9.9)) and individuals older than 65 years as “aged” (n=35, 65-97 years, mean age 83 years (SD±7.8). The expression of the dopaminergic markers tyrosine hydroxylase (TH) and LMX1B were similar in both populations indicating no apparent change of dopaminergic identity of analyzed melanin-positive dopaminergic neurons (Suppl Fig. 4A). Next, we compared the expression of repeat elements at the class, family and name level based on the repeat masker annotation implemented in the UCSC genome browser using a commonly used mapping strategy for repeats 48. No overt dysregulation of repeat elements at either level of repeat element hierarchy was observed (Suppl Fig. 4C-F), however there was a significant increase in several younger LINE-1 elements including L1HS and L1PA2 at the “name” level (Fig. 4A,B). This was not observed for HERVK-int, a human endogenous retrovirus family with some copies having retained coding potential (Fig. 4B) or other potentially active TEs like HERVH-int, HERV-Fc1 or SVA-F with the exception of a trend for an increase in AluYa5 transcripts in the >65y group, a young Alu family mobilized by the LINE-1 retrotransposition machinery (Suppl Fig. 4G). Interestingly, L1HS expression was highly correlated with L1PA2 expression and this correlation extended to almost all younger LINE-1 subfamilies weaning down with evolutionary distance (Fig. 4C). This was not true for other active TEs as L1HS was negatively correlated with HERVK-int expression (Fig. 4C). Several regulators of LINE-1 activity have been identified 17,49 and correlation of their expression with L1HS might allow to infer their relevance of interaction (activation or repression) with L1HS in human dopaminergic neurons. Spearman correlation analysis revealed three known repressors of LINE-1 activity; EN1 (Engrailed 1 17, Suppl Fig. 5A) with important functions for dopaminergic neuron homeostasis 50, CBX5/HP1a, a heterochromatin binding protein binding to the histone mark H3K9me3, thereby mediating epigenetic repression 51 (Suppl Fig. 5B) and XRCC5/6, also known as Ku86/Ku70, which are essential for DNA double-stranded break repair through the nonhomologous end joining (NHEJ) pathway and limit LINE-1 full-length insertions 52 (Suppl Fig. 5C). The transcripts of these genes showed, although not statistically significant, a trend for decreased expression in the elderly (Suppl Fig. 5D-G). Based on the increase of young LINE-1 families L1HS and L1PA2 in aged human dopaminergic neurons and the finding that ORF1p was increased in the aged mouse brain, we focused our attention on LINE-1 elements with coding potential for ORF1 and ORF2 according to the L1Basev2 annotation which are specific elements comprised in the L1HS and L1PA2 annotation at the “name” level. Most of the 146 full-length and coding LINE-1 termed UIDs (= Unique Identifier) in the L1Base are L1HS elements (76.03%), whereas the remaining 35 UIDs belong to the evolutionary older L1PA2 family (Suppl Fig. 6B). The L1Base annotation is based on the human reference genome (GRCh38) and annotates 146 human full-length (>6kB), intact LINE-1 elements (ORF1 and ORF2 intact) with a unique identifier from 1 to 146 4. Attribution of sequencing reads to a specific, individual TE copy is problematic 53 and several approaches have been proposed to circumvent this problem including the mapping of unique reads 48.While several tools using expectation maximization algorithms in assigning multi-mapping reads have been developed and successfully tested in simulations 48,54, we used a different approach in mapping unique reads to the L1Base annotation of full-length LINE-1. Specific “hot” LINE-1 loci in a given cellular context have been identified 3, but usage of the L1Base annotation enabled an unbiased approach albeit ignoring polymorphic LINE-1 sequences. Unique read mapping strategies for repeat elements, especially young LINE-1 elements, will unavoidably underestimate LINE-1 locus-specific expression levels 48, but will be most accurate in assigning reads while allowing the comparison of two different conditions analyzed in parallel. Assuming that expression of UIDs was correlated with mappability, we plotted a mappability count of each UID (see methods) against its mean normalized read count expression of the six individuals ≤65y. Non-parametric Spearman correlation revealed no correlation between UID mappability and expression (Suppl Fig. 6A) indicating no apparent bias between the two parameters. However, individual UID dependency of mappability on expression cannot be excluded, especially for high expressing UIDs like UID-16 for example (Suppl Fig. 6A,E). Expression of LINE-1 at the locus-level has been attributed to artefacts not representing autonomous transcription including differential high intronic read counts 55, pervasive transcription or reads attributable to passive co-transcription with genes when the LINE-1 element is intronic 56. To evaluate the latter, we determined the number of intronic (46.58%) and intergenic UIDs (78/146; Suppl Fig. 6C) and identified the corresponding genes for intronic UIDs (Suppl Fig. 6D). Of the 146 UIDs, 140 passed the threshold of >3 reads in at least 6 individuals. Differential expression of UID between “young” and “aged” dopaminergic neurons revealed several significantly deregulated full-length LINE-1 loci (Fig. 5A). However, while no single locus stood out, paired analysis of the expression of all UIDs indicated a general increase (Fig. 5B, especially of low expressed UIDs. The comparative analysis of the sum expression of UIDs per individual comparing young (≤65y) with elderly human dopaminergic neurons, however, did not reach statistical significance (Fig. 5C). Several specific loci were dysregulated in particular, for instance UID-68 (Fig. 5A), a L1HS element located on chromosome 7 (chr7: 141920659-141926712) in between two genes, CLEC5A (C-type lectin domain containing 5A) and OR9A4 (olfactory receptor family 9 subfamily A member 4). To rule out any influence of “hosting” gene transcription interference on measurable UID-68 expression differences (Fig. 5D, left), we performed Spearman correlation which did not indicate any correlation between CLEC5A or OR9A4 expression with UID-68 (Fig. 5D). Further, UID-68 had a high mappability count of 16 (range of all UIDs: 1-30, mean 9.0 (SD ±6.05), Suppl Fig. 6A) indicating that UID-68 might be a candidate for an age-dependent gain of activity. L1HS UID-129, located on chromosome 15 (chr15: 54926081-54932099) is intergenic, at ≈ 2Mb distance upstream and ≈ 200Mb downstream from the next genes. However, the mappability count of UID-129 was only 1, indicating possible mapping biases inherent to the unique read-based mapping strategy employed (Fig. 5E and Suppl Fig. 6A). Another UID increased in individuals >65y, the L1HS UID-37 (chr10:98,782,942-98,788,971, minus strand), located in intron 3 of the HPSE2 gene (mappability count of 13, minus strand), showed no correlation with its “hosting” gene indicating potential autonomous transcription of this LINE-1 element and suggesting its contribution to the increase of full-length LINE-1 transcripts (Suppl Fig. 7A). We also inspected UID-127 (chr13:40,356,291-40,362,321, mappability count: 14), a L1PA2 element which slightly decreased in elder individuals (Fig. 5A). We found a positive association with LINC00598 (intron 6 of 6; ≤65y: r=0.2, p=0.71; >65y: r=0.37, p=0.03), a non-coding RNA which hosts UID-127 in its 6th intron, indicating transcriptional co-regulation potentially indicative of non-autonomous transcription of UID-127 (Suppl Fig. 7B). The decrease in the expression of UID-137 (Fig. 5A) was mostly driven by one young individual with high expression (Suppl Fig. 7C) and thus not reflecting an overall decrease (Suppl Fig. 7C, p=0.23). In conclusion, TE expression analysis of this human dataset covering an age-span of 59 years indicates an increase in the expression of young LINE-1 elements including those which have coding potential in elderly dopaminergic neurons. A slight net sum increase of UID transcripts/cell might be sufficient for the production of “above steady-state” levels of ORF1p and ORF2p. Other TEs with coding potential, namely members of the HERV family, were not increased. Further, correlation analyses suggest that L1HS expression might possibly be controlled by the homeoprotein EN1, a protein specifically expressed in dopaminergic neurons in the ventral midbrain 50, the heterochromatin binding protein HP1, two known regulators of LINE-1, and the DNA repair proteins XRCC5/6.

Young LINE-1 elements are increased in aged human dopaminergic neurons

TE transcript expression in RNA-seq data of laser-captured micro-dissected post-mortem human dopaminergic neurons of brain-healthy individuals was analyzed using RepeatMasker (multimappers) or the L1Base (unique reads).

(A) Volcano plot of differential analysis of LINE-1 expression using DESeq2 comparing young (≤65y, n=6) or aged (>65y, n=35) human dopaminergic neurons at the “name” level of RepeatMasker. Young LINE-1 elements, including the two families L1HS and L1PA2 that have coding copies, are highlighted in red.

(B) Scatter plots of normalized read counts (“name” level) of the young L1HS and L1PA2 families as well as the human endogenous virus family HERVK-int, another TE family with coding potential comparing young (≤65y, n=6) or aged (>65y, n=35) human dopaminergic neurons. Mann-Whitney test, p<0.05.

(C) Correlation of the expression of LINE-1 elements with known regulators in human dopaminergic neurons. Spearman correlation of evolutionary close (L1HS, L1PA2-17) and distant LINE-1 (L1PB and L1MA5) as well as HERV elements with coding potential (HERV-Kint, HERV-Fc1, HERV-Fc2 and HERV-H-int) with known regulators of their expression. HERV-W and TREX1 did not pass the normalized read count threshold of >3 reads in >6 individuals.

Dysregulation of locus-specific full-length LINE-1 elements in aged human dopaminergic neurons

(A) Volcano plot of differential expression analysis of TE expression using DEseq2 comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons at the locus-level of specific full-length LINE-1 elements (140 of 146 “UID’s” as annotated in L1Base; threshold >3 reads in at least 6 individuals).

(B) Pairwise comparison of the expression of 140 out of 146 full-length LINE-1 elements comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons. Wilcoxon matched signed rank test, p<0.0001left panel).

(C) The sum of read counts of all UIDs per individual were plotted comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons.

(D-E) Dysregulated locus-specific full-length LINE-1 elements (UID-68 and UID-129) are plotted as scatter plots comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons. (D) UID-68 is located adjacent to the genes CLEC5A and OR9A4 (left). Spearman correlation analysis of the expression of UID-68 and CLEC5A (middle) or OR9A4 (right) in young (≤65y, n=6, black dots) or aged (>65y, n=35, red squares) human dopaminergic neurons. (E) UID-129 is intergenic.

Endogenous ORF1p interactors in the mouse brain

To go further in our understanding of steady-state neuronal ORF1p expression across the mouse brain, we immunoprecipitated ORF1p and performed quantitative label-free LC-MS/MS to identify potential protein partners of ORF1p in the mouse brain. We successfully immunoprecipitated endogenous ORF1p from whole brain lysates (Fig. 6A) and identified a total of 424 potential protein interactors associated with ORF1p (Suppl_Table2) in 5 independent experiments (n=5 mice). Using Gene Ontology (GO) analysis, we identified several interacting proteins belonging to GO terms related to known functions of the ORF1p protein in RNA binding, preferentially 57 but not exclusively in cis 58, for instance RNA decapping and mRNA catabolic process, or related to the known presence of ORF1p in ribonucleoprotein particles 59,60 (GO: cytoplasmic ribonucleotide granule) or the presence of ORF1p in p-bodies 58 as shown in Figure 6B and listed in Suppl_Table3. Other GO terms that emerged, to our knowledge not previously associated with ORF1p, were related to cGMP-mediated signaling (GO: cGMP-mediated signaling and 3’-5’phosphodiesterase activity: i.e. PDE4A, PDE4B, PDE4DIP) and the cytoskeleton (GO: microtubule depolymerization, cytoskeleton organization, microtubule and tubulin binding, cytoskeletal motor activity and protein binding). cGMP signaling is regulated by 3’-5’ phosphodiesterases (PDEs) which degrade 3’,5’-cyclic guanosine monophosphate (cGMP) and 3’,5’-cyclic adenosine monophosphate (cAMP), an activity essential for cell physiology for the integration of extra- and intracellular signals including neuronal excitability, synaptic transmission and neuroplasticity 61,62. Further, several ORF1p interacting proteins were constituents of the mating-type switching/sucrose nonfermenting complex (SWI/SNF complex), i.e. ARID1A, ARID1B, SMARCA2, SMARCB1, SMARCC2), an ATP-dependent chromatin remodeler complex disrupting nucleosome/DNA contacts to facilitate DNA/chromatin accessibility by shifting, removing or exchanging nucleosomes along DNA 63,64. Finally, we also observed proteins belonging to the GO term “neuronal cell body”, corroborating with the neuron-specific presence of ORF1p in the brain. A comparative analysis with previous mass spectrometry studies 60,65–70 aimed at identifying ORF1p interacting proteins unveiled significantly more common proteins than randomly expected (overrepresentation test; representation factor 2.6, p< 5.4e-08; Fig. 6C), including LARP1, STAU2, ATXN2, RALY, TARBP2 or DDX21 (for a full list see Suppl_Table4). The presence of a significant number of overlapping ORF1p interactors in different non-neuronal human cells (HEK60,65,66, HeLa67, human breast and ovarian tumors70 and hESCs68) and mouse brain cells (our study), suggest conserved key interactors between both species and between cell types, with a subset of these proteins regulating RNA degradation and translation potentially relevant for the LINE-1 lifecycle itself. ORF1p interactors found in mouse spermatocytes 69 were also present in our analysis including CNOT10, CNOT11, PRKRA and FXR2 among others (Suppl_Table4). To unravel the physical interactions between the identified interactors of endogenous ORF1p within the mouse brain, we used the STRING database (Search Tool for Recurring Instances of Neighboring Genes, https://string-db.org/). This analysis generated a network representation, where physical interactions are represented by edges (Fig. 6D). In analogy with the GO term analysis, ORF1p displayed interactions with various clusters, including well-known RNA decapping complexes directed against LINE-1 RNA, which also encompassed DCP2 and DCP1A which had not previously been identified as interacting with ORF1p 71. Furthermore, ORF1p exhibited interactions with the SWI/SNF complex (highlighted in red) as well as subunits of the RNA polymerase II complex suggesting a direct or indirect association with accessible chromatin, a hitherto unknown interaction of ORF1p with chromatin compartments within the nucleus. Notably, a multitude of novel interactors belonged to the “neuronal cell body” and “neuron projections” clusters, proposing potential neuron-specific partners of ORF1p such as Grm2/5, Bai1, Epha4, Kcnn2, Grik2 and Dmd among others. A last cluster, formed by Ncoa5 (Nuclear Receptor Coactivator 5), Nxf1 (Nuclear RNA Export Factor 1), Ranbp2 and Nup133 (both nucleoporins), might imply a role for these interactions in L1-RNA nuclear export and/or a mechanism for the LINE-1 RNP to gain access to the nucleus in post-mitotic neurons. Altogether, the identification of known and novel interactors of ORF1p in the mouse brain suggests roles of ORF1p in the LINE-1 life cycle (RNA binding and metabolism, RNP formation, nuclear access) but also suggests potential novel physiological roles of ORF1p in the brain related to cytoskeleton organization, cGMP signaling, neuron-specific functions (i.e. synaptic signaling, Suppl_Table3) and chromatin organization and/or transcription regulation.

Endogenous ORF1p interactors in the mouse brain

Immunoprecipitation (IP) of endogenous ORF1p from the mouse brain. WB against ORF1p showing ORF1p enrichment after IP but no signal in the IgG control. Five independent samples were then prepared for proteomic analysis by mass spectrometry (LC-MS/MS).

(B) GO slim enrichment analysis of proteins selected as endogenous ORF1p protein partners in the mouse brain after quantitative LC-MS/MS. ORF1p-immunoprecipitated proteins were categorized into GO slim terms. The nine GO slim term with the highest fold-change are plotted. Fold enrichment is depicted on the upper axis and displayed as bars, the FDR value appears on the lower axis and is represented by the black points. BP: Biological Process, CC: Cellular Component, MF: Molecular Function.

(C) Venn diagram showing common interactors (purple) between interactors of endogenous ORF1p in the mouse brain identified in this study (red) and known (published) interactors of ORF1p (blue). Statistical significance of the overlap between the two groups of proteins was tested by an overrepresentation test (http://nemates.org/MA/progs/overlap_stats.html).

(D) ORF1p associates with the SWI/SNF complex (red), RNA pol II complex (orange) and interactors belonging to GO terms related to neuronal cell body & neuron projection (green). Known interactors previously published 60,65–70 are indicated with a purple ring. STRING network of physical interactions where nodes represent proteins partners identified in (A) and edges thickness represents the strength of shared physical complexes. Only proteins sharing physical interactions were represented.

Discussion

While LINE-1 derepression in aging has been extensively explored in peripheral tissues and various pathologies, including cancer, our understanding of LINE-1, particularly ORF1p expression, in the central nervous system remains limited 20,72,73. A recent search of ORF1p peptides in mass spectrometry data spanning 29 different healthy tissues did not reveal the presence of ORF1p in the brain, suggesting that its presence might lie below detection limits 20. Only a few studies explored ORF1p encoded LINE-1 protein expression in the brain, most focusing on a specific region (in mice 17,33 in rats 74 and in human post-mortem brain 19), but it remained unclear if ORF1p is expressed throughout the entire brain, exhibits cell-type specificity, and most intriguingly, if its expression is influenced by the aging process. Here, we demonstrate that ORF1p is expressed throughout the entire mouse brain and in at least three regions of the human post-mortem brain at steady-state. Leveraging a comprehensive workflow that incorporates brain atlas registration and machine learning algorithms (as described in the methods section), we quantified tens of thousands of brain cells, enabling a profound analysis of cell proportions, cell identities, densities and ORF1p expression levels across the entire brain. Surprisingly, more than one-fifth of detected cells expressed ORF1p. Strikingly, regional variations in ORF1p expression levels were observed, with each region exhibiting distinct proportions, cell density, and signal intensity of ORF1p+ cells. In a non-neurologically diseased human brain, ORF1p is expressed in all three regions examined, that is the cingulate gyrus, the frontal cortex and the cerebellum. This is in accordance to an earlier study using histological staining, which found ORF1p expression in the human frontal cortex, the hippocampus, in basal ganglia, thalamus, midbrain and the spinal cord 19. This suggests, similarly to the mouse brain, a generalized expression across the human brain. On the transcriptomic level using long-read sequencing of GTEx tissues, brain and liver were highlighted as the organs displaying the highest expression of putatively active, full-length LINE-1 elements 75. However, when the authors looked at sub-regions, they found transcript expression in cerebellar hemispheres and the putamen, but not in the caudate and the anterior cingulate gyrus and frontal cortex 75. This is in contrast to our data and the data from Sur et al, where ORF1p was found to be expressed in the latter two regions using two different antibodies. We used the anti-human LINE-1 ORF1p antibody clone 4H1, a well characterized antibody 73,76. While the sample size for the staining of human post-mortem tissues certainly needs to be increased in order to draw quantitative conclusions, the presence of the protein in two independent studies does point to a steady-state expression of ORF1p in the human brain.

In the mouse brain, we find ORF1p to be expressed predominantly if not exclusively in neurons using immunofluorescence and fluorescence-activated cell sorting (FACS). This result is consistent with previous studies, such as the identification of ORF1p in excitatory neurons within the mouse frontal cortex 77, in parvalbumin neurons in the hippocampus 18, its presence in neurons in the ventral midbrain 17 and the recognition of morphological similarities between stained neurons and ORF1p+ cells in a post-mortem hippocampus sample of a healthy individual 19. We also detected ORF1p in Purkinje cells in the mouse and in bulk human cerebellum. Neuronal specificity or preference of LINE-1 expression was also shown on the transcriptomic level in recent studies investigating LINE-1 expression in the mouse hippocampus, where neuronal LINE-1 expression exceeded that of astrocytes and microglia by approximately twofold 20, is abundant in parvalbumin interneurons 18 and single-nuclei RNA-seq data from the mouse hippocampus and frontal cortex which confirmed globally that repetitive elements including LINE-1 are more active in neurons than in glial cells 77. In the human brain, LINE-1 transcripts are found in greater quantities in neurons compared to non-neuronal cells by single-nucleus sequencing 78. Furthermore, retrotransposition-competent LINE-1 elements (similar to UID) are found expressed exclusively in neurons 79. While ORF1p expression is suggested to be expressed in microglia under experimental autoimmune encephalomyelitis conditions in the spinal cord 80, no evidence of such expression was observed in non-neuronal cells under non-pathological condition.

On average, throughout the mouse brain, the majority of neurons was positive for ORF1p and in some regions (i.e. the thalamus) around 80% of neurons expressed ORF1p. Comparing the results of both imaging approaches, the percentages of neurons expressing ORF1p in the ventral midbrain and frontal cortex were roughly similar (around 70% of neurons expressed ORF1p as quantified by confocal imaging and about 60% of neurons were identified as ORF1p+ using the slide scanner approach). In the human cingulate gyrus, we found that 37.2% of neurons express ORF1p and that 80% of cells expressing ORF1p were neurons, which are proportions similar to some regions of the mouse brain. It is however possible that these percentages are underestimated due to technical issues inherent to the machine-learning based algorithm for cell detection as our observations often indicated a positive signal in neurons which were classified as negative due to a particular shape or our stringent intensity threshold. A question which arises based on these findings is whether specific features distinguish ORF1p+ and ORF1p-neurons. One hint comes from a recent study suggesting that in the mouse hippocampus, it is the parvalbumin positive neurons that predominantly express ORF1p 18. We have made a similar observation in the mouse ventral midbrain, where TH-positive dopaminergic neurons express higher levels of ORF1p compared to surrounding, non-dopaminergic neurons 17 (Fig. 1B, panel 8). In the cerebellum, we observed ORF1p staining in Purkinje cells but not in the surrounding granular and molecular layer neurons (Fig. 1B, panel 10). Parvalbumin positive neurons are inhibitory neurons, so are Purkinje cells. However, dopaminergic neurons are modulatory neurons exerting excitatory and inhibitory effects depending on the brain region they act on. Specific neurons in the granular layer (i.e. Golgi and unipolar brush cells) of the cerebellum are inhibitory, but ORF1p negative, indicating that the decisive feature might not be the excitatory or inhibitory nature of a neuron. Another possibility is a cell-type specific chromatin organization permissive for the expression of LINE-1 and future single-cell studies in the mouse and human brain might reveal those differences.

Because transposable elements are known to become active in somatic tissues during aging 15,16,28,81,82, we aimed to investigate whether there was a corresponding increase at the protein level. In aged mice, ORF1p expression significantly increased throughout the mouse brain consistent with a previously documented increase in ORF1p outside the central nervous system in aged rats 74,83 and aged mice 81. By quantifying the mean intensity of ORF1p in over 70 000 cells identified as ORF1p+, we were able to characterize the extent of this increase in each anatomical sub-region. Remarkably, apart from the isocortex which did not show any change, ORF1p expression increased in all other brain regions by 7% to 27%, indicating a generalized increase of ORF1p expression in neurons throughout the brain (13%). We did not detect any change in cell identity of ORF1p expressing cells, that is, ORF1p expression remained predominantly if not exclusively neuronal. Intriguingly, we observed a loss of NeuN+ cells, particularly those expressing ORF1p, throughout the brain which was more pronounced in specific regions (ventral midbrain) than in others (isocortex). The loss of NeuN+ cells could either be due to a loss of neuronal identity, as described recently in the context of neuronal aging 84 and in the context of Alzheimer disease and related tauopathies 85, to a neurodegenerative process per se or to gliosis related to the aging process 86 (as we observe a slight increase of ORF1p-/NeuN-cells per mm2).

Interestingly, a region with a strong increase in ORF1p expression with aging (ventral midbrain) also had a significant loss of NeuN+ cells while a region with no change in ORF1p expression with aging (isocortex, frontal cortex) did not lose NeuN+ cells. However, further investigations are necessary to validate a correlation and to investigate an underlying mechanism. An increase of ORF1p might have several direct or indirect consequences on a cell or here, on a neuron. As ORF1p is translated from a polycistronic LINE-1 RNA together with ORF2p, albeit in much higher amounts (the estimated ratio ORF1p to ORF2p is 240:1) 87, it can be expected that a LINE-1 ribonucleotide particles are formed and ORF2-dependent cell toxicity in form of genomic instability 17,21 and single-stranded cytoplasmic DNA triggered inflammation 27,28,81 might result. This has been shown in mouse dopaminergic neurons where oxidative stress induced LINE-1 causally contributed to neurodegeneration 17.

Neurodegeneration was partially prevented by anti-LINE-1 strategies among which NRTIs 17 and similar LINE-1 protein-dependent neuronal toxicity has been shown in drosophila 31,32 and the mouse cerebellum 33.

In order to test whether an increase in LINE-1 is a feature of human brain aging, we turned to a unique RNA-seq dataset of human laser-captured dopaminergic neurons of 41 individuals ranging from 38 to 99 years 47. In accordance with our focus on LINE-1 sequences which are full-length and coding, we developed a rationale to interrogate LINE-1 families with representatives that are coding (L1HS, L1PA2, multimappers; RepeatMasker) and to specifically investigate full-length LINE-1 elements that have intact open reading frames for ORF1p and ORF2p (unique reads; L1Basev2 4). Indeed, we find an increase in L1HS and L1PA2 elements in individuals ≥65y as well as an increase in specific full-length LINE-1 elements but only a trend for increase of all full-length LINE-1 in sum in the elderly. This analysis has technical limitations inherent to transcriptomic analysis of repeat elements especially as it is based on short-read sequences. Nevertheless, we tried to rule out several biases by demonstrating that mappability did not correlate with expression and that the expression of intronic full-length LINE-1 elements is not correlated with the expression of their “hosting” gene. Interestingly, dysregulated full-length LINE-1 elements in aged dopaminergic neurons did not correspond to those identified in bladder cancer 88 indicating the intricate nature of this expression across tissues and pathological conditions. Overall, a slight net sum increase of UID transcripts/cell might be sufficient for the production of “above steady-state” levels ORF1p and ORF2p. Further, a dissociation of LINE-1 transcript and protein levels in aging has been observed recently in excitatory neurons of the mouse cortex. In the absence of transcriptional changes of LINE-1, protein levels of ORF1p were increased 77.

We can only speculate about the reason for an increase in ORF1p in the aged brain. A recent single-cell epigenome analysis of the mouse brain suggested a specific decay of heterochromatin in excitatory neurons of the mouse brain with age which was paralleled by an increase in ORF1p, albeit equally in excitatory and inhibitory neurons, again not indicating any dependency of ORF1p regulation on the excitatory or inhibitory nature of neurons 77. Chromatin and particularly heterochromatin disorganization are a primary hallmark of aging 82 but other repressive cellular pathways which control the LINE-1 life cycle might also fail with aging 13. Another possibility is a loss of accessibility of repressive factors to the LINE-1 promoter or an age-dependent decrease in their expression. Matrix correlation analysis of several known LINE-1 regulators, both positive and negative, revealed possible regulators of young LINE-1 sequences in human dopaminergic neurons. Despite known and most probable cell-type unspecific repressive factors like the heterochromatin binding protein CBX5/HP1 51 or the DNA repair proteins XRCC5 and XRCC6 49, we identified the homeoprotein EN1 as negatively correlated with young LINE-1 elements including L1HS and L1PA2. EN1 is an essential protein for mouse dopaminergic neuronal survival 50 and binds, in its properties as a transcription factor, to the promoter of LINE-1 17. As EN1 is specifically expressed in dopaminergic neurons in the ventral midbrain, our findings suggests that EN1 controls LINE-1 expression in human dopaminergic neurons as well and serves as an example for a neuronal sub-type specific regulation of LINE-1.

The heterogenous, brain-wide presence of ORF1p expression at steady-state is intriguing. In cancer cell lines or mouse spermatocytes, ORF1p interacts with several “host” proteins, some if not most of which are related to the LINE-1 life cycle. However, a profile of endogenous ORF1p interactors in the mouse brain might inform on possible other and organ-specific functions besides its binding to the LINE-1 RNA in “cis” 69. Among the total 424 potential interactors of endogenous ORF1p in the mouse brain, 38 partners had been previously identified by mass spectrometry in human cancers, cancerous cell lines and mouse spermatocytes 60,65–70 (Suppl_Table4). Further, GO term analysis contained expected categories like “P-body”, mRNA metabolism related categories and “ribonucleoprotein granule”. We also identified NXF1 as a protein partner of ORF1p, a protein found to interact with LINE-1 RNA related to its nuclear export 89. This suggests the conservation of key interactors probably essential for completing or repressing the LINE-1 life cycle in both species, despite the divergence of mouse and human ORF1p protein sequences 90. Along these lines, several ORF1p protein partners we identified might complete the list of post-transcriptional regulators implicated in LINE-1 silencing. Recent work conducted on human cancerous cell lines has demonstrated that MOV10 orchestrates the recruitment of DCP2 for LINE-1 RNA decapping 71. In our analysis, we identified DCP2 along with DCP1A, known to enhance the decapping activity of DCP2 91, and DCP1b, a pivotal component of the mRNA decapping complex 92. Intriguingly, MOV10 was not detected in our mass spectrometry analysis, despite its established role in recruiting DCP2 and forming a complex with L1-RNP to mediate LINE-1 RNA decapping, as reported by Liu et al 71. However, we found two enhancers of mRNA decapping, EDC3 and EDC4, both core components of P-bodies, a membrane-less organelle known to contain L1-RNP 58. Multiple ubiquitin-ligase proteins were found although not appearing as a significantly enriched GO term. These results complete the picture of the post-transcriptional and translational control of ORF1p and suggest that these mechanisms, despite a steady-state expression, are operational in neurons. Further, several neuron-specific interactors were identified belonging to GO term categories “neuron projection” (75 proteins) and “neuronal cell body” (5 proteins), again pointing to the neuron-predominant expression of ORF1p in the mouse brain. Other interesting aspects were raised from this analysis. Among significantly enriched GO terms, several were related to the cytoskeleton, the functional consequences of which need to be determined in future studies. Our screen also identified PDE10A as an interactor of ORF1p in the mouse brain, a PDE almost exclusively expressed in medium spiny neurons of the striatum and a target for treatment of neurological diseases related to basal ganglia function like Huntington’s disease, schizophrenia and Tourette syndrome 93. Interestingly, PDE10A inhibition is related to beta-catenin signaling, another GO term which emerged from our screen 94. Finally, we found components of RNA polymerase II and the SWI/SNF complex as partners of ORF1p. This further indicates that ORF1p has access to the nucleus in mouse brain neurons as described for other cells 95,96, implying that ORF1p potentially has access to chromatin. These findings give rise to intriguing questions regarding the potential function of ORF1p in neuron in health and disease as (i) ORF1p is widely distributed throughout the brain under normal physiological conditions, (ii) ORF1p shows a wide range expression levels within and inbetween regions, (iii) ORF1p is expressed predominantly if not exclusively in neurons, (iv) but not in all neurons and (v) interacts with proteins that might not directly relate to the LINE-1 life cycle, some of which are neuron-specific. In addition, physicochemical properties of ORF1p to form compacted nucleic acid - bound complexes with sequestration potential were shown 90,97. Future loss-of-function studies should help to shed light on the necessity of ORF1p for neuronal functions if they exist. Our data spurs the idea of a possible “physiological” function of ORF1p as an integrative protein with exapted function in neuronal homeostasis and a loss of restriction in the aged brain limiting LINE-1 expression to steady-state levels.

Materials & Methods

Animals

Swiss OF1 wild-type mice (Janvier) were housed on a 12h light/dark cycle with free access to water and food. Mice were sacrificed at 3-month or 16-month. Animal experiments were performed according to the EU directive 2010/63/EU.

Mouse tissue dissection and protein extraction

Tissues were extracted from 3-month-old and 16-month-old Swiss/OF1 mice. Briefly, the two hemispheres were separated in ice cold PBS -/-. For each mouse, one hemisphere was rinsed and fixed in 4% PFA for 1h followed by 24h of incubation in 30% sucrose. Hemispheres were kept at −20°C until being sliced on a freezing microtome (Epredia, HM 450) with a 20µm thickness. The other hemisphere was dissected in ice cold PBS -/- 1X and 6 brain regions were rinsed, cut in small pieces and dissociated separately using a large (21G) to small gauge (27G) needle in RIPA lysis buffer for 5 min. Lysates were kept on ice for 25 min, were sonicated for 15 min and supernatants were collected after a 30 min centrifugation at 4°C at 14 000 rpm.

Proteins were quantified and Laemmli buffer was added before boiling for 10min at 95°C to be used for Western Blot.

Human Samples

Cerebellum, frontal cortex and cingulate gyrus human samples were provided by Neuro-CEB biobank from a 78-year-old healthy male individual and conserved at −80°C.

Human Samples pulverization and protein extraction

We used the dry pulverizer Cryoprep (Covaris) for pulverization of tissue blocs. Each sample was disposed in a liquid-nitrogen precooled Tissue-tube bag and dry cryo-pulverized with one impact at the maximum level. The pulverized brain sample was then weighed and resuspended in lysis buffer (mg/v) (0.32M sucrose, 5mM CaCl2, 3mM Mg(CH3COOH)2, 0.1mM EDTA, 10mM Tris-HCL pH8, 1mM DTT, 0.1% TritonX-100 and Protease Inhibitors), kept on ice for 30 min with gentle up-and-down pipetting until homogenization. We added 2X RIPA buffer (v/v) to totals fractions for 30 min on ice. We then sonicated samples 2 times for 15 min. AtlasSupernatants were collected after a 30 min centrifugation at 14 000 rpm at 4°C, proteins were quantified and Laemmli buffer was added to be used for Western Blot. All samples were boiled 10 min at 95°C to be used for Western Blot.

Western-Blot

We used 1.5mm NuPAGE 4-12% Bis-Tris-Gel (Invitrogen™). Proteins samples were loaded and gel migration was performed with NuPAGE™ MES SDS Running Buffer (Invitrogen™) for 45 minutes at 200mV. Gels were transferred onto a methanol activated PVDF membrane (Immobilon) in a buffer containing: Tris 25 mM, pH=8,3 and Glycine 192 mM, during 1h30 at 400 mA. Membranes were blocked 30 min with 5% milk in TBST (0,2% Tween 20, 150 mM NaCl, 10 mM Tris pH:8). Primary antibodies were diluted in 5% milk in TBST, and membranes were incubated o/n at 4 C°. After 3 x 10 min washing in TBST, membranes were incubated for 1h30 with the respective secondary antibodies diluted at a concentration of 1/2000 in 5% milk TBST. Membranes were washed 3 x 10 min in TBST and were revealed by the LAS-4000 Fujifilm system using Clarity Western ECL Substrate (Bio Rad) or Maxi Clarity Western ECL Substrate (Bio Rad).

Immunostaining

Sagittal mouse brains slices were fixed for 10min in PFA 4% and rinsed 3 times for 10min in PBS -/-. Slices were then incubated 20 min in glycine 100mM, washed 3 times for 5min in PBS and immersed in 10mM citrate pH 6 at 62°C during 45min for antigen retrieval. Slices were then immersed 3 times in PBS with Triton X-100 0.2% and incubated in blocking buffer for 1,5 h (PBS with Triton X-100 0.2% and FBS (10%) previously inactivated 20min at 56°C (Gibeco, 16141061). Primary antibodies (ORF1p antibody: abcam ab216324; NeuN antibody: GeneTex GTX00837) were diluted (1/200 and 1/500 respectively) in blocking buffer and incubated with slices overnight at 4°C and then washed 3 times for 10 min with PBS. For validation, an in-house ORF1p antibody was used (09) (guinea pig, 1/200). For non-neuronal marker (GFAP antibody: Millipore AB5541; Iba1 antibody: GeneTex GTX101495; Sox9 antibody: RnDsystems AF3075; S100β antibody: Sigma S2532), antibodies were diluted at 1/500.

Suitable secondary antibodies (Invitrogen) and Hoechst (Invitrogen, 15586276) were incubated for 1,5h at 1/2000 in PBS with inactivated FBS (10%) and washed 3 times 10 min in PBS. To quench tissue autofluorescence, especially lipofuscin, TrueBlack Plus (Biotium) in PBS was used during 10min. Slices were rinsed 3 times in PBS and mounted with Fluoromount (Invitrogen™). For human cingulate gyrus stainings, the same protocol was performed, with the difference that a human ORF1p antibody (Abcam 245249) was used. Mouse and human brain slices were imaged by the Axioscan 7 Digital Slide Scanner (Zeiss) or a Spinning Disk W1 confocal microscope (Yogogawa).

Blocked peptide

The ORF1p antibody (abcam ab216324) was incubated 2h on a turning wheel with excess (4:1) of mouse ORF1p recombinant protein as in 17 before the blocked antibody was used in the above-described immunofluorescence protocol.

Quantification of confocal acquisitions

Analysis was performed with a custom-written plugin developed for the Fiji software, using Bio-Formats 98 and 3D Image Suite 99 libraries. Code is freely available online at https://github.com/orion-cirb/DAPI_NEUN_ORF1P. Nuclei were detected using Hoechst channel downscaled by a factor of 2 with the 2D-stitched version of Cellpose 100 (percentile normalization = [1-99], model = ‘cyto’, diameter = 30, flow threshold = 0.4, cell probability threshold = 0.0, stitching threshold = 0.75). The segmented image is rescaled to its original size, and the obtained 3D nuclei are filtered by volume to avoid false positive detections. NeuN and ORF1p+ cells were detected in their respective channel in the same manner as nuclei, but with different Cellpose settings (model = ‘cyto2’, diameter = 40, flow threshold = 0.4, cell probability threshold = 0.0, stitching threshold = 0.75). Then, each cell was associated with a nucleus having at least half of its volume in contact with it. Cells without any associated nucleus were filtered out. Each nucleus was thus labeled according to NeuN and/or ORF1p positivity.

ABBA Registration and Qupath analysis

We registered each sagittal brain section with the Allen Mouse Brain Atlas (CCFv3 101), using the Aligning Big Brains & Atlases plugin 102 in Fiji. Registration results were imported into QuPath software 103 for downstream processing.

In each brain subregion, cell analysis was done with custom Groovy scripts developed for QuPath. Code is freely available online at https://github.com/orion-cirb/QuPath_ORF1P. In brief, Hoechst nuclei were detected with StarDist 2D 104, applying the DSB 2018 pretrained model with the following parameters: percentile normalization = [1-99], probability threshold = 0.75, overlap threshold = 0.25. Cells in NeuN and ORF1p channels were detected with Cellpose 2D (percentile normalization = [1-99], model = ‘cyto2’, diameter = 30, flow threshold = 0.4, cell probability threshold = 0.0). Nuclei and cells were then filtered by area and intensity in their respective channel, once again to avoid false positive detections. Minimum intensity threshold was based on the channel background noise. This noise was estimated for each subregion as the mean intensity of pixels not belonging to any nucleus or cell detected in the corresponding channel. Finally, each cell was associated with a nucleus having its centroid located inside the cell mask. Cells without any associated nucleus were filtered out. Each cell containing a nucleus was thus identified as NeuN+ or NeuN- and ORF1p+ or ORF1p-. Intensity values were normalized by subtracting the background noise computed in the corresponding channel and subregion. As a last step, subregional results were merged into regional ones and data were analyzed using the Pandas Python library 105.

FACS

Mouse brains were dissociated with Adult Brain Dissociation kit (Miltenyi Biotec, 130-107-677) and incubated with the coupled antibody NeuN Alexa 647 (Abcam, ab190565) or the control isotype IgG Alexa 647 (Abcam, ab199093). Stained cells were filtered a last time with a 40µm filter before FACS sorting (FACS ARIA II). Neuronal and non-neuronal cells were separately collected in PBS -/- 2m EDTA and then centrifugate (5min at 700rpm). Pellets were resuspended in RIPA for protein extraction in an appropriate volume in order to achieve equal cell concentrations (10 000 cells/µl).

RNA-seq analysis

The RNA-seq dataset from Dong et al 47 was downloaded from dbGAP (phs001556.v1.p1) and contains unstranded paired-end 50bp and 75bp reads from pooled laser-capture micro-dissected dopaminergic neurons from human post-mortem brain (107 samples) from 93 individuals w/o brain disease. RNA-seq had been done on total and linearly amplified RNA. We focused our analysis on data obtained with 50bp reads, in order to avoid mappability bias, while still regrouping all age categories (n=41; with ages ranging from 38 to 97 (mean age: 79.88 (SD ±12.07); n=6 ≤65y; n=35 >65y; mean PMI: 7.07 (SD ±7.84), mean RIN: 7.09 (±0.94)). Sequencing reads were aligned on the Human reference genome (hg38) using the STAR mapper (v2.7.0a) 3 and two different sets of parameters. Genome-wide individual repeat quantification was performed using uniquely mapped reads and the following STAR parameters: -- outFilterMultimapNmax 1–-alignEndsType EndToEnd–-outFilterMismatchNmax 999–- outFilterMismatchNoverLmax 0.06. Repeats class, family and name quantification was performed using a random mapping procedure and the following parameters : -- outFilterMultimapNmax 5000–-outSAMmultNmax 1–-alignEndsType EndToEnd–- outFilterMismatchNmax 999–-outFilterMismatchNoverLmax 0.06–-outSAMprimaryFlag OneBestScore–-outMultimapperOrder Random. Repeats annotations were downloaded from the UCSC Table Browser (repeatMasker database: https://genome.ucsc.edu/cgi-bin/hgTables) and coordinates of LINE-1 full length and coding elements were downloaded from the L1base database 2 (http://l1base.charite.de/l1base.php; 4) selecting LINE-1 full length elements containing two predicted complete open reading frames for ORF1 and ORF2 (UID= Unique IDentifier) from the LINE-1 database (http://l1base.charite.de/l1base.php) and corrected genomic intervals with the repeat masker annotation of the corresponding genomic locus. Repeat quantification from the aligned data was done using a gtf file composed of all genes (Gencode v29) and all individual repeat elements. This strategy was used to avoid overestimation of repeat elements due to overlaps with expressed genes. For individual repeat quantification of the full length L1 elements (L1base), we therefore used a gtf of all genes and all L1base entries, and ran the FeatureCounts tool 106 with the following parameters: -g gene_id -s 0 -p. In the context of the family-based analysis, we used a gtf with all genes and all annotated repeats elements and ran FeatureCounts with -g gene_family -s 0 -p -M. Before DeSeq2 analysis, we remove all genes and repeat elements with less than 10 reads in a minimum of n individuals, n being the number of individuals in the condition containing the fewest individuals (“young” condition: n=6, 38-65y, mean 57.5 years). We use the same conditions with genes and UIDs with less than 3 reads in a minimum of n individuals. Finally, we calculated the scaling factors using DeSeq2 on all genes + all repeat elements or all gene + UID according to the quantification method and then applied these scaling factors to the corresponding counts tables.

In order to test for the mappability of each UID (= full-length and coding LINE-1), we extracted the bed track « main on human:umap50 (genome hg38) from the UCSC genome browser (≈ 7Mio regions) directly into Galaxy (usegalaxy.org) and joined genomic intervals with a minimum overlap of 45bp of this dataset with a dataset containing the annotation of UIDs extracted from L1Basev2 4 corrected in length with repeat masker and completed with information on whether the UID is intra- or intergenic and, if intragenic, in which gene (NM_ID, chr, strand, start, end, gene length, number of exons, gene symbol) the fl-LINE-1 is located, which resulted in 1266 regions. We then used the “group on data and group by” function in Galaxy and counted the number of overlapping 50kmers with all 146 UIDs (=mappability score). Correlation analysis (non-parametric Spearman) was then done between the mappability score and the normalized read counts.

Immpunoprecipitation of ORF1p from the mouse brain

For immunoprecipitation, we used ORF1p (abcam, ab245122) and IgG rabbit (abcam, ab172730) antibodies. The antibodies were coupled to magnetic beads using the Dynabeads® Antibody Coupling Kit (Invitogen, 14311D) according to the manufacture’’s recommendations. We used 5µg of antibody for 1 mg of beads and used 1.5mg of beads for IP. Individual mouse brain lysates (n=5), homogenized using dounce and sonicated, were incubated with ORF1p or IgG-control coupled beads and a small fraction was kept as input. Each of these two tubes containing coupled beads and brain lysates were diluted in 5 ml buffer (10 mM Tris HCl, 150 mM NaCl, protease inhibitor). The samples were then incubated overnight on a wheel at 4°C. Samples were then washed 3 times with 1 ml buffer (10 mM Tris HCl pH 8, 200 mM NaCl) using a magnet and then resuspended in the same buffer. The samples were boiled in Laemmli buffer (95°C, 10 min) and 20 µl of each sample were loaded on a 4-12% Nupage gel (Invitrogen, NP0336) to be revealed by WB. For samples used in Mass Spectrometry study, beads were washed with buffer (10 mM Tris HCl pH 8, 200 mM NaCl) using a magnet. After 3 washes with 1ml buffer the beads were washed twice with 100 µL of 25 mM NH4HCO3 (ABC buffer). Finally, beads were resuspended in 100 μl of 25mM ABC buffer and digested by adding 0.20 μg of trypsine/LysC (Promega) for 1 hour at 37 °C. A second round of digestion was applied simultaneously on the beads by adding 100 µL of 25 mM ABC buffer and to the previous digest by adding 0.20 µg of trypsin/LysC for 1 hour at 37 °C. Samples were then loaded into homemade C18 StageTips packed by stacking three AttractSPE® disk (#SPE-Disks-Bio-C18-100.47.20 Affinisep) into a 200 µL micropipette tip for desalting. Peptides were eluted using a ratio of 40:60 CH3CN:H2O + 0.1% formic acid and vacuum concentrated to dryness with a SpeedVac device. Peptides were reconstituted in 10 µL of injection buffer in 0.3% trifluoroacetic acid (TFA) before liquid chromatography-tandem mass spectrometry (LC-MS/MS) analysis.

Mass Spectrometry

Online chromatography was performed with an RSLCnano system (Ultimate 3000, Thermo Scientific) coupled to a Q Exactive HF-X with a Nanospay Flex ion source (Thermo Scientific). Peptides were first trapped on a C18 column (75 μm inner diameter × 2 cm; nanoViper Acclaim PepMapTM 100, Thermo Scientific) with buffer A (2/98 MeCN/H2O in 0.1% formic acid) at a flow rate of 2.5 µL/min over 4 min. Separation was then performed on a 50 cm x 75 μm C18 column (nanoViper Acclaim PepMapTM RSLC, 2 μm, 100Å, Thermo Scientific) regulated to a temperature of 50°C with a linear gradient of 2% to 30% buffer B (100% MeCN in 0.1% formic acid) at a flow rate of 300 nL/min over 91 min. MS full scans were performed in the ultrahigh-field Orbitrap mass analyzer in ranges m/z 375–1500 with a resolution of 120 000 at m/z 200. The top 20 intense ions were subjected to Orbitrap for further fragmentation via high energy collision dissociation (HCD) activation and a resolution of 15 000 with the intensity threshold kept at 1.3 × 105. We selected ions with charge state from 2+ to 6+ for screening. Normalized collision energy (NCE) was set at 27 and the dynamic exclusion of 40s. For identification, the data were searched against the Mus Musculus (UP000000589_10090 012019) Uniprot database using Sequest HT through proteome discoverer (version 2.4). Enzyme specificity was set to trypsin and a maximum of two-missed cleavage sites were allowed. Oxidized methionine, Met-loss, Met-loss-Acetyl and N-terminal acetylation were set as variable modifications. Maximum allowed mass deviation was set to 10 ppm for monoisotopic precursor ions and 0.02 Da for MS/MS peaks. The resulting files were further processed using myProMS97 v3.10.0. FDR calculation used Percolator and was set to 1% at the peptide level for the whole study. The label free quantification was performed by peptide Extracted Ion Chromatograms (XICs), reextracted by conditions and computed with MassChroQ version 2.2.21 107. For protein quantification, XICs from proteotypic peptides shared between compared conditions (TopN matching) with missed cleavages were used. Median and scale normalization at peptide level was applied on the total signal to correct the XICs for each biological replicate (n=5). To estimate the significance of the change in protein abundance, a linear model (adjusted on peptides and biological replicates) was performed, and p-values were adjusted using the Benjamini–Hochberg FDR procedure. Proteins with at least three peptides, identified in each biological replicates of ORF1p condition, a 10-fold enrichment and an adjusted p-value ≤ 0.05 were considered significantly enriched in sample comparisons. Unique proteins were considered with at least three peptides in all replicates. Protein selected with these criteria were used for Gene Ontology enrichment analysis and string network analysis.

The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consortium (http://proteomecentral.proteomexchange.org) via the PRIDE partner repository 108 with the dataset identifier PXD047160.

GO term and STRING network analysis

Gene Ontology analysis was performed using GO PANTHER (PANTHER: Making genome-scale phylogenetics accessible to allPaul D. Thomas, Dustin Albou and Huaiyu MiProtein Society. 2022;31(1):8-22. doi:10.1002/pro.4218) and String network physical interactions were retrieved using the STRING database v11.5 (https://string-db.org/) and then implemented in Cytoscape software 109.

Statistical analyses

In column comparisons, data in each column were tested for normality using two normality and lognormality tests (Shapiro-Wilk test and Kolmogorov-Smirnov test). Data which passed the normality tests were analyzed subsequently by a parametric test, data which did not pass the normality tests were analyzed by a non-parametric statistical test as indicated in the figure legends. The significance threshold was defined as p<0.05. Statistical analyses were done with PRISM software (v10).

Supporting information

eLife Assessment

Bonnifet et al. present data on the expression and interacting partners of the transposable element L1 in the mammalian brain. The work includes important findings addressing the potential role of L1 in aging and neurodegenerative disease. However, several aspects of experimental evidence presented are preliminary and the study remains incomplete in its current form.

Reviewer #1 (Public Review):

Summary:

In this study, Bonnifet et al. profile the presence of L1 ORF1p in the mouse and human brain. They claim that ORF1p is expressed in the human and mouse brain at a steady state and that there is an age-dependent increase in expression. This is a timely report as two recent papers have extensively documented the presence of full-length L1 transcripts in the mouse and human brain (PMID: 38773348 & PMID: 37910626). Thus, the finding that L1 ORF1p is consistently expressed in the brain is not surprising, but important to document.

Strengths:

Several parts of this manuscript appear to be well done and include the necessary controls. In particular, the evidence for steady-state expression of ORF1p in the mouse brain appears robust.

Weaknesses:

Several parts of the manuscript appear to be more preliminary and need further experiments to validate their claims. In particular, the data suggesting expression of L1 ORF1p in the human brain and the data suggesting increased expression in the aged brain need further validation. Detailed comments:

(1) The expression of ORF1p in the human brain shown in Figure 1j is not convincing. Why are there two strong bands in the WB? How can the authors be sure that this signal represents ORF1p expression and not non-specific labelling? Additional validations and controls are needed to verify the specificity of this signal.

(2) The data shown in Figure 2g are not convincing. How can the authors be sure that this signal represents ORF1p expression and not non-specific labelling? Extensive additional validations and controls are needed to verify the specificity of this signal.

(3) The data showing a reduction in ORF1p expression in the aged mouse brain is confusing and maybe even misleading. Although there is an increase in the intensity of the ORF1p signal in ORF1p+ cells, the data clearly shows that fewer cells express ORF1p in the aged brain. If these changes indicate an overall loss or gain of ORF1p, expression in the aged brain is not resolved. Thus, conclusions should be more carefully phrased in this section. It is important to show the quantification of NeuN+ and NeuN- cells in young vs aged (not only the proportions as shown in Figure 3b) to determine if the difference in the number of ORF1p+ cells is due to loss of neurons or perhaps a sampling issue. More so, it would be essential to perform WB and/or proteomics experiments to complement the IHC data for the aged mouse samples.

(4) The transcriptomic data presented in Figure 4 and Figure 5 are not convincing. Quantification of transposon expression on short read sequencing has important limitations. Longer reads and complementary approaches are needed to study the expression of evolutionarily young L1s (see PMID: 38773348 & PMID: 37910626 for examples of the current state of the art). Given the read length and the unstranded sequencing approach, I would at least ask the authors to add genome browser tracks of the upregulated loci so that we can properly assess the clarity of the results. I would also suggest adding the mappability profile of the elements in question. In addition, since this manuscript focuses on ORF1p, it would be essential to document changes in protein levels (and not just transcripts) in the ageing human brain.

(5) More information is needed on RNAseq of microdissections of dopaminergic neurons from 'healthy' post-mortem samples of different ages. No further information on these samples is provided. I would suggest adding a table with the clinical information of these samples (especially age, sex, and cause of death). The authors should also discuss whether this experiment has sufficient power. The human ageing cohort seems very small to me.

(6) The findings in this manuscript apply to both human and mouse brains. However, the landscape of the evolutionarily young L1 subfamilies between these two species is very different and should be part of the discussion. For example, the regulatory sequences that drive L1 expression are quite different in human and mouse L1s. This should be discussed.

(7) On page 3 the authors write: "generally accepted that TE activation can be both, a cause and consequence of aging". This statement does not reflect the current state of the field. On the contrary, this is still an area of extensive investigation and many of the findings supporting this hypothesis need to be confirmed in independent studies. This statement should be revised to reflect this reality.

Reviewer #2 (Public Review):

Summary:

Bonnifet et al. sought to characterize the expression pattern of L1 ORF1p expression across the entire mouse brain, in young and aged animals, and to corroborate their characterization with Western blotting for L1 ORF1p and L1 RNA expression data from human samples. They also queried L1 ORF1p interacting partners in the mouse brain by IP-MS.

Strengths:

A major strength of the study is the use of two approaches: a deep-learning detection method to distinguish neuronal vs. non-neuronal cells and ORF1p+ cells vs. ORF1p- cells across large-scale images encompassing multiple brain regions mapped by comparison to the Allen Brain Atlas, and confocal imaging to give higher resolution on specific brain regions. These results are also corroborated by Western blotting on six mouse brain regions. Extension of their analysis to post-mortem human samples, to the extent possible, is another strength of the paper. The identification of novel ORF1p interactors in the brain is also a strength in that it provides a novel dataset for future studies.

Weaknesses:

The main weakness of the study is that cell type specificity of ORF1p expression was not examined beyond neuron (NeuN+) vs non-neuron (NeuN-). Indeed, a recent study (Bodea et al. 2024, Nature Neuroscience) found that ORF1p expression is characteristic of parvalbumin-positive interneurons, and it would be very interesting to query whether other neuronal subtypes in different brain regions are distinguished by ORF1p expression. The data suggesting that ORF1p expression is increased in aged mouse brains is intriguing, although it seems to be based upon modestly (up to 27%, dependent on brain region) higher intensity of ORF1p staining rather than a higher proportion of ORF1+ neurons. Indeed, the proportion of NeuN+/Orf1p+ cells actually decreased in aged animals. It is difficult to interpret the significance and validity of the increase in intensity, as Hoechst staining of DNA, rather than immunostaining for a protein known to be stably expressed in young and aged neurons, was used as a control for staining intensity. The main weakness of the IP-MS portion of the study is that none of the interactors were individually validated or subjected to follow-up analyses. The list of interactors was compared to previously published datasets, but not to ORF1p interactors in any other mouse tissue.

The authors achieved the goals of broadly characterizing ORF1p expression across different regions of the mouse brain, and identifying putative ORF1p interactors in the mouse brain. However, findings from both parts of the study are somewhat superficial in depth.

This provides a useful dataset to the field, which likely will be used to justify and support numerous future studies into L1 activity in the aging mammalian brain and in neurodegenerative disease. Similarly, the list of ORF1p interacting proteins in the brain will likely be taken up and studied in greater depth.

Reviewer #3 (Public Review):

The question about whether L1 exhibits normal/homeostatic expression in the brain (and in general) is interesting and important. L1 is thought to be repressed in most somatic cells (with the exception of some stem/progenitor compartments). However, to our knowledge, this has not been authoritatively / systematically examined and the literature is still developing with respect to this topic. The full gamut of biological and pathobiological roles of L1 remains to be shown and elucidated and this area has garnered rapidly increasing interest, year-by-year. With respect to the brain, L1 (and repeat sequences in general) have been linked with neurodegeneration, and this is thought to be an aging-related consequence or contributor (or both) of inflammation. This study provides an impressive and apparently comprehensive imaging analysis of differential L1 ORF1p expression in mouse brain (with some supporting analysis of the human brain), compatible with a narrative of non-pathological expression of retrotransposition-competent L1 sequences. We believe this will encourage and support further research into the functional roles of L1 in normal brain function and how this may give way to pathological consequences in concert with aging. However, we have concerns with conclusions drawn, in some cases regardless of the lack of statistical support from the data. We note a lack of clarity about how the 3rd party pre-trained machine learning models perform on the authors' imaging data (validation/monitoring tests are not reported), as well as issues (among others) with the particular implementation of co-immunoprecipitation (ORF1p is not among the highly enriched proteins and apparently does not reach statistical significance for the comparison) - neither of which may be sufficiently rigorous.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In this study, Bonnifet et al. profile the presence of L1 ORF1p in the mouse and human brain. They claim that ORF1p is expressed in the human and mouse brain at a steady state and that there is an age-dependent increase in expression. This is a timely report as two recent papers have extensively documented the presence of full-length L1 transcripts in the mouse and human brain (PMID: 38773348 & PMID: 37910626). Thus, the finding that L1 ORF1p is consistently expressed in the brain is not surprising, but important to document.

Thank you for recognizing the importance of this study. The two cited papers have indeed reported the presence of full-length transcripts in the mouse and human brain. However, the first (PMID: 38773348) report has shown evidence of full-length LINE-1 RNA and ORF1 protein expression in the mouse hippocampus (but not elsewhere) and the second (PMID: 37910626) shows full-length LINE-1 RNA expression and H3K4me3-ChIP data in the frontal and temporal lobe of the human brain, but not protein expression.

Strengths:

Several parts of this manuscript appear to be well done and include the necessary controls. In particular, the evidence for steady-state expression of ORF1p in the mouse brain appears robust.

Weaknesses:

Several parts of the manuscript appear to be more preliminary and need further experiments to validate their claims. In particular, the data suggesting expression of L1 ORF1p in the human brain and the data suggesting increased expression in the aged brain need further validation. Detailed comments:

(1) The expression of ORF1p in the human brain shown in Figure 1j is not convincing. Why are there two strong bands in the WB? How can the authors be sure that this signal represents ORF1p expression and not nonspecific labelling? Additional validations and controls are needed to verify the specificity of this signal.

We have validated the antibody against human ORF1p (Abcam 245249-> https://www.abcam.com/enus/products/primary-antibodies/line-1-orf1p-antibody-epr22227-6-ab245249), which we use for Western blotting experiments (please see Fig1J and new Suppl Fig.2A,B and C), by several means.

(1) We have done immunoprecipitations and co-immunoprecipitations followed by quantitative mass spectrometry (LC-MS/MS; data not shown as they are part of a different study). We efficiently detect ORF1p in IPs (Western blot now added in Suppl Fig2B) and by quantitative mass spectrometry (5 independent samples per IP-ORF1p and IP-IgG: ORF1p/IgG ratio: 40.86; adj p-value 8.7e-07; human neurons in culture; data not shown as they are part of a different study). We also did co-IPs followed by Western blot using two different antibodies, either the Millipore clone 4H1 (https://www.merckmillipore.com/CH/en/product/Anti-LINE-1-ORF1p-Antibody-clone-4H1,MM_NF-MABC1152?ReferrerURL=https%3A%2F%2Fwww.google.com%2F) or the Abcam antibody to immunoprecipitate and the Abcam antibody for Western blotting on human brain samples. Indeed, the Millipore antibody does not work well on Western Blots in our hands. We consistently revealed a double band indicating that both bands are ORF1p-derived. We have added an ORF1p IP-Western blot as Suppl Fig. 2B which clearly shows the immunoprecipitation of both bands by the Abcam antibody. Abcam also reports a double band, and they suspect that the lower band is a truncated form (see the link to their website above). ORF1p Western blots done by other labs with different antibodies have detected a second band in human samples

Sato, S. et al. LINE-1 ORF1p as a candidate biomarker in high grade serous ovarian carcinoma. Sci Rep 13, 1537 (2023) in Figure 1D

McKerrow, W. et al. LINE-1 expression in cancer correlates with p53 mutation, copy number alteration, and S phase checkpoint. Proc. Natl. Acad. Sci. U.S.A. 119, e2115999119 (2022)) showing a Western blot of an inducible LINE-1 (ORFeus) detected by the MABC1152 ORF1p antibody from Millipore Sigma in Figure 7 - Walter et al. eLife 2016;5:e11418. (DOI: 10.7554/eLife.11418) in mouse ES cells with an antibody made inhouse (gift from another lab; in Figure 2B)

The lower band might thus be a truncated form of ORF1p or a degradation product which appears to be shared by mouse and human ORF1p. We have now mentioned this in the revised version of the paper (lines 183-189).

(2) We have used the very well characterized antibody from Millipore ((https://www.merckmillipore.com/CH/en/product/Anti-LINE-1-ORF1p-Antibody-clone-4H1,MM_NFMABC1152?ReferrerURL=https%3A%2F%2Fwww.google.com%2F)) for immunostainings and detect ORF1p staining in human neurons in the very same brain regions (Fig 2H, new Suppl Fig. 2E) including the cerebellum in the human brain. We added a 2nd antibody-only control (Suppl Fig. 2E).

(3) We also did antibody validation by siRNA knock-down. However, it is important to note, that these experiments were done in LUHMES cells, a neuronal cell line which we differentiated into human dopaminergic neurons. In these cells, we only occasionally detect a double band on Western blots, but mostly only reveal the upper band at ≈ 40kD. The results of the knockdown are now added as Suppl Fig. 2C.

Altogether, based on our experimental validations and evidence from the literature, we are very confident that it is indeed ORF1p that we detect on the blots and by immmunostainings in the human brain.

(2) The data shown in Figure 2g are not convincing. How can the authors be sure that this signal controls are needed to verify the specificity of this signal. represents ORF1p expression and not non-specific labelling? Extensive additional validations and

In line 117-123 of the manuscript, we had specified “Importantly, the specificity of the ORF1p antibody, a widely used, commercially available antibody [18,34–38], was confirmed by blocking the ORF1p antibody with purified mouse ORF1p protein resulting in the complete absence of immunofluorescence staining (Suppl Fig. 1A), by using an inhouse antibody against mouse ORF1p[17] which colocalized with the anti-ORF1p antibody used (Suppl Fig. 1B, quantified in Suppl Fig. 1C), and by immunoprecipitation and mass spectrometry used in this study (see Author response image 1)”.

Figure 2G shows a Western blot using an extensively used and well characterized ORF1p antibody from abcam (mouse ORF1p, Rabbit Recombinant Monoclonal LINE-1 ORF1p antibody-> (https://www.abcam.com/enus/products/primary-antibodies/line-1-orf1p-antibody-epr21844-108-ab216324; cited in at least 11 publications) after FACS-sorting of neurons (NeuN+) of the mouse brain. We have validated this ORF1p antibody ourselves in IPs (please see Fig 6A) and co-IP followed by mass spectrometry (LC/MS-MS; see Fig 6, where we detect ORF1p exclusively in the 5 independent ORF1p-IP samples and not at all in 5 independent IgG-IP control samples, please also see Suppl Table 2). In this analysis, we detect ORF1p with a ratio and log2fold of ∞ , indicating that this proteins only found in IP-ORF1p samples (5/5) and not in the IP-control samples ((not allowing for the calculation of a ratio with p-value), please see Suppl Table 2)

In addition, we have added new data showing the entire membrane of the Western blot in Fig1H (now Suppl Fig.1E) and a knock-down experiment using siRNA against ORF1p or control siRNA in mouse dopaminergic neurons in culture (MN9D; new Suppl Fig.1D). This together makes us very confident that we are looking at a specific ORF1p signal. The band in Figure 2G is at the same height as the input and there are no other bands visible (except the heavy chain of the NeuN antibody, which at the same time is a control for the sorting). We added some explanatory text to the revised version of the manuscript in lines 120-124 and lines 253-256).

Please note that in the IP of ORF1p shown in Fig6A, there is a double band as well, strongly suggesting that the lower band might be a truncated or processed form of ORF1p. As stated above, this double band has been detected in other studies (Walter et al. eLife 2016;5:e11418. DOI: 10.7554/eLife.11418) in mouse ES cells using an in-house generated antibody against mouse ORF1p. Thus, with either commercial or in-house generated antibodies in some mouse and human samples, there is a double band corresponding to full-length ORF1p and a truncated or processed version of it.

We noticed that we have not added the references of the primary antibodies used in Western blot experiments in the manuscript, which was now corrected in the revised version.

(3) The data showing a reduction in ORF1p expression in the aged mouse brain is confusing and maybe even misleading. Although there is an increase in the intensity of the ORF1p signal in ORF1p+ cells, the data clearly shows that fewer cells express ORF1p in the aged brain. If these changes indicate an overall loss or gain of ORF1p, expression in the aged brain is not resolved. Thus, conclusions should be more carefully phrased in this section. It is important to show the quantification of NeuN+ and NeuN- cells in young vs aged (not only the proportions as shown in Figure 3b) to determine if the difference in the number of ORF1p+ cells is due to loss of neurons or perhaps a sampling issue. More so, it would be essential to perform WB and/or proteomics experiments to complement the IHC data for the aged mouse samples.

We thank the reviewer for this comment and we agree that the representation has been confusing, which is why we added data to Suppl Fig.5 (F-K) using a different representation. As suggested by the reviewer, in new Suppl Fig. 5F-K, we now show the number of ORF1p+, NeuN+ or NeuN- cells per mm2. These graphs indicate that the number per mm2 of ORF1p+ cells overall do not decrease significantly (with the dorsal striatum as an exception, but possibly due to technical limitations which we now discuss in the results section, line 332-335). Globally, there is thus no loss of ORF1p+ expressing cells. There is also no global nor region-specific decrease in the number of neuronal cells (NeuN+ per mm2) although proportions change (Suppl Fig 2E, confocal acquisitions), thus most likely due to a gain of non-neuronal cells in this region. Concerning Western blots on mouse brain tissues from young and aged individuals, we unfortunately ran into limits regarding tissue availability of aged mice.

(4) The transcriptomic data presented in Figure 4 and Figure 5 are not convincing. Quantification of transposon expression on short read sequencing has important limitations. Longer reads and complementary approaches are needed to study the expression of evolutionarily young L1s (see PMID: 38773348 & PMID: 37910626 for examples of the current state of the art). Given the read length and the unstranded sequencing approach, I would at least ask the authors to add genome browser tracks of the upregulated loci so that we can properly assess the clarity of the results. I would also suggest adding the mappability profile of the elements in question. In addition, since this manuscript focuses on ORF1p, it would be essential to document changes in protein levels (and not just transcripts) in the ageing human brain.

We agree that there are limitations to the analysis of TEs with short read sequencing and we have added more text on this aspect in the revised version (results section) and highlighted the problem of limited and disequilibrated sample size in the discussion (line 638-644). The approaches shown in PMID: 38773348 & PMID: 37910626 or even a combination of them, would be ideal of course. However, here we re-analyzed a unique preexisting dataset (Dong et al, Nature Neuroscience, 2018; http://dx.doi.org/10.1038/s41593-018-0223-0), which contains RNA-seq data of human post-mortem dopaminergic neurons in a relatively high number of brain-healthy individuals of a wide age range including some “young” individuals which is rare in post-mortem studies. Such data is unfortunately not available with long read sequencing or any other more appropriate approach yet. Limitations are evident, but all limitations will apply equally to both groups of individuals that we compare. The general mappability profile of the full-length LINE-1 “UIDs” was shown in old Suppl Fig 6A. We have colorhighlighted now in new Suppl Fig 8C the specific elements in this graph. Most importantly, we have now used, as a condensate of suggestions by all reviewers, a combination of mappability score, post-hoc power calculation, visualization and correlation with adjacent gene expression in order to retain a specific locus with confidence or not. Using these criteria, we retained UID-68 (Fig 5D) which has a relatively high mappability score (Suppl Fig.8C) plus an overlap of umap 50 mappability peaks and read mapping when visualizing the locus in IGV (new Fig. 5E), very high post-hoc power (96.6%; continuous endpoint, two independent samples, alpha 0.05) and no correlation with adjacent gene expression per individual (Fig. 5F, G). Based on these criteria, we had to exclude UID-129, UID-37, UID-127 and UID-137, reinforcing the notion that a combination of quality control criteria might be crucial to retain a specific locus with confidence. This is now mentioned in the manuscript in the discussion in line 427430).

We will not be able to document changes in protein levels in aged human dopaminergic neurons as we do not have access to this material. We have tried to obtain human substantia nigra tissues but were not able to get sufficient amounts to do laser-capture microdissection or FACS analyses, especially of young individuals. There are still important limitations to tissue availability, especially of young individuals, and even more so of specific regions of interest like the substantia nigra pars compacta affected in Parkinson disease.

(5) More information is needed on RNAseq of microdissections of dopaminergic neurons from 'healthy' postmortem samples of different ages. No further information on these samples is provided. I would suggest adding a table with the clinical information of these samples (especially age, sex, and cause of death). The authors should also discuss whether this experiment has sufficient power. The human ageing cohort seems very small to me.

This is a re-analysis of a published dataset (Dong et al, Nat Neurosci, 2018; doi:10.1038/s41593-018-0223-0), available through dbgap (phs001556.v1.p1). In this original article, the criteria for inclusion as a brain-healthy control were as follows:

“…Subjects… were without clinicopathological diagnosis of a neurodegenerative disease meeting the following stringent inclusion and exclusion criteria. Inclusion criteria: (i) absence of clinical or neuropathological diagnosis of a neurodegenerative disease, for example, PD according to the UKPDBB criteria[47], Alzheimer’s disease according to NIA-Reagan criteria[48], or dementia with Lewy bodies by revised consensus criteria[49]; for the purpose of this analysis incidental Lewy body cases (not meeting clinicopathological diagnostic criteria for PD or other neurodegenerative disease) were accepted for inclusion; (ii) PMI ≤ 48 h; (iii) RIN[50] ≥ 6.0 by Agilent Bioanalyzer (good RNA integrity); and (iv) visible ribosomal peaks on the electropherogram. Exclusion criteria were: (i) a primary intracerebral event as the cause of death; (2) brain tumor (except incidental meningiomas); (3) systemic disorders likely to cause chronic brain damage.”

We do not have access to the cause of death, but we have added available metadata as Suppl_Table 5 to the manuscript.

We have performed a post-hoc power analysis (using the “Post-hoc Power Calculator” https://clincalc.com/stats/Power.aspx, which evaluates the statistical power of an existing study and added the results to the revision. Due to this analysis, we have indeed taken out Suppl Fig 7 as a whole which had shown data of three full-length LINE-1 loci (UID-37, UID-127 and UID-137) with low power (between 17-66% power). The locus shown in Fig. 5D of the UID-68) had a post-hoc power score of 96.6% which increases our confidence in this full-length LINE-1 element being upregulated in aged dopaminergic neurons. UID-129 had a post-hoc power score of 97%. However, visualization and mappability analysis of the UID-129 locus led us to exclude this UID.

The post-hoc power analysis for L1HS and L1PA2 revealed a low power (28.4% and 32.8% respectively). We have added these results to the manuscript (line 359-362), but decided to keep the data in as this will hopefully be a motivation for future confirmation studies knowing that the availability of similar data from brain-healthy human dopaminergic neurons especially of young individuals will be low.

(6) The findings in this manuscript apply to both human and mouse brains. However, the landscape of the evolutionarily young L1 subfamilies between these two species is very different and should be part of the discussion. For example, the regulatory sequences that drive L1 expression are quite different in human and mouse L1s. This should be discussed.

Indeed, they are different. We have added a paragraph to the discussion (lines 539-548).

(7) On page 3 the authors write: "generally accepted that TE activation can be both, a cause and consequence of aging". This statement does not reflect the current state of the field. On the contrary, this is still an area of extensive investigation and many of the findings supporting this hypothesis need to be confirmed in independent studies. This statement should be revised to reflect this reality.

We agree, this is overstated, we have changed this sentence accordingly to:

“It is now, 31 years after the initial proposition of the “transposon theory of aging” by Driver and McKechnie [14], still a matter of debate whether TE activation can be both, a cause and a consequence of aging [15,16].”

Reviewer #2 (Public Review):

Summary:

Bonnifet et al. sought to characterize the expression pattern of L1 ORF1p expression across the entire mouse brain, in young and aged animals, and to corroborate their characterization with Western blotting for L1 ORF1p and L1 RNA expression data from human samples. They also queried L1 ORF1p interacting partners in the mouse brain by IP-MS.

Strengths:

A major strength of the study is the use of two approaches: a deep-learning detection method to distinguish neuronal vs. non-neuronal cells and ORF1p+ cells vs. ORF1p- cells across large-scale images encompassing multiple brain regions mapped by comparison to the Allen Brain Atlas, and confocal imaging to give higher resolution on specific brain regions. These results are also corroborated by Western blotting on six mouse brain regions. Extension of their analysis to post-mortem human samples, to the extent possible, is another strength of the paper. The identification of novel ORF1p interactors in the brain is also a strength in that it provides a novel dataset for future studies.

Thank you for highlighting the strength of our study.

Weaknesses:

The main weakness of the study is that cell type specificity of ORF1p expression was not examined beyond neuron (NeuN+) vs non-neuron (NeuN-). Indeed, a recent study (Bodea et al. 2024, Nature Neuroscience) found that ORF1p expression is characteristic of parvalbumin-positive interneurons, and it would be very interesting to query whether other neuronal subtypes in different brain regions are distinguished by ORF1p expression.

We agree that this point is important to address. We have mentioned in the manuscript our previous work, which showed that in the mouse ventral midbrain, dopaminergic neurons (TH+/NeuN+) express ORF1p and that these neurons express higher levels of ORF1p than adjacent non-dopaminergic neurons (TH-/NeuN+; Blaudin de Thé et al, EMBO J, 2018). Others have shown evidence of full-length L1 RNA expression in both excitatory and inhibitory neurons but much less expression in non-neuronal cells (Garza et al, SciAdv, 2023). Further, ORF1p expression was documented in excitatory (CamKIIa-positive) and CamKIIa-negative neurons in the mouse frontal cortex (Zhang et al, Cell Res, 2022, doi.org/10.1038/s41422-022-00719-6). We do detect ORF1p staining in mouse (Fig. 1B, panel 10) and human Purkinje cells (based on morphology and in accordance with data from Takahashi et al, Neuron, 2022; DOI: 10.1016/j.neuron.2022.08.011) and most probably basket cells (based on anatomical location in the molecular layer near Purkinje cells) of the cerebellum (Suppl Fig.4). Some Purkinje cells express PV in mice (https://doi.org/10.1016/j.mcn.2021.103650 and 10.1523/JNEUROSCI.22-1607055.2002), as do stellate and basket cells of the molecular layer (10.1523/JNEUROSCI.22-16-07055.2002). While ORF1p is expressed in PV cells of the hippocampus (Bodea et al, Nat Neurosci, 2024) and in the human and mouse cerebellum in PV-expressing neurons, it does not seem as if ORF1p expression is restricted to PV cells overall. To adress this question experimentally, we have now performed ORF1p stainings in different brain regions (hippocampus, cortex, hindbrain, thalamus, ventral midbrain and cerebellum) together with parvalbumin (PV) stainings and in some cases including the lectin WFA (Wisteria floribunda agglutinin, which specifically stains glycoproteins surrounding PV+ neurons). We have added this data to the manuscript as Suppl Fig.4. While PV-positive neurons often co-stain with ORF1p, not all ORF1p positive cells are PV-positive. We have also deepened the discussion of this aspect in the revised manuscript (line 579-599).

The data suggesting that ORF1p expression is increased in aged mouse brains is intriguing, although it seems to be based upon modestly (up to 27%, dependent on brain region) higher intensity of ORF1p staining rather than a higher proportion of ORF1+ neurons. Indeed, the proportion of NeuN+/Orf1p+ cells actually decreased in aged animals. It is difficult to interpret the significance and validity of the increase in intensity, as Hoechst staining of DNA, rather than immunostaining for a protein known to be stably expressed in young and aged neurons, was used as a control for staining intensity.

We have now separated the analysis of NeuN+, ORF1p+ and NeuN- cells (please see new Suppl Fig5F-K) which highlights the fact that there is indeed no change in the number of ORF1p+ cells in the young compared to the aged mouse brain. However, while neuronal cell numbers throughout the brain do not change significantly (new Suppl Fig.5F), while cell proportions in the ventral midbrain (confocal microscopy based quantifications) change, possibly due to a combination of a slight loss in neurons and a gain in non-neuronal cell numbers (Suppl Fig3E). Please also keep in mind that the ventral midbrain region on images taken on a confocal microscope are a much smaller region than the midbrain motor region as specified by ABBA on images taken by the slide scanner. A different marker than DNA as a control requires the use of a protein that is stably expressed throughout the brain and throughout age. We are not aware of a protein for which this has been established. To nevertheless try to address this issue, we used whole-brain imaging intensity data for the protein Rbfox3 (NeuN) which we originally used as a marker for cell identity. We have now added the quantifications of the protein Rbfox3 (NeuN) to Fig3 (new Fig3B). As shown in this figure, NeuN intensity is not stable from one individual to another, neither in control mice nor in the aged control group. Most importantly, NeuN staining intensity does not increase in aged mice. As we did not use NeuN intensity but presence or absence of NeuN as a marker for cell identity, the instability of NeuN intensity from one individual mouse to another does not have an influence on the data presented in this manuscript. It does indicate however, that the overall increase of ORF1p in aged mice is not a mere reflection of a general decrease in protein turnover. As stated above, the DNA staining with Hoechst controls for technical artefacts. Using Hoechst and NeuN as control, we have thus provided evidence for the fact that the increase in ORF1p intensity per cell is indeed specific for ORF1p. This is now added to the results section (line 299-301).

The main weakness of the IP-MS portion of the study is that none of the interactors were individually validated or subjected to follow-up analyses. The list of interactors was compared to previously published datasets, but not to ORF1p interactors in any other mouse tissue.

As stated in the manuscript, the list of previously published datasets does include a mouse dataset with ORF1p interacting proteins in mouse spermatocytes (please see line 479-480: “ORF1p interactors found in mouse spermatocytes were also present in our analysis including CNOT10, CNOT11, PRKRA and FXR2 among others (Suppl_Table4).”) -> De Luca, C., Gupta, A. & Bortvin, A. Retrotransposon LINE-1 bodies in the cytoplasm of piRNA-deficient mouse spermatocytes: Ribonucleoproteins overcoming the integrated stress response. PLoS Genet 19, e1010797 (2023)). We indeed did not validate any interactors for several reasons (economic reasons and time constraints (post-doc leaving)). However, we feel that the significant overlap with previously published interactors highlights the validity of our data and we anticipate that this list of ORF1p protein interactors in the mouse brain will be of further use for the community.

The authors achieved the goals of broadly characterizing ORF1p expression across different regions of the mouse brain, and identifying putative ORF1p interactors in the mouse brain. However, findings from both parts of the study are somewhat superficial in depth.

This provides a useful dataset to the field, which likely will be used to justify and support numerous future studies into L1 activity in the aging mammalian brain and in neurodegenerative disease. Similarly, the list of ORF1p interacting proteins in the brain will likely be taken up and studied in greater depth.

Reviewer #3 (Public Review):

The question about whether L1 exhibits normal/homeostatic expression in the brain (and in general) is interesting and important. L1 is thought to be repressed in most somatic cells (with the exception of some stem/progenitor compartments). However, to our knowledge, this has not been authoritatively / systematically examined and the literature is still developing with respect to this topic. The full gamut of biological and pathobiological roles of L1 remains to be shown and elucidated and this area has garnered rapidly increasing interest, year-by-year. With respect to the brain, L1 (and repeat sequences in general) have been linked with neurodegeneration, and this is thought to be an aging-related consequence or contributor (or both) of inflammation. This study provides an impressive and apparently comprehensive imaging analysis of differential L1 ORF1p expression in mouse brain (with some supporting analysis of the human brain), compatible with a narrative of non-pathological expression of retrotransposition-competent L1 sequences. We believe this will encourage and support further research into the functional roles of L1 in normal brain function and how this may give way to pathological consequences in concert with aging. However, we have concerns with conclusions drawn, in some cases regardless of the lack of statistical support from the data. We note a lack of clarity about how the 3rd party pre-trained machine learning models perform on the authors' imaging data (validation/monitoring tests are not reported), as well as issues (among others) with the particular implementation of co-immunoprecipitation (ORF1p is not among the highly enriched proteins and apparently does not reach statistical significance for the comparison) - neither of which may be sufficiently rigorous.

Thank you for your comments on our manuscript.

We have addressed the concerns about the machine learning paradigm (see Author response image 1). Concerning the co-IP-MS, we can confirm that ORF1p is among the highly enriched proteins as it was not found in the IgG control (in 5 independent samples), only in the ORF1p-IP (in 5 out of 5 independent samples). This is what the infinite sign in Suppl Table 2 indicates and this is why there is no p-value assigned as infinite/0 doesn’t allow to calculate a pvalue. We have made this clearer in the revised version of the manuscript and added a legend to Suppl Table 2.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

I would recommend the authors remove the human data and expand the analysis of the aged mice. This would most likely result in a much stronger manuscript.

We do think that the imaging data and the Western blots are convincing (please also see our detailed response above to the criticism concerning the antibody we used and the newly added data) and very much reflects what we find in the mouse brain, i.e. concerning the percentage of neurons expressing ORF1p and the percentage of ORF1p+ cells being neuronal. When it comes to the transcriptomic data on aged dopaminergic neurons, we have further discussed the limitations of this study in the revised manuscript and hope that the findings inspire others in the field to redo these types of analyses using the now state-of-the-art NGS technologies to address the question and validate what we have found.

Reviewer #2 (Recommendations For The Authors):

The characterization of ORF1p expression across the mouse brain would be vastly more informative if cell identity was established beyond NeuN+/NeuN---the neuronal predominance of L1 activity in the brain has long been observed. Indeed, even corroboration of the PV+ interneuron signature previously reported would both lend credence to the present study and provide valuable confirmation to the field.

We agree. Please see our response above as well as the new experimental data we added (Suppl Fig5.F-K).

The increased intensity (but not prevalence in terms of % of Orf1p positive cells) of Orf1p expression in aged mouse brains would be more convincing with further context and perhaps better controls. Is overall protein turnover in aged neurons simply slower than in neurons from younger brains? Immunostaining with another protein marker, rather than Hoescht staining of DNA, to demonstrate that increased staining intensity is unique to Orf1p, would make this result more compelling.

To address this question, we have now added the quantifications of the protein Rbfox3 (NeuN) to Fig3 (Fig. 3B). As shown in this figure, NeuN intensity is not stable from one individual to another, neither in control mice nor in the aged control group. As we did not use NeuN intensity but presence or absence of NeuN as a marker for cell identity, this does not have any influence on the data presented in this manuscript. It does indicate however, that the overall increase of ORF1p in aged mice is not a mere reflection of a general decrease in protein turnover. As stated above, the DNA staining with Hoechst controls for technical artefacts. Using Hoechst and NeuN as control, we have thus provided evidence for the fact that the increase in ORF1p intensity per cell is indeed specific for ORF1p.

Western blotting on cell lysates from aged vs young NeunN+ sorted cells would also strengthen this conclusion, although I appreciate the technical challenge of physically isolating whole mature neuronal cells.

Indeed, this would be feasible but only after FACS sorting, which is technically challenging on whole brain cells (less so on nuclei). We unfortunately do not have the possibility to embark on this right now.

Concerning data presentation, Figure 3A would be much more informative if the graph was broken down to show the proportion of ORF1p+ and ORF1p- cells, regardless of NeuN status, and the proportion of NeuN+ and NeuN- cells shown independently of Orf1p status. It is difficult to ascertain the relationship of either of these variables to age, as the graph is presented now.

We followed the suggestions of the reviewer agreeing that breaking down this figure into either ORF1p+ or NeuN+ or NeuN- cells without double attribution is easier to interpret. However, we also chose to use cell densities (cell numbers/ per mm2) to represent the data (new Suppl Fig.5F-K) which is even more precise while proportions are now shown in Suppl Fig.3A-E. Indeed, while it is important to realize that the variables ORF1p+/- or NeuN+/- are not completely independent of each other (as shown in proportions of old Fig4A and B, new Suppl Fig3A and B) as they form four categories (NeuN+/ORF1p+; NeuN+/ORF1p-. NeuN-/ORF1p+, NeuN-/ORF1p-), we can see from the data that there is no overall change in neuron number in the mouse brain between 3 month and 16 months of age. There isn’t an overall change of the density of ORF1p+ cells nor NeuN- cells in the mouse brain with the exception of a decrease in cell density of ORF1p-positive cells in the dorsal striatum accompanied by an increase in non-neuronal cell density (but as discussed above and in the manuscript (line 332-337), this might be due to technical limitations). Thus, while ORF1p intensities per cell increase significantly in older mice, here is no significant change in ORF1p+ cell number.

Reviewer #3 (Recommendations For The Authors):

(1) According to the description in Materials and Methods on the analysis of the confocal images (lines 731-743) the authors used Cell-Pose for both the nuclei and cell segmentation tasks, using model=cyto and diameter=30 for the first (nuclei) and model=cyto2 and diameter=40 for the second (cell). Description of analysis of sagittal brain regions (lines 746-764) indicates the pre-trained model DSB2018 from StarDist 2D was used for nuclei detection, and Cell-Pose using model cyto2 and diameter=30 for cell segmentation. Detected nuclei were then matched to segmented cell areas based on overlap criteria and each nucleus was labeled as 'positive' or 'negative' for either OFR1P or NEU-N.

As described in its three publications (1, 2, 3), Cell-Pose as a segmentation tool is trained in different datasets, with cyto2 being trained on a more varied dataset than cyto. In their library they also offer a model specific for nuclei2. Some description and explanation on the reasons two different models were used for nuclei detection and not choosing the offered specific pre-trained model by Cell-Pose in either case.

According to the cellpose library documentation "Changing the diameter will change the results that the algorithm outputs. When the diameter is set smaller than the true size then cellpose may over-split cells. Similarly, if the diameter is set too big then cellpose may over-merge cells.". It would be useful to offer the justification of the pixels chosen for the analysis (possibly average pixel counts in a subsample of Hoechst images).

Answers to questions 1-5:

Regarding ABBA, slices were first positioned and oriented manually along the Z-axis, without using DeepSlice. Automated affine registration was then applied in the XY plane, followed by manual refinement. 1 slice per mouse brain, 4 mouse brains per condition.

Regarding the gradient heatmap, as stated in the figure legend of Fig3F; Represented is the fold-change in percent (aged vs young) of the “mean of the mean” ORF1p expression per ORF1p+ cell quantified mapped onto the nine different regions analyzed. More precisely, the heatmap shows the percentage increase in the mean of all mean cell intensities in the aged condition, normalized to the mean of all mean cell intensities in the young condition. The pre-trained models and hyperparameters were selected based on their optimal performance across our image datasets. For slide scanner images, the StarDist DSB 2018 model was chosen over a Cellpose model because it more effectively avoided detecting out-of-focus nuclei, which were common in slide scanner images due to the lack of optical sectioning. This issue was not present in confocal images, where Cellpose cyto model was used instead. To assess the performance of each model and diameter setting, we computed the average precision (AP) metric, which is defined as AP = TP/(TP+FP+FN), where TP = true positives, FP = false positives, and FN = false negatives. The AP was calculated at the commonly used Intersection over Union (IoU) threshold of 0.5. For confocal images, Cellpose models and hyperparameters were evaluated on eight images per channel, capturing intensity variability across different mouse ages and brain regions. A total of approximately 2,000 nuclei and 1,000 NeuN and ORF1p cells were manually annotated. The AP values at an IoU threshold of 0.5 were: 0.995 for nuclei, 0.960 for NeuN, and 0.974 for ORF1p cells. These high AP values confirm that the selected models and diameter settings were well-suited for analyzing the entire dataset. For slide scanner images, nuclei and cell detection were evaluated on 14 images per channel, with approximately 800 nuclei and 400 NeuN and ORF1p cells manually annotated. The AP values were lower compared to confocal images, mainly due to a lower signal-to-noise ratio, which led to an increased number of false positives and false negatives: 0.806 for nuclei, 0.675 for NeuN, and 0.695 for ORF1p cells. This decline in performance was expected given the challenges posed by slide scanner images, including background noise and out-of-focus objects. Notably, the observed false positives primarily correspond to small-sized nuclei/cells or those with low intensity, which evade the stringent filters that were applied. While fine-tuning the models could further enhance detection robustness, we considered that the selected models and diameter settings were suitable for processing the entire dataset.

We added a paragraph to the materials & methods section with this new information; for confocal images (line 847-855), slide scanner images (line 878-885).

(2) Next to no information is offered regarding the brain segment registration and how the results were analyzed: The ABBA plug-in has two modules manual and automatic, via a DL pre-trained model called DeepSlice. The authors should report which mode of ABBA they used, how many slices per mouse brain, and how many brains. Moreover, there is no explanation of how the gradient heatmap of the brain regions (Figure 3G) was calculated.

Please see above

(3) Even the best algorithms produce some False predictions. In this application of the (3rd party) cellpose, StarDist, and ABBA pre-trained models, such cases of wrong predictions would have amplified downstream effects on the analysis e.g., wrongly characterizing certain cells as 'negative' (falsely not detected cell, falsely detected nucleus), or worse, biasing against certain cell subgroups (falsely not detected 'type' of nuclei). This is even more troubling with the variety of models used for the nuclei segmentation task, and the parameters in each. It is possible the authors performed optimizations and reported exactly such optimized values for their dataset, they should however still explicitly offer these detailed validation and optimization processes. The low statistical significance throughout the quantified results from these IF experiments (Figures 1-3) is also a cause for needing an explicit description of how these algorithms perform on the authors' data.

It is good practice that a pre-trained model when applied to a new dataset like the one that the authors produced for this work, would require basic monitoring for how it performs in the new, previously unseen dataset, even when the model's generalizability has been reported previously as great. It would be best if the authors had handannotated a few images as the validation set and produced some model performance metrics as a supplemental table for all pre-trained models they used, in the datasets they used them at. Alternatively, the authors are offered the ability by the cellpose team to fine-tune the model for their data, and this could be used to perform the experiments for this work instead if the performance metrics of the used cellpose (cyto and cyto2) models prove to be poor.

Please see above

(4) The legend for Figure 1A indicates that Cell-Pose was used for cell detection and StarDist for nuclei detection in the confocal images (line 960). This needs clarification and correction.

Please see above

(5) Some explanation of why the models used were changed when using confocal or the slide scanner microscope would be nice.

Please see above

(6) The legend title of Figure 3 (line 1040) "Fig. 3: ORF1p expression is increased throughout the whole mouse brain in the context of aging" is misleading as half the panels in the figure demonstrate a decrease in ORF1pexpressing cells. The two can be both true, but in a more nuanced relationship. A more modest representation of the data in the title is also warranted by the unimpressive statistical significance achieved (notably with no correction for multiple testing, which would further inflate them).

We have toned down the tile of Fig. 3 to “ORF1p expression is increased in some regions of the aged mouse brain” while leaving its meaning as globally. There is indeed no significant loss of ORF1p expressing cells (Suppl Fig. 5F; except in the dorsal striatum (Supl Fig. 5I, please see also discussion above), but there is a significant increase in ORF1p intensity per cell overall (Fig. 3A,C,F) and in several regions of the mouse brain (Fig E, G and H).

(7) Figure 4 suffers for significance. For example in panel A, the few genes with the highest -log10P value, ie above 1.3 (p-value of ~0.05) have a log2-fold change of 0.2-0.3 (fold change 1.14-1.23). There are no hits with even the modest log2-fold change of 0.5 (fold-change 1.4). The big imbalance between young/old samples for these RNA seq experiments (6 vs 36 mice) could be an issue here too.

The reviewer refers to mouse samples (“6 to 36 mice”), but this is data of human post-mortem dopaminergic neurons from brain-healthy individuals which were laser-captured and sequenced as reported by Dong et al, Nat Neurosci, 2018. There is indeed a big imbalance between young and old samples which are linked to the difficulties in availability of brain-healthy post-mortem tissues from young individuals which are obviously much rarer than from older people. We agree that the fold-enrichment are modest and p-values rather high, but we argue to keep this data in as it is based on rare post-mortem human brain tissues which were difficult to obtain and will be very difficult to obtain in sufficient number in future studies. We hope however, that these results will encourage such studies in the future and motivate researchers to further look into the expression of TEs in aging brain tissues with higher sample sizes and more suitable sequencing techniques. We have now in the revised version toned down some sentences (i.e. line 359: modest, but significant increase in several young…) and have now also added a post-hoc power analysis (results section line 359-362: “There was a modest but significant increase in several younger LINE-1 elements including L1HS and L1PA2 at the “name” level (Fig. 4A, B), an analysis which was however underpowered (post-hoc power calculation; L1HS: 28.4%; L1PA2: 32.8%) and thus awaits further confirmation in independent studies.”)

(8) Figure legend 4C (line 1088) should offer more explanation on what is compared for these correlations: the young vs old results, all intensities of all experiments, and intensities separately for each sample.

We have added the missing information to Figure legend 4C (line 1209-1215): “Correlation of the RNA expression levels of LINE-1 elements with known transposable element regulators in human dopaminergic neurons (all ages included). What was compared are the expression levels of LINE-1 elements with known regulators of TEs for each individual sample, all ages included.”

(9) Figure 5, panel D. The regressions are all driven by 1-2 outliers. Should be removed as they don't add anything.

We agree and therefore have performed an outlier test (ROUT (Q=1%) and identified outliers (1 in each graph) have been taken out from the analysis. We argue that the information of a non-correlation of UID-68 and adjacent gene expression is important as it rules out a dependency of expression of the full-length LINE-1 depending on neighboring gene expression (see new Fig5E-G).

(10) Figure 6 panel B. It is unexpected that the GO terms with the highest enrichment also show weak significance and vice-versa. Fold enrichment in the PANTHER tool is defined as the % of GO-term genes in the sample divided by the %GO-term genes in the background (organism).

This is not unexpected as GO terms contain different numbers of proteins. Indeed, the significance can be different if the GO term contains for example 3 or 300 proteins. A GO term containing only few proteins with a high fold change between the conditions (here: ORF1p-IP vs whole mouse genome) will lead to a rather low significance for example. If you look at the last 6 categories in Fig 6B, you can appreciate that they have very similar values for enrichment but very different significance levels (FDR).

(11) Many citations in the References sections are referred to by doi and "Published online" date. These should be corrected to include the citation in standard format (journal name, volume, issue, pages, etc).

We apologize for this and have corrected this in the revised version.

(12) (line 970) Legend of Figure 1 is missing label referencing panel C (ie (C) Bar plot showing the total....).

Thank you for pointing this out, this has been corrected.

(13) The bottom violin plot in Figure 1C lacks sufficient explanation (what are the M1-4 categories?). The same problem with panel G (same Figure 1).

This has now been better explained. The M1-M4 categories denominate individual mice numbered from 1 to 4 for (results are shown per individual).

-> specified in line 1098-1099 (Fig.1C) and new text (1117-1118: Fig.1G): Four three-month-old Swiss/ OF1 mice (labeled as M1 to M4) are represented each by a different color, the scattered line represents the median. ****p<0.0001, nested one-way ANOVA. Total cells analyzed = 4645

(14) Figure 1B; confocal image 2 (Hippocampus) does not seem to tell the same story as the main slide scanner image. Overall, more explicit phrasing regarding how the Images in Figure 1B are not blow-outs of the bigger one but different, confocal images of the same regions.

We have changed the sentence to: “Representative images acquired on a confocal microscope of immunostainings showing ORF1p expression (orange) in 10 different regions of the mouse brain.”, which hopefully helps to indicate that these images are indeed not blow-outs of the slide scanner image.

(15) Young are defined as 3 months and 'old' as 16 months mice. 16-month group name would be better as "adults". Example of age range considered 'old': "Young (3-6-month-old) and aged (18-27-month-old) male mice were age- and source-matched for each experiment." https://www.cell.com/cell-metabolism/fulltext/S1550-4131(23)00462X?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS15504131230 0462X%3Fshowall%3Dtrue

This is true, but the 16-month age group does not have a designation when looking at Mouse Life history stages in C57Bl/6 mice from the Jackson laboratory (see https://www.jax.org/news-and-insights/jax-blog/2017/november/when-are-mice-considered-old#), they are neither middle-aged nor old. We therefore believe that the designation as “aged” still holds true.

(16) Lines 63-65 > To our understanding, both ORF1 and ORF2 proteins are thought to exhibit cis preference.

Yes, that is true, but the sentence as it is does not make a claim about ORF2p not having cis-preference.

(17) Figure 1I is only referred to as "Figure I". Twice. Page 8, line 173 & 176.

Thank you, has been corrected.

(18) Lines 178-182 >To investigate intra-individual expression patterns of ORF1p in the post-mortem human brain, we analyzed three brain regions of a neurologically healthy individual (Figure 1J) by Western blotting. ORF1p was expressed at different levels in the cingulate gyrus, the frontal cortex, and the cerebellum underscoring a widespread expression of human ORF1p across the human brain." > It is difficult for us to gauge how believable the blots are without knowing the amount of protein loaded.

We have loaded 10ug of tissue lysate per lane (tissue pulverized with a Covaris Cryoprep; amount now mentioned in the materials & methods section). We have added some more information on the antibody in the revised manuscript (line 183-194).

We say this from our experience conducting similar blots of anti-ORF1p IPs from human brain tissues using the same antibody (4H1) without successful detection of enriched protein by western blot (of course there can be many reasons for that, but knowing the amount of protein loaded is important for reproducibility). In addition, we find the "double" ORF1p bands they see in almost every blot atypical.

In our hands, the 4H1 antibody does not work well on Western blots, but it immunoprecipitates well and works very well on immunostainings. However, the abcam AB 245249 works well for Western blotting (and IPs) which is why we used this antibody for these applications, respectively. As described above, there is evidence that the double band is not atypical, but rather frequent, which we now also mention in the revised manuscript line 183191: “To investigate intra-individual expression patterns of ORF1p in the post-mortem human brain, we analyzed three brain regions of a neurologically-healthy individual (Fig. 1J, entire Western blot membrane in Suppl Fig. 2A) by Western blotting using a commercial and well characterized antibody which we further validated by several means. The double band pattern in Western blots has been observed in other studies for human ORF1p outside of the brain (Sato et al, SciRep, 2023, McKerrow et al, PNAS, 2022) as well as for mouse ORF1p (Walter et al, eLife, 2016). We also validated the antibody by immunoprecipitation and siRNA knock-down in human dopaminergic neurons in culture (differentiated LUHMES cells, Suppl Fig. 2B and 2C) where we detect however in most cases the upper band only. The nature of the lower band is unknown, but might be due to truncation, specific proteolysis or degradation. ORF1p was expressed at different levels in the human post-mortem cingulate gyrus, the frontal cortex and the cerebellum underscoring a widespread expression of human ORF1p across the human brain. This was in accordance with ORF1p immunostainings of the human post mortem cingulate gyrus (Fig. 2H and Suppl Fig. 2E) and frontal cortex (Suppl Fig. 2E), with an absence of ORF1p staining when using the secondary antibody only (Suppl Fig. 2E).”

In some images a band is labeled as IgG heavy chain (e.g. presumably from the FACS, Figure 2G, and IP, Figure 6A - which could contain residual antibody) - however, this is avoidable by using a different antibody for capture than detection - which also helps reduce false positive results.

Unfortunately, we have only an antibody raised in rabbit available to perform IPs and Western blots on mouse tissues and therefore cannot avoid the detection of the IgG heavy chain.

Aside from these, there seem to be persistent 'double bands' in the region of ORF1p. Generally, we are unaccustomed to seeing such 'double bands' in human anti-ORF1p western blots and IP-western blots, and since, in this study, this is seen in both mouse and human blots, it raises some doubts. Having the molecular mass ladder on each blot to at least allow for the assessment of migration consistency and would therefore be very helpful.

We have added the molecular weights on the Western blots (Fig.1H, Fig. 2G and Suppl Fig.1D and E). As discussed also above, there is accumulating evidence that in some tissues, there are persistent double bands detected using ORF1p antibodies in both, mouse and human tissues.

Human ORF1p detection:

We have validated the antibody against human ORF1p (Abcam 245249-> https://www.abcam.com/enus/products/primary-antibodies/line-1-orf1p-antibody-epr22227-6-ab245249), which we use for Western blotting experiments (please see Fig1J and new Suppl Fig.2A,B and C), by several means.

(1) We have done immunoprecipitations and co-immunoprecipitations followed by quantitative mass spectrometry (LC-MS/MS; data not shown as they are part of a different study). We efficiently detect ORF1p in IPs (Western blot now added in Suppl Fig2B) and by quantitative mass spectrometry (5 independent samples per IP-ORF1p and IP-IgG: ORF1p/IgG ratio: 40.86; adj p-value 8.7e-07; human neurons in culture; data not shown as they are part of a different study). We also did co-IPs followed by Western blot using two different antibodies, either the Millipore clone 4H1 (https://www.merckmillipore.com/CH/en/product/Anti-LINE-1-ORF1p-Antibody-clone- 4H1,MM_NF-MABC1152?ReferrerURL=https%3A%2F%2Fwww.google.com%2F) or the Abcam antibody to immunoprecipitate and the Abcam antibody for Western blotting on human brain samples. Indeed, the Millipore antibody does not work well on Western Blots in our hands. We consistently revealed a double band indicating that both bands are ORF1p-derived. We have added an ORF1p IP-Western blot as Suppl Fig. 2B which clearly shows the immunoprecipitation of both bands by the Abcam antibody. Abcam also reports a double band, and they suspect that the lower band is a truncated form (see the link to their website above). ORF1p Western blots done by other labs with different antibodies have detected a second band in human samples

Sato, S. et al. LINE-1 ORF1p as a candidate biomarker in high grade serous ovarian carcinoma. Sci Rep 13, 1537 (2023) in Figure 1D

McKerrow, W. et al. LINE-1 expression in cancer correlates with p53 mutation, copy number alteration, and S phase checkpoint. Proc. Natl. Acad. Sci. U.S.A. 119, e2115999119 (2022)) showing a Western blot of an inducible LINE-1 (ORFeus) detected by the MABC1152 ORF1p antibody from Millipore Sigma in Figure 7 - Walter et al. eLife 2016;5:e11418. (DOI: 10.7554/eLife.11418) in mouse ES cells with an antibody made inhouse (gift from another lab; in Figure 2B)

The lower band might thus be a truncated form of ORF1p or a degradation product which appears to be shared by mouse and human ORF1p. We have now mentioned this in the revised version of the paper (lines 183-189).

(2) We have used the very well characterized antibody from Millipore ((https://www.merckmillipore.com/CH/en/product/Anti-LINE-1-ORF1p-Antibody-clone-4H1,MM_NF-MABC1152?ReferrerURL=https%3A%2F%2Fwww.google.com%2F)) for immunostainings and detect ORF1p staining in human neurons in the very same brain regions (Fig 2H, new Suppl Fig. 2E) including the cerebellum in the human brain. We added a 2nd antibody-only control (Suppl Fig. 2E).

(3) We also did antibody validation by siRNA knock-down. However, it is important to note, that these experiments were done in LUHMES cells, a neuronal cell line which we differentiated into human dopaminergic neurons. In these cells, we only occasionally detect a double band on Western blots, but mostly only reveal the upper band at ≈ 40kD. The results of the knockdown are now added as Suppl Fig. 2C.

Altogether, based on our experimental validations and evidence from the literature, we are very confident that it is indeed ORF1p that we detect on the blots and by immmunostainings in the human brain.

Mouse ORF1p detection: In line 117-123 of the manuscript, we had specified “Importantly, the specificity of the ORF1p antibody, a widely used, commercially available antibody [18,34–38], was confirmed by blocking the ORF1p antibody with purified mouse ORF1p protein resulting in the complete absence of immunofluorescence staining (Suppl Fig. 1A), by using an inhouse antibody against mouse ORF1p[17] which colocalized with the anti-ORF1p antibody used (Suppl Fig. 1B, quantified in Suppl Fig. 1C), and by immunoprecipitation and mass spectrometry used in this study (see Author response image 1)”.

Figure 2G shows a Western blot using an extensively used and well characterized ORF1p antibody from abcam (mouse ORF1p, Rabbit Recombinant Monoclonal LINE-1 ORF1p antibody-> (https://www.abcam.com/enus/products/primary-antibodies/line-1-orf1p-antibody-epr21844-108-ab216324; cited in at least 11 publications) after FACS-sorting of neurons (NeuN+) of the mouse brain. We have validated this ORF1p antibody ourselves in IPs (please see Fig 6A) and co-IP followed by mass spectrometry (LC/MS-MS; see Fig 6, where we detect ORF1p exclusively in the 5 independent ORF1p-IP samples and not at all in 5 independent IgG-IP control samples, please also see Suppl Table 2). In this analysis, we detect ORF1p with a ratio and log2fold of ∞ , indicating that this proteins only found in IP-ORF1p samples (5/5) and not in the IP-control samples ((not allowing for the calculation of a ratio with p-value), please see Suppl Table 2)

In addition, we have added new data showing the entire membrane of the Western blot in Fig1H (now Suppl Fig.1E) and a knock-down experiment using siRNA against ORF1p or control siRNA in mouse dopaminergic neurons in culture (MN9D; new Suppl Fig.1D). This together makes us very confident that we are looking at a specific ORF1p signal. The band in Figure 2G is at the same height as the input and there are no other bands visible (except the heavy chain of the NeuN antibody, which at the same time is a control for the sorting). We added some explanatory text to the revised version of the manuscript in lines 120-124 and lines 253-256).

Please note that in the IP of ORF1p shown in Fig6A, there is a double band as well, strongly suggesting that the lower band might be a truncated or processed form of ORF1p. As stated above, this double band has been detected in other studies (Walter et al. eLife 2016;5:e11418. DOI: 10.7554/eLife.11418) in mouse ES cells using an in-house generated antibody against mouse ORF1p. Thus, with either commercial or in-house generated antibodies in some mouse and human samples, there is a double band corresponding to full-length ORF1p and a truncated or processed version of it.

We noticed that we have not added the references of the primary antibodies used in Western blot experiments in the manuscript, which was now corrected in the revised version.

(19) Figure 1H, 1J, 6A: Show/indicate molecular weight marker.

The molecular weight markers were added (please see Fig.1H, Fig. 2G and Suppl Fig.1D and E).

(20) Page 10, line 223. " ...expressing ORF1p and ORF1p"?

Thank you, this was corrected.

(21) Lines 279-280 "An increase of ORF1p expression was also observed in three other regions albeit not significant." > This means it is not distinguishable as a change under the assumptions and framework of the analysis; please remove this statement.

We agree, we removed this sentence.

(22) Page 13, line 301. Labeling the group with a mean age of 57.5 as "young" might be a bit misleading.

This is why we put the “young” in quotation marks.

(23) Lines 309-311 "however there was a significant increase in several younger LINE-1 elements including L1HS and L1PA2 at the "name" level (Figure 4A, B)". > Effect size is tiny; is this really viable as biologically significant? Maybe just remove the volcano plot? Does panel A add anything not covered by B?

We would like to keep the Volcano plot, even though effect sizes are small (which we acknowledge in the manuscript line 359-362: “There was a modest but significant increase in several younger LINE-1 elements including L1HS and L1PA2 at the “name” level (Fig. 4A, B), an analysis which was however underpowered (posthoc power calculation; L1HS: 28.4%; L1PA2: 32.8%) and thus awaits further confirmation in independent studies.” The reason for this decision is to illustrate a general increase in expression (even with a small effect size) of several LINE-1 elements at the name level with the youngest LINE-1 elements being amongst those with the highest effect.

(24) Lines 327-328 "The transcripts of these genes showed, although not statistically significant, a trend for decreased expression in the elderly (Supplementary Figure 5D-G). > I do not recommend doing this.

We agree and take it out.

(25) Lines 339-342 "While several tools using expectation maximization algorithms in assigning multi-mapping reads have been developed and successfully tested in simulations 48,54, we used a different approach in mapping unique reads to the L1Base annotation of full-length LINE-1" > Generally, this section is not clear - what is the rationale for the approach (compared to the stated norms)? Ideally, justify this analytical choice and provide a basic comparison to other more standard approaches (even if briefly in a supplement).

We thank the reviewer for his comment. Indeed, randomly assigning multi-mapping reads is usually a good strategy to quantify the expression of repeats at the family level (Teissandier et al. 2019) which we did in the first part of the analysis (class, family and name level). However, our main goal was to focus on specific single fulllength LINE elements which can encode ORF1p. We therefore decided to only use uniquely mapped reads, which is by definition the only way to be sure that a sequencing read really comes from a specific genomic location, and which will to not over-estimate their expression level. In this sense, we have added some explanatory text to this specific section. We also added a section to the discussion (line 638-644): This analysis has technical limitations inherent to transcriptomic analysis of repeat elements especially as it is based on short-read sequences and on a limited and disequilibrated number of individuals in both groups. Nevertheless, we tried to rule out several biases by demonstrating that mappability did not correlate with expression overall and used a combination of visualization, post-hoc power analysis and analysis of the mappability profile of each differentially expressed fulllength LINE-1 locus.

(26) Page 16, line 389. The age span covered is 59 years although the difference in mean age between the two groups is only 25.5 years - please indicate both metrics.

We have added this additional metric in line 432.

(27) Lines 394-397 "Further, correlation analyses suggest that L1HS expression might possibly be controlled by the homeoprotein EN1, a protein specifically expressed in dopaminergic neurons in the ventral midbrain 50, the heterochromatin binding protein HP1, two known regulators of LINE-1, and the DNA repair proteins XRCC5/6." > This reads like a drastic reach unless framed explicitly as a 'tempting speculation' (or similar). I don't think this claim should be made as it is without further validation.

We believe to have used careful language (“correlation analysis suggests”.“might possibly be controlled”) in the results section as well as in the discussion (line 660-671): “Matrix correlation analysis of several known LINE-1 regulators, both positive and negative, revealed possible regulators of young LINE-1 sequences in human dopaminergic neurons. Despite known and most probable cell-type unspecific regulatory factors like the heterochromatin binding protein CBX5/HP1 [51] or the DNA repair proteins XRCC5 and XRCC6 [49], we identified the homeoprotein EN1 as negatively correlated with young LINE-1 elements including L1HS and L1PA2. EN1 is an essential protein for mouse dopaminergic neuronal survival [50] and binds, in its properties as a transcription factor, to the promoter of LINE-1 in mouse dopaminergic neurons [17]. As EN1 is specifically expressed in dopaminergic neurons in the ventral midbrain, our findings suggests that EN1 controls LINE-1 expression in human dopaminergic neurons as well and serves as an example for a neuronal sub-type specific regulation of LINE-1.” To this we added: “Although these proteins are known regulators of LINE-1, this correlative relationship awaits experimental validation.”

(28) Mouse protein/gene names are all capital letters on page 17/18. Changes on page 18/19. This should be consistent.

Thank you, this has been corrected (all capital).

(29) Page 23, line 559. The estimated ORF1p/ORF2p ratio referenced is based on an overexpression of L1 from a plasmid (ref87). > It should be made clear to the reader that it is still unknown whether such a ratio is representative of native conditions.

OK, this is indeed true. Thank you for pointing this out. (line 621-622)

(30) Lines 613-616 "Further, GO term analysis contained expected categories like "P-body", mRNA metabolism related categories, and "ribonucleoprotein granule". We also identified NXF1 as a protein partner of ORF1p, a protein found to interact with LINE-1 RNA related to its nuclear export 89." > There is no reason to speculate that the proteins in the pulldown are specific to L1 RNAs.

We did not speculate that the proteins in the pulldown are specific to LINE-1 RNA. We just mentioned that NXF1 was an ORF1p protein partner and that it had been found previously as a LINE-1 RNA interactor.

ORF1p is present in large heterogeneous assemblies - not every protein should be assigned an L1-related function and many proteins will be participating in general RNA-granule functions (given L1 ORFs are known to accumulate in such structures). Moreover, the granules are not the same in every cell type. IP is done in low salt and overnight incubation (poorly controlled for non-specific accumulation).

We state that these key interactors are “probably” essential for completing or repressing the LINE-1 life cycle. It is true that we cannot affirm this. We therefore added a sentence to the discussion (line 679): “This supports the validity of the list of ORF1p partners identified, although we cannot rule out the possibility that unspecific protein partners might be pulled down due to colocalization in the same subcellular compartment.”

(31) Lines 629-631" These results complete the picture of the post-transcriptional and translational control of ORF1p and suggest that these mechanisms, despite a steady-state expression, are operational in neurons." > Stating that these results complete the picture, which is still very much open for completion (granted, these results add to the picture), is an unneeded over-reach.

We agree. We changed “complete” to “add to “ the picture.

(32) Lines 641-644 "Finally, we found components of RNA polymerase II and the SWI/SNF complex as partners of ORF1p. This further indicates that ORF1p has access to the nucleus in mouse brain neurons as described for other cells 95,96, implying that ORF1p potentially has access to chromatin." > There is no way to know if this is a post-lysis effect - we have no real specificity information. The mock IP control is insufficient for this conclusion without further validation.

We added: “however a bias due to a post-lysis effect cannot be excluded.” Line 711

(33) ab216324 for IF and ab245122 for IP - why? What is the difference? Both are rated equally for IF and IP - please provide a rationale for reagent selection and use.

These two antibodies are the same except their storage buffer. ab245122 is azide and BSA-free, while ab216324 contains the preservative sodium azide (0.01%) and the following constituents: PBS, 40% Glycerol (glycerin, glycerine), 0.05% BSA. As azide and BSA can affect coupling of antibodies to beads, antibodies which do not contain these components in their buffer are preferred for IPs (but can be stored less long).

(34) Page 35, line 862. "1.3 x 105" should be "1.3 x 105".

We added a regular x but we are not sure if this is what the reviewer was referring to ?

(35) MS comparison in Figure 6. Why is the comparison not being made between young vs. old brain/neurons? This would be more informative instead of just showing what they IP over a mock IgG control and the comparison would track better with other experiments in the rest of the paper.

Yes, that is true. However, we did not do this at the time as we did not have old mouse brain tissue available. Services from official animal providers in France have unfortunately only recently expanded their offer with regard to the availability of aged animals.

(36) Supplementary Table 2 (MS data) is lacking information. How many peptides (unique/total) were discovered for each protein? Why are all ratios and p-values not listed for every protein in the table? LFQ protein intensity values should also be listed. Each supplementary table should have a legend as a separate tab in the document.

As stated in the SupplTable2 and now made clearer in an independent tab file in SupplTable2 which contains a legend to the table, some proteins do not have associated p values and ratios as these proteins are found only in the ORF1p IP and not in the IgG control. This is why these proteins have an indefinite sign instead of a foldenrichment and no p-value assigned as we cannot calculate a ratio with X/0 which again makes it impossible to obtain a p-value. Concerning the absence of LFQ protein intensity values, as stated in the materials & methods section, we did not use these values (linear model) but instead the intensity values of the peptides: “The label free quantification was performed by peptide Extracted Ion Chromatograms (XICs), reextracted by conditions and computed with MassChroQ version 2.2.21 109. For protein quantification, XICs from proteotypic peptides shared between compared conditions (TopN matching) with missed cleavages were used. Median and scale normalization at peptide level was applied on the total signal to correct the XICs for each biological replicate (n=5). To estimate the significance of the change in protein abundance, a linear model (adjusted on peptides and biological replicates) was performed, and p-values were adjusted using the Benjamini–Hochberg FDR procedure.”

The number of peptides unique/total for each protein has been added to Suppl_Table2 along other available information.

(37) Poor overlap in 6C could in part be explained by the use of different sample/tissue types, but more likely the big difference could come from the very different conditions at which the IPs were performed (buffers and incubation times etc.).

The overlap seems poor, but nevertheless is bigger as by chance (representation factor 2.6, p<5.4e-08). We agree that this can be in part explained by different experimental conditions which we now added to the discussion (line 478: “However, differences in experimental conditions could also influence this overlap.”)

(38) Figure 6D is a very uninspiring representation of the data. What is the point of showing several binary interactions? Was the IgG control proteome also analyzed? Have proteins displayed in Figure 6 been corrected for that?

The point of showing these interactions is that OFR1p interacts with clustered proteins. ORF1p interacts with proteins that belong to specific GO terms (Fig6b), but these proteins are also interacting with each other more than expected (Fig6C). This is the benefit of showing a STRING (Search Tool for the Retrieval of Interacting Genes/Proteins) representation, which is a database of known and predicted protein–protein interactions. Indeed, proteins in Fig6 have been corrected for the IgG proteome. We only show proteins that were enriched or uniquely present in the ORF1p IP condition compared to the IgG control (please see Suppl_Table2).

Version 2

Abstract

Abstract

Recent studies have established a reciprocal causal link between aging and the activation of transposable elements, characterized in particular by a de-repression of LINE-1 retrotransposons. These LINE-1 elements represent 21% of the human genome, but only a minority of these sequences retain the coding potential essential for their mobility. LINE-1 encoded proteins can induce cell toxicity implicated in aging and neurodegenerative diseases. However, our knowledge of the expression and localization of LINE-1-encoded proteins in the central nervous system is limited. Using a novel approach combining atlas-based brain mapping with deep-learning algorithms on large-scale pyramidal brain images, we unveil a heterogeneous, neuron-predominant and widespread ORF1p expression throughout the murine brain at steady-state. In aged mice, ORF1p expression increases significantly which is corroborated in human post-mortem dopaminergic neurons by an increase in young LINE-1 elements including those with open reading frames. Mass spectrometry analysis of endogenous mouse ORF1p revealed novel, neuron-specific protein interactors. These findings contribute to a comprehensive description of the dynamics of LINE-1 and ORF1p expression in the brain at steady-state and in aging and provide insights on ORF1p protein interactions in the brain.

Introduction

Only about 2 % of the human genome are DNA sequences that will be translated into protein. The remaining 98 % are comprised of introns, regulatory elements, non-coding RNA, pseudogenes and repetitive elements including transposable elements. However, some sequences in what is generally considered “non-coding genome” do in fact contain sequences which encode proteins. This is true for specific lncRNAs which can encode peptides or functional proteins 1 but also for a few copies of two transposable element families, Long INterspersed Element-1 (LINE-1) and Human Endogenous RetroViruses (HERV). Non-functional copies of retrotransposons, to which LINE-1 and HERV belong, cover about 44% 2 of the human genome as remnants of an evolutionary ancient activity. Depending on the source, about 100 3 to 146 4 full-length LINE-1 elements with two open reading frames encoding ORF1 and ORF2 are present in the Human reference genome (GRCh38 Genome Assembly) and several incomplete HERV sequences encoding either or any combination of envelope (env), gag, pro or pol 5. The LINE-1 encoded protein ORF1p, an RNA binding protein with “cis” preference 6,7, and ORF2p, an endonuclease and reverse transcriptase 8,9 are required for the mobility of LINE-1 elements. As many other transposable elements (TEs) including HERVs, LINE-1 elements are repressed by multiple cellular pathways. It was thus generally thought that TEs are repressed in somatic cells with no expression at steady-state 10–12. However, the aging process reduces the reliability of these repressive mechanisms 13. It is now, 31 years after the initial proposition of the “transposon theory of aging” by Driver and McKechnie 14, still a matter of debate whether TE activation can be both, a cause and a consequence of aging 15,16.

Sparse data has shown that the LINE-1 encoded protein ORF1p is expressed at steady-state in the mouse ventral midbrain 17, the mouse hippocampus 18 and in some regions of the human post-mortem brain 19 and recent data informed about the presence of full-length transcripts in cancer cells, human epithelial cells and mouse hippocampal neurons 20. Repression of LINE-1 might thus be incomplete and if so, it remains unclear how cells then prevent cell toxicity associated with LINE-1 encoded protein activity. Indeed, LINE-1 encoded proteins have been demonstrated to induce genomic instability (ORF2p endonuclease-mediated 17,21–26) and inflammation (ORF2p reverse transcriptase-mediated 27–29) and these cellular activities might be causally related to organismal aging, cancer, autoimmune and neurological diseases 30. For instance, LINE-1 activation can drive neurodegeneration of mouse dopaminergic neurons 17, of drosophila neurons 31,32 and of mouse Purkinje neurons 33 which can be at least partially rescued with nucleoside analogue reverse transcriptase inhibitors (NRTIs) or other anti-LINE-1 strategies. NRTIs are currently being tested in several clinical trials designed to target either the RT of HERVs or the RT encoded by the LINE-1 ORF2 protein. It is not known today, however, to which extend LINE-1 encoded proteins are expressed at steady-state throughout the mouse and human brain, whether there is cell-type specificity and whether activation of LINE-1 encoded proteins is associated with brain aging or human neurodegeneration. Here, using a deep-learning assisted cellular detection methodology applied to pyramidal large-scale images of the mouse brain mapped to the Allen mouse brain atlas combined with post-mortem human brain imaging, co-IP mass spectrometry and transcriptomic analysis of LINE-1 expression, we describe a brain-wide map of ORF1p expression and interacting proteins at steady-state and in the context of aging. We find a heterogenous but widespread expression of ORF1p in the mouse brain with predominant expression in neurons. In aged mice, neuronal ORF1p expression increases brain-wide and in some brain regions to up to 27%. In human dopaminergic neurons, young LINE-1 transcripts and specific full-length and coding LINE-1 copies are increased in aged individuals. We further describe endogenous mouse ORF1p interacting proteins revealing known interactors and unexpected interacting proteins belonging to GO categories related to RNA metabolism, chromatin remodeling, cytoskeleton and the synapse.

Results

Widespread and heterogenous expression of the LINE-1 encoded ORF1p protein in the wildtype mouse brain

To investigate the expression pattern and intensities of endogenous LINE-1 encoded ORF1p protein throughout the entire mouse brain, we devised a deep-learning assisted cellular detection methodology applied to pyramidal large-scale images using a comprehensive workflow complemented by an approach based on confocal imaging as schematized in Figure 1A. Briefly, starting from sagittal slide scanner images of the mouse brain, we defined anatomical brain regions by mapping the Allen Brain Atlas onto the slide scanner images using Aligning Big Brains & Atlases (ABBA). We then employed a deep-learning detection method to identify all cell nuclei (Hoechst) and categorize all detected cells into neuronal cells (NeuN+) or non-neuronal cells (NeuN-) and ORF1p-expressing cells (ORF1p+) or cells that do not express ORF1p (ORF1p-). This workflow allowed us then to characterize the cell identity of ORF1p+ cells and ORF1p intensity throughout the whole brain but also in specific anatomical regions. In parallel, we completed the approach using confocal microscopy on selected anatomical regions allowing for comparison with higher resolution. Importantly, the specificity of the ORF1p antibody, a widely used, commercially available antibody 18,34–38, was confirmed by blocking the ORF1p antibody with purified mouse ORF1p protein resulting in the complete absence of immunofluorescence staining (Suppl Fig. 1A), by using an in-house antibody against mouse ORF1p17 which colocalized with the anti-ORF1p antibody used (Suppl Fig. 1B, quantified in Suppl Fig. 1C), by immunoprecipitation and mass spectrometry used in this study (Fig 6A, Suppl Table 2) and by siRNA-mediated knock-down of ORF1 in a differentiated mouse dopaminergic cell line (MN9D; Suppl Fig. 1D). Unexpectedly, we found a generalized and widespread expression of ORF1p throughout the brain of wildtype mice (Fig. 1B; Swiss OF1 mice, three months-old; whole brain except regions with particularly high cellular density (cerebellum, hippocampus, olfactory bulb) which impedes nuclei detection by deep-learning. ORF1p is detectable in all regions and subregions analyzed with heterogenous expression patterns (density and intensity) per region/subregions. The ten regions shown in Figure 1B exemplify visible different densities of ORF1p+ cells with varying levels of expression. Notably, the expression pattern of ORF1p in the hippocampus is similar to what has recently been published 18 (Fig. 1B, panel 2). Throughout the entire brain, the mean density of ORF1p+ cells per mm² was ≈ 305 ±18 (mean ±SEM), representing up to 20% of all detected cells (Fig. 1C). ORF1p+ cells in each mouse brain analyzed showed up to eight-fold disparities in intensity between low- and high-expressed cells (Fig. 1C). We then quantified nine anatomical regions according to the Allen Brain Atlas on four brains of three-month old mice (Fig. 1D) using the automated workflow (Fig. 1A) with regard to cell density (Fig. 1D), cell proportions (Fig. 1E) and fluorescent intensity of ORF1p+ cells (Fig. 1F). This approach permitted the analysis of about 10 000 ORF1p+ cells per animal highlighting the power of our large-scale analysis. Densities of ORF1p+ cells ranged from the lowest density in the hindbrain with 154 ±19 cells per mm2 (mean ±SEM) to the highest density of ORF1p+ cells in the isocortex with 451 ±44 cells per mm2 (mean ±SEM) and the thalamus with 446 ±50 cells per mm2 (mean ±SEM). The proportion of ORF1p+ cells per anatomical region fluctuated between 10% ±2.1 (ventral striatum, mean ±SEM) and 31% ±1.6 (thalamus, mean ±SEM). The dorsal striatum (“striatal dorsal” in the Allen Brain Atlas denomination) exhibited the lowest ORF1p expression intensity (658 ±3 mean ±SEM) of all regions tested, the hindbrain the highest mean intensity of ORF1p per cell (mean ±SD 1221 ±548) as illustrated in Figure 1B and quantified in Figure 1D and 1F. Interestingly, cell density did not correlate with expression levels. Dorsal and ventral striatum for instance displayed similar ORF1p intensities per cell but exhibited significant differences in ORF1p cell density and proportion. The “midbrain motor” region as defined by the Allen Brain Atlas showed an intermediate cell density (mean ±SEM 265 ±16 cells per mm2) and a rather high ORF1p expression intensity (mean ±SEM 1006 ±533). Statistical analysis comparing mean density of ORF1p+cells per mm2 or mean intensity per ORF1p+ cells among regions confirmed the heterogeneity concerning ORF1p expression throughout the mouse brain (Fig. 1D, 1F). Slide scanner and confocal images revealed an exceptionally high ORF1p expression intensity in the ventral region of the midbrain which we identified as the Substantia nigra pars compacta (SNpc). This region displayed an important density of ORF1p+ cells and a comparatively high level of ORF1p expression as illustrated by confocal imaging (Fig. 1B, panel 8), but could not be quantified independently with our brain-wide approach due to the geometrically-complex anatomy of this region and its small size (subregion-level in the Allen Brain Atlas hierarchy). Another region which could not be included in our brain-wide analysis was the cerebellum due to its extremely high density of cell nuclei. However, slide scanner and confocal imaging (Fig. 1B, panel 10) revealed that ORF1p is expressed in Purkinje cells, while not detectable in the molecular or granular layers.

Widespread and heterogenous expression of ORF1p protein in the mouse brain

(A) Schematic representation of the unbiased cell detection pipeline on large-scale and confocal images. Immunofluorescent images on sagittal mouse brain slices were acquired on a digital pathology slide scanner or on a confocal microscope (DNA stain: Hoechst, neuronal marker; NeuN, protein of interest: ORF1p). Pyramidal images aquired with the slide scanner were then aligned with the hierarchical anatomical annotation of the Allen Brain Atlas using ABBA. Once the regions defined, a deep-learning based detection of cell nuclei (Hoechst staining, Stardist) and cell cytoplasm (NeuN staining, Cellpose) was performed on each sub-region of the atlas. Objects were filtered according to the background intensity measured in each sub-region for each channel (NeuN and ORF1p). The identity and intensity measures were analyzed at the regional and whole brain level. In parallel, confocal images (multiple z-stacks) of two selected regions (frontal cortex and ventral midbrain) were also acquired and identity and intensity were quantified using Cellpose.

(B) Widespread and heterogenous expression of the LINE-1 encoded protein ORF1p in the mouse brain. Representative image of ORF1p immunostaining (orange) of a sagittal section of the brain of a young (three months-old) mouse acquired on a slide scanner. Scale bar = 1mm. (1-10) Representative images of confocal images of immunostainings showing ORF1p expression (orange) in 10 different regions of the mouse brain. Nuclei are represented in blue (Hoechst), scale bar = 50µm. (1) Isocortex, (2) Hippocampus, (3) Striatum dorsal, (4) Thalamus, (5) Midbrain motor, (6) Pallidum, (7) Hypothalamus, (8) Substantia nigra pars compacta, (9) Hindbrain, (10) Cerebellum. ORF1p expression profile in the mouse brain. The entire mouse brain with the exception of the olfactory bulb and the cerebellum were analyzed according to the pipeline on large-scale images described in (A).

(C) Bar plot showing the total number of ORF1p+ cells per mm² in the mouse brain. Data is represented as mean ±SEM, n=4 mice (top). Bar plot indicating the proportion of ORF1p+ cells compared to all cells detected. Data is represented as mean ±SEM, n=4 mice (labeled M1 to M4), 202001 total cells analyzed (middle). Scatter plot showing the mean intensity of ORF1p per ORF1p+ cell. Data is represented as mean ±SD, n=4 mice, 40999 ORF1p+ cells analyzed (bottom).

(D-F) ORF1p expression profile (density, proportion and expression) in defined anatomical regions of the mouse brain. Nine anatomical regions as defined by the Allen Brain Atlas and mapped onto sagittal brain slices (four three-month-old Swiss/ OF1) with ABBA were analyzed using the pipeline on large-scale images described in (A). (D) ORF1p+ cell density in 9 different regions. Bar plot showing the number of ORF1p+ cells per mm². Data is represented as mean ±SEM; *p<0.05; **p<0.01; adjusted p-value, one-way ANOVA followed by a Benjamin-Hochberg test (E) Proportion of ORF1p positive cells in 9 different regions. Bar plot showing the proportion of ORF1p+ cells among all cells detected per region. Data is represented as mean ±SEM. (F) Mean ORF1p expression per cell in 9 different regions. Dot plot showing the mean intensity of ORF1p signal per ORF1p+ cell in 9 different regions. Data is represented as mean ±SD. The number of analyzed cells per region is indicated in the figure. *p<0.05; **p<0.01; ***p<0.001; ****p<0.0001; adjusted p-value, nested one-way ANOVA followed by Sidak’ multiple comparison test.

(G) ORF1p expression in the mouse frontal cortex and ventral midbrain. Confocal images with multiple z-stacks. Dot plot representing the mean intensity levels of ORF1p per ORF1p+ cells. Four three-month-old Swiss/ OF1 mice (labeled as M1 to M4) are represented each by a different color, the scattered line represents the median. ****p<0.0001, nested one-way ANOVA. Total cells analyzed = 4645.

(H-I) ORF1p expression in the mouse frontal cortex and the ventral midbrain. (H) Western blots showing ORF1p (top) and actin expression (bottom) in four individual mice per region which were quantified in (I) using actin as a reference control. The signal intensity is plotted as the fold change of ORF1p expression in the ventral midbrain to ORF1p expression in the frontal cortex. *p<0.05; two-sided, unpaired student’s-test.

(J) ORF1p expression in three regions of the human brain. Western blot showing human ORF1p expression in the cingulate gyrus (CG), frontal cortex (FC) and cerebellum (CB) of post-mortem tissues from a healthy individual. ORF1p (Top), Actin (bottom). For the full Western blot image please see Suppl Fig. 2A.

In order to confirm ORF1p expression by an independent method, we performed Western blot analysis on six micro-dissected regions from the mouse brain (Swiss/OF1 mouse, three-month old). As shown in Suppl Fig. 1E, ORF1p is expressed in all six regions with varying expression levels confirming the overall presence of ORF1p throughout the brain. We then chose two regions with significantly divergent ORF1p expression intensities as identified and quantified on pyramidal large-scale images: the frontal cortex (low) and the ventral midbrain (intermediate to high). We confirmed a significant higher expression of ORF1p in the ventral midbrain compared to the frontal cortex using an approach based on the unbiased, automated quantification of multiple z-stacks using a confocal microscope (Fig. 1G) and by Western-blotting on micro-dissected regions (Fig. 1H, 1I). In concordance with the findings stemming from the large-scale image quantification pipeline (Fig. 1F), the ventral midbrain showed ≈ 2-times higher expression of ORF1p than the frontal cortex as quantified in Figure 1G (1.8-fold) and Figure 1I (2.3-fold) validating our cellular detection methodology for pyramidal large-scale imaging and underscoring the heterogeneity of ORF1p expression levels in the mouse brain.

To investigate intra-individual expression patterns of ORF1p in the post-mortem human brain, we analyzed three brain regions of a neurologically-healthy individual (Fig. 1J, entire Western blot membrane in Suppl Fig. 2A) by Western blotting using a commercial and well characterized antibody which we further validated by several means. The double band pattern in Western blots has been observed in other studies for human ORF1p outside of the brain 39,40 as well as for mouse ORF1p 41. We also validated the antibody by immunoprecipitation and siRNA knock-down in human dopaminergic neurons in culture (differentiated LUHMES cells, Suppl Fig. 2B and 2C) where we detected however in most cases the upper band only. The nature of the lower band is unknown, but might be due to truncation, specific proteolysis or degradation. ORF1p was expressed at different levels in the human post-mortem cingulate gyrus, the frontal cortex and the cerebellum underscoring a widespread expression of human ORF1p across the human brain. This was in accordance with ORF1p immunostainings of the human post mortem cingulate gyrus (Fig. 2H and Suppl Fig. 2E) and frontal cortex (Suppl Fig. 2E), with an absence of ORF1p staining when using the secondary antibody only (Suppl Fig. 2E).

ORF1p is predominantly expressed in neurons in the mouse brain

(A) ORF1p expression is absent in the white matter (corpus callosum) and predominantly expressed in neurons. Proportion of ORF1p+/NeuN+, ORF1p+/NeuN-, ORF1p-/NeuN+ and ORF1p-/NeuN-cells in the white matter (corpus callosum) and the grey matter (left) and in nine different regions (right) analyzed by the cell detection pipeline on large-scale images presented in Figure1A. Exact values can be found in Suppl_Table1.

(B) Representative confocal microscopy images showing ORF1p (red) and NeuN expression (green) in two different regions of the mouse brain. The bottom images show the merge of the two stainings, an overlap of both markers is represented in orange. z-projections; scalebar = 25µm.

(C) Proportion of neurons expressing ORF1p in the frontal cortex and ventral midbrain quantified on confocal images. ns: non-significant; chi-square test on the cell number of the different cell-types analyzed; n=4 mice, data is represented as mean ± SEM.

(D) Proportion of ORF1p+ cells identified as NeuN+ or NeuN-in two different regions, analyzed by confocal microscopy on multiple z-stacks. ns: non-significant; chi-square test, n=4 mice.

(E) ORF1p does not colocalize with glial or microglial cell markers. Representative confocal microscopy images showing ORF1p staining (red) and three different glial cell (GFAP, Sox9, S100β) or microglial (Iba1) markers (white). Note that Iba1 antibody (rabbit) was used with the ORF1p 09 antibody (guinea pig, in house) z-projections, scalebar = 50µm.

(F-G) Separation of neuronal and non–neuronal - cells by FACS confirms predominant neuronal expression of ORF1p. (F) Neuronal (NeuN+) and non-neuronal (NeuN-) cells isolated by fluorescent activated cell sorting (FACS). Dot plots showing autofluorescence versus an appropriate control antibody (IgG rabbit 647; left) and an antibody against NeuN (AB 657, right). The P4 window represents isolated NeuN+ cells (pink) and the P5 fraction NeuN-cells (orange) containing the same number of cells as sorted in P4 for comparison, others NeuN- are represented in blue. (G) Western blot showing ORF1p expression in NeuN- and NeuN+ FACS-sorted cells stemming from Figure F (A and B representing two different FACS experiments).

(H) Representative confocal microscopy images showing ORF1p (red), NeuN (green) and Hoechst (blue) in the cingulate gyrus of the human brain. z-projection; scalebar = 25µm (left). Examples of individuals neurons expressing ORF1p or not are shown on the right panel. z-projection; scalebar = 5 µm (right).

(I) Proportion of ORF1p+ cells identified as NeuN+ or NeuN- in the human cingulate gyrus, analyzed by confocal microscopy on multiple z-stacks.

(J) Proportion of neurons expressing ORF1p in the human cingulate gyrus, analyzed by confocal microscopy on multiple z-stacks.

In summary, our findings reveal the consistent presence of ORF1p expression throughout the mouse brain in all anatomical regions analyzed with high regional variability in terms of density of ORF1p+ cells and ORF1p+ cell intensity. ORF1p is also expressed in the human brain in at least three brain regions. This finding raises several questions concerning cell-type identity of ORF1p expressing cells and potential functions or consequences of ORF1p expression in the mouse and human brain at steady-state.

ORF1p is predominantly expressed in neurons

Following our observation of a wide-spread expression of endogenous ORF1p throughout the brain, we first addressed the question of the cellular identity of ORF1p+ cells. To this end, we used the neuron-specific marker NeuN, commonly used to identify post-mitotic neurons in the central nervous system 42. This allowed us to determine the proportion of neuronal (NeuN+) or non-neuronal cells (NeuN-) expressing ORF1p (ORF1p+) or not (ORF1p-). Making use of our large-scale imaging approach (Fig. 1A), we observed drastic dissimilarities in detected cellular proportions between the white and grey brain matter. As expected, we observed only 1% of NeuN+ cells in the white matter (corpus callosum; Fig. 2A) validating both, the neuronal marker NeuN as such and the ABBA superposition of the Allen Brain Atlas onto the sagittal brain slices. In the grey matter, our approach detected 30.5% NeuN+ cells (dark red and yellow bars in Fig. 2A) which, according to the literature, should include all post-mitotic neurons with only minor exceptions 42–45 and corresponds to the reported proportion of neurons present in the mouse brain 46. The nine identified grey matter regions in Fig. 2A display the proportions of the different cell types per region. The proportion of all cells in a given region which are positive for ORF1p (dark red bars) differed between regions (lowest proportion: hindbrain: 7%; highest proportion dorsal striatum: 26.6%). In the isocortex and the midbrain motor-related regions, the majority of neurons detected express ORF1p (54% and 59% by large-scale analysis, Suppl Fig. 3A; 68.7% and 68.8% by confocal imaging, Fig. 2B, quantified in C), while in the midbrain sensory related regions the proportion dropped to 25% whereas it reached 82% in the thalamus (Suppl Fig. 3A). Altogether, nearly half of all NeuN+ cells throughout the mouse brain expressed ORF1p (mean of all regions: 48.2%; Suppl Fig. 3A). Regarding the cell identity of ORF1p+ cells brain-wide, more than 70% were identified as neuronal by the large-scale approach (Suppl Fig. 3B). This contrasted somewhat with results obtained by the second approach using confocal imaging on multiple z-stacks which indicated that 91.3% (frontal cortex) and 88.5% (ventral midbrain) of ORF1p+ cells were neuronal (Fig. 2D). This difference in percentages of ORF1p+ expressing neurons among all neurons between the large-scale image cell detection methodology and the confocal workflow is most probably due to technical limitations inherent to the large-scale pipeline. Indeed, with the latter approach, region-dependent differences in cell density and signal intensity might be the cause for an underestimation of the proportion of ORF1p+ cells being neuronal due to difficulties in cell detection by StarDist/Cellpose (high cell density) on a single focal plan, technical difficulties which are widely reduced by the multiple z-stack based approach when using a confocal microscope. Moreover, the large-scale pipeline involves background measurements in each sub-region in order to apply stringent filtering. This, however, results in a loss of true positives cells, but avoids cells which are out-of-focus, the presence of which is inherent to the slide scanner microscope which lacks optical sectioning (see Materials & methods section for evaluated model performance). Notably, frontal cortex and ventral midbrain present similar proportion of neurons expressing ORF1p (Fig. 2C), although the percentage of NeuN+ cells between these two regions is significantly different (Suppl Fig. 3C). As we could not rule out that ORF1p might also be expressed in non-neuronal cells, we turned to non-neuronal markers specific for different glial cell populations using two different astrocytic markers (GFAP, Sox9), the astro- and oligodendrocytic marker S100β and the microglial marker Iba1 46,47 and performed co-staining with ORF1p followed by confocal imaging as illustrated in Figure 2E. We screened multiple images of frontal cortex, ventral midbrain, hippocampus and striatum and did not find a single ORF1p+ cell which could unambiguously be defined as non-neuronal. This indicated that ORF1p is not or only very rarely expressed in non-neuronal cells. To further confirm the predominant presence of expression of ORF1p in neurons and the absence of ORF1p expression in non-neuronal cells, we used fluorescence-activated cell sorting (FACS) to isolate neurons (using a NeuN antibody) and non-neuronal cells (NeuN-) from the adult mouse brain followed by Western blotting with an antibody against ORF1p (Fig. 2F, 2G). As described above, this antibody is well characterized, extensively used and was validated further in this study. After FACS-sorting of neurons from the adult mouse brain using an antibody against NeuN (Fig. 2F, G), we detected ORF1p exclusively in the neuronal population (NeuN+, internal control = heavy chain), confirming the results based on two different imaging approaches. Finally, to assess whether predominant, if not exclusive ORF1p expression in neurons is mouse brain specific or a pattern also applicable to the human brain, we investigated the identity of ORF1p expressing cells in the post-mortem cingulate gyrus of a healthy human brain. Similar to what we found in the mouse brain, we observed sparse NeuN expression in the white and extensive NeuN staining in the grey matter corresponding to the cortical layers (Suppl Fig. 2D, grey and white matter separated by a dashed line) with ORF1p+ cells predominantly located in the grey matter (confocal images in Fig. 2H, Suppl Fig 2E are located in the grey matter). All cells stained by ORF1p were co-stained with NeuN indicating that ORF1p was expressed in neuronal cells in the human brain (Fig. 2H). However, due to the lower signal quality inherent to human post-mortem sections compared to mouse sections, the identity of ORF1p+ cells was estimated to be 80% neuronal by the automated image analysis pipeline of confocal images (Fig. 2I), although no ORF1p+ / NeuN-cells could be clearly identified. Of all neurons identified, 37.2% were ORF1p+ (Fig. 2J), indicating that, similar to the mouse brain, only a fraction of neurons express ORF1p (Fig. 2H, right).

Next, we asked the question of a neuron-subtype specific expression of ORF1p. Our previous study had revealed a higher expression of ORF1p in tyrosine-hydroxylase (TH) positive neurons compared to TH-negative neurons in the mouse ventral midbrain 17. A recent study reported that endogenous LINE-1 RNA and ORF1p expression were higher in parvalbumin (PV)-positive interneurons compared to PV-negative neurons in the mouse hippocampus 18. To address the question of a generalized co-expression of ORF1p and PV cells, we co-stained sagittal brain sections of young mice with antibodies against ORF1p and PV and, in some cases the lectin Wisteria floribunda agglutinin (WFA), which specifically stains glycoproteins surrounding PV+ neurons. Confocal imaging on several brain regions including the hippocampus, cortex, cerebellum, hindbrain, ventral midbrain and thalamus revealed ORF1p+ neurons co-expressing PV, but also many examples of equally intense ORF1p+ neurons that do not express PV (Suppl Fig. 4). In summary, ORF1p expression in the mouse and human brain is widely restricted to neurons of which a proportion express ORF1p. This raises the question of the function and consequences of ORF1p expression specifically in neurons but also on the dynamic regulation of this expression upon exogenous (exposome) or endogenous (aging) challenges.

ORF1p expression is increased in the aged mouse brain

ORF1p is expressed at steady-state throughout the brain, but whether this expression is dynamically regulated remains unknown. Aging has been linked to LINE-1 regulation in some studies 16,48 potentially as both, a trigger and as a consequence of LINE-1 activation, but whether this is true for the brain and if yes, whether this might be region-specific has not been investigated brain-wide. We therefore addressed the question whether advanced age was paralleled by a change of expression patterns or expression levels of ORF1p in the brain. We first analyzed ORF1p expression levels comparing young (3-month) to aged (16-month) mouse brains using the cell detection workflow applied to large-scale images described in Figure 1A. Interestingly, the mean intensity of ORF1p expression increased moderately but significantly with advanced age throughout the brain (13% increase brain-wide; n=4 young mice; n=4 aged mice; p=0.03; Fig. 3A). This was in contrast to another protein, NeuN, which we used as a control and whose intensity did not change between young and aged brains (n=4 young mice; n=4 aged mice; p=0.27; Fig. 3B). Frequency distribution analysis unveiled a shift in ORF1p mean expression per cell in aged mice (Fig. 3C). Importantly, the Hoechst mean intensity within nuclei of ORF1p+ cells, serving as an internal control, showed no significant change (Fig. 3D). Among nine analyzed regions, five demonstrated a general increase in ORF1p mean intensity per cell in aged mice (p≤0.05), a change independent from inter-individual variations in both young and aged mice (Fig. 3E). The increase of ORF1p expression (fold change intensity) throughout the brain, reaching nearly a 30% increase in some regions, is represented on the heatmap in Figure 1F. These results were confirmed by the confocal imaging approach; ORF1p expression in the frontal cortex remained unchanged but increased significantly in the ventral midbrain region in aged mice as shown in Figure 3G and quantified in Figure 3H.

ORF1p expression is increased in some regions of the aged mouse brain

(A) ORF1p mean intensity per ORF1p+ cell in the brain, analyzed on large-scale images. Dot plot showing the ORF1p mean expression per ORF1p+ cell in young (Y1-4) and aged (O1-4) mice in the whole brain (except cerebellum and olfactory bulb). 74985 total cells were analyzed; * p<0.05, nested two-way ANOVA; n=4 mice per condition, data is represented as mean ± SEM.

(B) NeuN mean intensity per NeuN+ cell in the brain, analyzed on large-scale images. Dot plot showing the NeuN mean expression per NeuN+ cell in young (Y1-4) and aged (O1-4) mice in the whole brain (except cerebellum and olfactory bulb). Nested two-way ANOVA; n=4 mice per condition, data is represented as mean ± SEM.

C) Frequency distribution of ORF1p mean intensity in ORF1p+ cells of young (blue) and aged (orange) mice. ***p<0.001, Kolmogorov-Smirnov test.

(D) Frequency distribution of Hoechst mean intensity in the nuclei of OrF1p+ cells of young (blue) and aged (orange) mice. ns: non-significant, Kolmogorov-Smirnov test.

(E) Mean ORF1p expression per ORF1p+ cell in nine different anatomical regions. Dot plot showing the ORF1p mean expression per ORF1p positive cell (n=74985). Adjusted p-value are represented, two-tailed nested t-test followed by a Benjamin, Krieger and Yukutieli test; n=4 mice per region, data is represented as mean ± SEM.

(F) Color-coded representation of fold-changes of ORF1p expression comparing young and aged brains. Represented is the fold-change in percent (aged vs young) of the “mean of the mean” ORF1p expression per ORF1p+ cell quantified and mapped onto the nine different regions analyzed as shown in (J).

(G) Representative confocal microscopy images showing increased ORF1p expression (red) in the ventral midbrain region of aged mice (one z plan is shown). Cell nuclei are shown in blue (Hoechst staining). Scalebar = 50µm.

(H) ORF1p expression is increased in the ventral midbrain of aged mice. Dot plot representing ORF1p expression in two different regions of young and aged mice analyzed on confocal images with multiple z-stacks; total cells analyzed = 8381 ns: non-significant *p<0.05, two-tailed one-way ANOVA; dashed lines represent the medians.

We then asked whether the increase in ORF1p expression levels observed in several brain areas in aged compared to young mice was also accompanied by a change in expression patterns. We therefore analyzed cell proportions and densities comparing young and aged mouse brains. Globally, we observed a reduction in the proportion of ORF1p+/NeuN+ cells in aged mouse brains using the cell detection workflow applied to large-scale images described in Figure 1A, a phenomenon mainly driven by the midbrain motor, the dorsal striatum, the pallidum and the thalamus regions (Suppl Fig. 5A, dark red bars, Suppl_Table1). The confocal approach applied to two regions, the frontal cortex and the ventral midbrain (Suppl Fig. 5B), confirmed this reduction in ORF1p+/NeuN+ cell proportions in favor of the ORF1-/NeuN-cell population in the ventral midbrain with no change in cell proportions in the frontal cortex in accordance with the large-scale imaging approach (Suppl Fig. 5A). The predominantly neuronal identity of ORF1p+ cells, however, was unchanged in the ventral midbrain (Suppl Fig. 5C) just as the proportion of neurons expressing ORF1p (Suppl Fig. 5D). We observed a significant shift in NeuN+/- cell proportions (Suppl Fig. 5E) which could either suggest a decrease in NeuN+ cells or a gain of NeuN-cells in this region with age. While proportions are less sensitive to technical variability and can identify cell population shifts, cell densities allow for absolute comparisons. When quantifying global cellular densities throughout the brain, we did not observe a significant reduction of ORF1p expressing cells (Suppl Fig. 5F), neurons (Suppl Fig. 5G) or non-neuronal cells (Suppl Fig. 5H). When analyzing cell densities in the nine brain regions separately, there were no significant changes in ORF1p positive (Suppl Fig. 5I), NeuN positive (Suppl Fig. 5J) or NeuN negative (Suppl Fig. 5K) cell densities in eight out of nine brain regions. The only exception was the dorsal striatum, but technical limitations applying to this particular brain region might account for these changes. Indeed, the dorsal striatum is different from the other brain regions as it represents the only region consisting of a single ABBA sub-region resulting in only one overall background measurement. Taken together, while there were no major changes in cell proportions, densities nor in ORF1p+ cell identities, we observed an age-dependent increase in ORF1p expression per cell of up to 27%.

Coding LINE-1 transcripts are increased in aged human dopaminergic neurons

Following the observation of increased ORF1p expression in the aged mouse brain, among which the ventral midbrain, and given the age-related susceptibility of dopaminergic neurons in the SNpc to cell death and to degeneration in PD 49, we turned to a previously published RNA-seq dataset of laser-captured micro-dissected post-mortem human dopaminergic neurons of brain-healthy individuals 50, in order to interrogate full-length LINE-1 mRNA expression profiles as a function of age. To avoid read-length bias to which TE analysis is particularly sensitive, we analyzed only the data derived from 50bp paired-end reads of linearly amplified total RNA as this dataset represented all age categories (n=41; with ages ranging from 38 to 97; mean age: 79.88 (SD ±12.07); n=6 ≤65y; n=35 >65y; mean PMI: 7.07 (SD ±7.84), mean RIN: 7.09 (±0.94), metadata available in Suppl_Table5). As age-related dysregulation of TEs might not be linear, we considered individuals with ages-at-death younger or equal to 65 years as “young” (n=6, 38-65 years, mean age 57.5 years (SD ±9.9)) and individuals older than 65 years as “aged” (n=35, 65-97 years, mean age 83 years (SD±7.8). The expression of the dopaminergic markers tyrosine hydroxylase (TH) and LMX1B were similar in both populations indicating no apparent change of dopaminergic identity of analyzed melanin-positive dopaminergic neurons (Suppl Fig. 6A, B). Next, we compared the expression of repeat elements at the class, family and name level based on the repeat masker annotation implemented in the UCSC genome browser using a commonly used mapping strategy for repeats consisting of randomly assigning multi-mapping reads 51. No overt dysregulation of repeat elements at either level of repeat element hierarchy was observed (Suppl Fig. 6C-F). There was a modest but significant increase in several younger LINE-1 elements including L1HS and L1PA2 at the “name” level (Fig. 4A, B), an analysis which was however underpowered (post-hoc power calculation; L1HS: 28.4%; L1PA2: 32.8%) and thus awaits further confirmation in independent studies. No expression changes were observed for HERVK-int, a human endogenous retrovirus family with some copies having retained coding potential (Fig. 4B) or other potentially active TEs like HERVH-int, HERV-Fc1, SVA-F or AluYa5 transcripts in the >65y group (Suppl Fig. 6G). Interestingly, L1HS expression was highly correlated with L1PA2 expression and this correlation extended to almost all younger LINE-1 subfamilies weaning down with evolutionary distance (Fig. 4C). This was not true for other active TEs as L1HS was negatively correlated with HERVK-int expression (Fig. 4C). Several regulators of LINE-1 activity have been identified 17,52 and correlation of their expression with L1HS might allow to infer their relevance of interaction (activation or repression) with L1HS in human dopaminergic neurons. Spearman correlation analysis revealed three known repressors of LINE-1 activity whose expression was negatively correlated with LINE-1 expression; EN1 (Engrailed 1 17, Suppl Fig. 7A) with important functions for dopaminergic neuron homeostasis 53, CBX5/HP1a, a heterochromatin binding protein binding to the histone mark H3K9me3, thereby mediating epigenetic repression 54 (Suppl Fig. 7B) and XRCC5/6, also known as Ku86/Ku70, which are essential for DNA double-stranded break repair through the nonhomologous end joining (NHEJ) pathway and limit LINE-1 full-length insertions 55 (Suppl Fig. 7C). The transcripts of these genes showed, although not statistically significant, a trend for decreased expression in the elderly (Suppl Fig. 7D-G). Based on the increase of young LINE-1 families L1HS and L1PA2 in aged human dopaminergic neurons and the finding that ORF1p was increased in the aged mouse brain, we focused our attention on LINE-1 elements with coding potential for ORF1 and ORF2 according to the L1Basev2 annotation which are specific elements comprised in the L1HS and L1PA2 annotation at the “name” level. Most of the 146 full-length and coding LINE-1 termed UIDs (= Unique Identifier) in the L1Base are L1HS elements (76.03%), whereas the remaining 35 UIDs belong to the evolutionary older L1PA2 family (Suppl Fig. 7A). The L1Base annotation is based on the human reference genome (GRCh38) and annotates 146 human full-length (>6kB), intact LINE-1 elements (ORF1 and ORF2 intact) with a unique identifier from 1 to 146 4. Attribution of sequencing reads to a specific, individual TE copy is problematic 56 and several approaches have been proposed to circumvent this problem including the mapping of unique reads 51.While several tools using expectation maximization algorithms in assigning multi-mapping reads have been developed and successfully tested in simulations 51,57, we used a different approach in mapping unique reads to the L1Base annotation of full-length LINE-1. Specific “hot” LINE-1 loci in a given cellular context have been identified 3, but usage of the L1Base annotation enabled an unbiased approach albeit ignoring polymorphic LINE-1 sequences. Unique read mapping strategies for repeat elements, especially young LINE-1 elements, will unavoidably underestimate LINE-1 locus-specific expression levels 51, but will be most accurate in assigning reads to a specific genomic location while allowing the comparison of two different conditions analyzed in parallel. Of the 146 full-length LINE-1 elements in the L1Base annotation, 111 were of the L1HS family and 35 belonged to the L1PA2 family (Suppl Fig. 8A). Assuming that expression of UIDs was correlated with mappability, we plotted a mappability count of each UID against its mean normalized read count expression of the six individuals ≤65y (Suppl Fig. 8B, C). Non-parametric Spearman correlation revealed no correlation between UID mappability and expression (Suppl Fig. 8C) indicating no apparent bias between the two parameters. However, individual UID dependency of mappability on expression cannot be excluded, especially for high expressing UIDs like UID-16 for example (Suppl Fig. 8B, C). Expression of LINE-1 at the locus-level has been attributed to artefacts not representing autonomous transcription including differential high intronic read counts 58, pervasive transcription or reads attributable to passive co-transcription with genes when the LINE-1 element is intronic 59. To evaluate the latter, we determined the number of intronic (46.58%) and intergenic UIDs (78/146; Suppl Fig. 8D) and identified the corresponding genes for intronic UIDs (Suppl Fig. 8E). Of the 146 UIDs, 140 passed the threshold of >3 reads in at least 6 individuals. Differential expression of UID between “young” and “aged” dopaminergic neurons revealed several significantly deregulated full-length LINE-1 loci (Fig. 5A). Paired analysis of the expression of all UIDs indicated a general increase (Fig. 5B), especially of low expressed UIDs. The comparative analysis of the sum expression of UIDs per individual comparing young (≤65y) with elderly human dopaminergic neurons, however, did not reach statistical significance (Fig. 5C). Several specific loci were dysregulated, in particular UID-68 (Fig. 5A and 5D), a L1HS element located on chromosome 7 (chr7: 141920659-141926712) in between two genes, OR9A4 (olfactory receptor family 9 subfamily A member 4) and CLEC5A (C-type lectin domain containing 5A; Fig. 5E and Suppl Fig.7E). This specific full-length LINE-1 element had a high mappability count of 16 (range of all UIDs: 1-30, mean 9.0 (SD ±6.05), Suppl Fig. 7C) and a post-hoc power analysis score of 96.6% (continuous endpoint, two independent samples, alpha 0.05). To rule out any influence of “hosting” gene transcription interference on measurable UID-68 expression differences, we performed Spearman correlation which did not indicate any correlation between OR9A4 (Fig. 5F) or CLEC5A (Fig. 5G) expression with UID-68. Together, this indicated that UID-68 might be a candidate for an age-dependent gain of activity. Other dysregulated UIDs (i.e. UID-129, UID-37, UID-127 and UID-137) had either a low mappability score, a low post-hoc power or did not pass the visualization check in IGV, reinforcing the notion that a combination of quality control criteria is crucial to retain a specific locus with confidence. In conclusion, TE expression analysis of this human dataset covering an age-span of 59 years (mean age difference between both groups 25.5 years) indicates an increase in the expression of young LINE-1 elements including those which have coding potential in elderly dopaminergic neurons, particularly a specific full-length LINE-1 element on chromosome 7 (UID-68). A slight net sum increase of UID transcripts/cell might be sufficient for the production of “above steady-state” levels of ORF1p and ORF2p. Other TEs with coding potential, namely members of the HERV family, were not increased. Further, correlation analyses suggest that L1HS expression might possibly be controlled by the homeoprotein EN1, a protein specifically expressed in dopaminergic neurons in the ventral midbrain 53, the heterochromatin binding protein HP1 and the DNA repair proteins XRCC5/6.

Young LINE-1 elements are increased in aged human dopaminergic neurons

TE transcript expression in RNA-seq data of laser-captured micro-dissected post-mortem human dopaminergic neurons of brain-healthy individuals was analyzed using RepeatMasker (multimappers) or the L1Base (unique reads).

(A) Volcano plot of differential analysis of LINE-1 expression using DESeq2 comparing young (≤65y, n=6) or aged (>65y, n=35) human dopaminergic neurons at the “name” level of RepeatMasker. Young LINE-1 elements, including the two families L1HS and L1PA2 that have coding copies, are highlighted in red.

(B) Scatter plots of normalized read counts (“name” level) of the young L1HS and L1PA2 families as well as the human endogenous virus family HERVK-int, another TE family with coding potential comparing young (≤65y, n=6) or aged (>65y, n=35) human dopaminergic neurons. Mann-Whitney test, p<0.05.

(C) Correlation of the RNA expression levels of LINE-1 elements with known transposable element regulators in human dopaminergic neurons (all ages included). Spearman correlation of evolutionary close (L1HS, L1PA2-17) and distant LINE-1 (L1PB and L1MA5) as well as HERV elements with coding potential (HERV-Kint, HERV-Fc1, HERV-Fc2 and HERV-H-int) with known regulators of transposable elements for each individual sample, all ages included. HERV-W and TREX1 did not pass the normalized read count threshold of >3 reads in >6 individuals.

Dysregulation of locus-specific full-length LINE-1 elements in aged human dopaminergic neurons

(A) Volcano plot of differential expression analysis of TE expression using DEseq2 comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons at the locus-level of specific full-length LINE-1 elements (140 of 146 “UID’s” as annotated in L1Base; threshold >3 reads in at least 6 individuals).

(B) Pairwise comparison of the expression of 140 out of 146 full-length LINE-1 elements comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons. Wilcoxon matched signed rank test, p<0.0001left panel).

(C) The sum of read counts of all UIDs per individual were plotted comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons.

(D) UID-68 is dysregulated in aged human dopaminergic neurons. Normalized read counts of uniquely mapping reads mapping to the full-length LINE-1 element “UID-68” per individual were plotted comparing human post-mortem dopaminergic neurons from young (≤65y, n=6) and aged (>65y, n=35) individuals; two-tailed Mann-Whitney test (* p=0.046).

(E) IGV window of the locus around the full-length LINE-1 UID-68 (chr7:141.918.055-141.932.948). UID-68 is located adjacent to the genes CLEC5A (right) and OR9A4 (left). Coverage and mappability of the locus including UID-68 is shown in tracks. Coverage is represented by average bigwig profiles (same scale, black: </=65y, red >65y); mappability of the genomic locus is depicted by Umap 50 tracks showing peaks overlapping with the peaks of reads (bigwig averages).

(F) Spearman correlation analysis of the expression of UID-68 and OR9A4 in young (≤65y, n=6, black dots; Spearman r= 0.66, p=0.17)) or aged (>65y, n=35, red squares, Spearman r= 0.15, p=0.40) human dopaminergic neurons.

(G) Spearman correlation analysis of the expression of UID-68 and CLEC5A in young (≤65y, n=6, black dots, Spearman r= 0.75, p=0.11) or aged (>65y, n=35, red squares; Spearman r= - 0.03, p=0.87) human dopaminergic neurons.

Endogenous ORF1p interactors in the mouse brain

To go further in our understanding of steady-state neuronal ORF1p expression across the mouse brain, we immunoprecipitated ORF1p and performed quantitative label-free LC-MS/MS to identify potential protein partners of ORF1p in the adult mouse brain. We successfully immunoprecipitated endogenous ORF1p from whole brain lysates (Fig. 6A), where we detect ORF1p exclusively in the 5 independent ORF1p-IP samples (and not at all in 5 independent IgG-IP control samples; Suppl_Table2) and identified a total of 424 potential protein interactors associated with ORF1p (Suppl_Table2; n=5 mice). Using Gene Ontology (GO) analysis, we identified several interacting proteins belonging to GO terms related to known functions of the ORF1p protein in RNA binding, preferentially 60 but not exclusively in cis 61, for instance RNA decapping and mRNA catabolic process, or related to the known presence of ORF1p in ribonucleoprotein particles 62,63 (GO: cytoplasmic ribonucleotide granule) or the presence of ORF1p in p-bodies 61 as shown in Figure 6B and listed in Suppl_Table3. Other GO terms that emerged, to our knowledge not previously associated with ORF1p, were related to cGMP-mediated signaling (GO: cGMP-mediated signaling and 3’-5’phosphodiesterase activity: i.e. PDE4A, PDE4B, PDE4DIP) and the cytoskeleton (GO: microtubule depolymerization, cytoskeleton organization, microtubule and tubulin binding, cytoskeletal motor activity and protein binding). cGMP signaling is regulated by 3’-5’ phosphodiesterases (PDEs) which degrade 3’,5’-cyclic guanosine monophosphate (cGMP) and 3’,5’-cyclic adenosine monophosphate (cAMP), an activity essential for cell physiology for the integration of extra- and intracellular signals including neuronal excitability, synaptic transmission and neuroplasticity 64,65. Further, several ORF1p interacting proteins were constituents of the mating-type switching/sucrose nonfermenting complex (SWI/SNF complex), i.e. ARID1A, ARID1B, SMARCA2, SMARCB1, SMARCC2), an ATP-dependent chromatin remodeler complex disrupting nucleosome/DNA contacts to facilitate DNA/chromatin accessibility by shifting, removing or exchanging nucleosomes along DNA 66,67. Finally, we also observed proteins belonging to the GO term “neuronal cell body”, corroborating with the neuron-specific presence of ORF1p in the brain. A comparative analysis with previous mass spectrometry studies 34,63,68–72 aimed at identifying ORF1p interacting proteins unveiled significantly more common proteins than randomly expected (overrepresentation test; representation factor 2.6, p< 5.4e-08; Fig. 6C), including LARP1, STAU2, ATXN2, RALY, TARBP2 or DDX21 (for a full list see Suppl_Table4). The presence of a significant number of overlapping ORF1p interactors in different non-neuronal human cells (HEK 63,68,69, HeLa 70, human breast and ovarian tumors 72 and hESCs 71) and mouse brain cells (our study), suggest conserved key interactors between both species and between cell types, with a subset of these proteins regulating RNA degradation and translation potentially relevant for the LINE-1 lifecycle itself. However, differences in experimental conditions in between studies could also influence this overlap. ORF1p interactors found in mouse spermatocytes 34 were also present in our analysis including CNOT10, CNOT11, PRKRA and FXR2 among others (Suppl_Table4). To unravel the physical interactions between the identified interactors of endogenous ORF1p within the mouse brain, we used the STRING database (Search Tool for Recurring Instances of Neighboring Genes, https://string-db.org/). This analysis generated a network representation, where physical interactions are represented by edges (Fig. 6D). In analogy with the GO term analysis, ORF1p displayed interactions with various clusters, including well-known RNA decapping complexes directed against LINE-1 RNA, which also encompassed DCP2 and DCP1A which had not previously been identified as interacting with ORF1p 73. Furthermore, ORF1p exhibited interactions with the SWI/SNF complex (highlighted in red) as well as subunits of the RNA polymerase II complex suggesting a direct or indirect association with accessible chromatin, a hitherto unknown interaction of ORF1p with chromatin compartments within the nucleus. Notably, a multitude of novel interactors belonged to the “neuronal cell body” and “neuron projections” clusters, proposing potential neuron-specific partners of ORF1p such as GRM2/5, BAI1, EPHA4, KCNN2, GRIK2 and DMD among others. A last cluster, formed by NCOA5 (Nuclear Receptor Coactivator 5), NXF1 (Nuclear RNA Export Factor 1), RANBP2 and NU133 (both nucleoporins), might imply a role for these interactions in L1-RNA nuclear export and/or a mechanism for the LINE-1 RNP to gain access to the nucleus in post-mitotic neurons. Altogether, the identification of known and novel interactors of ORF1p in the mouse brain suggests roles of ORF1p in the LINE-1 life cycle (RNA binding and metabolism, RNP formation, nuclear access) but also suggests potential novel physiological roles of ORF1p in the brain related to cytoskeleton organization, cGMP signaling, neuron-specific functions (i.e. synaptic signaling, Suppl_Table3) and chromatin organization and/or transcription regulation.

Endogenous ORF1p interactors in the mouse brain

Immunoprecipitation (IP) of endogenous ORF1p from the mouse brain. WB against ORF1p showing ORF1p enrichment after IP but no signal in the IgG control. Five independent samples were then prepared for proteomic analysis by mass spectrometry (LC-MS/MS).

(B) GO slim enrichment analysis of proteins selected as endogenous ORF1p protein partners in the mouse brain after quantitative LC-MS/MS. ORF1p-immunoprecipitated proteins were categorized into GO slim terms. The nine GO slim term with the highest fold-change are plotted. Fold enrichment is depicted on the upper axis and displayed as bars, the FDR value appears on the lower axis and is represented by the black points. BP: Biological Process, CC: Cellular Component, MF: Molecular Function.

(C) Venn diagram showing common interactors (purple) between interactors of endogenous ORF1p in the mouse brain identified in this study (red) and known (published) interactors of ORF1p (blue). Statistical significance of the overlap between the two groups of proteins was tested by an overrepresentation test (http://nemates.org/MA/progs/overlap_stats.html).

(D) ORF1p associates with the SWI/SNF complex (red), RNA pol II complex (orange) and interactors belonging to GO terms related to neuronal cell body & neuron projection (green). Known interactors previously published 34,63,68–72 are indicated with a purple ring. STRING network of physical interactions where nodes represent proteins partners identified in (A) and edges thickness represents the strength of shared physical complexes. Only proteins sharing physical interactions were represented.

Discussion

While LINE-1 derepression in aging has been extensively explored in peripheral tissues and various pathologies, including cancer, our understanding of LINE-1, particularly ORF1p, in the central nervous system remains limited 20,40,74. A recent search of ORF1p peptides in mass spectrometry data spanning 29 different healthy tissues did not reveal the presence of ORF1p in the brain, suggesting that its presence might lie below detection limits 20. Only a few studies explored and detected ORF1p expression in the brain, most focusing on a specific region (in mice 17,33 in rats 75 and in human post-mortem brain 19), but it remained unclear if ORF1p is expressed throughout the entire brain, exhibits cell-type specificity, and most intriguingly, if its expression is influenced by the aging process. Here, using well characterized and validated antibodies, we demonstrate that ORF1p is expressed throughout the entire mouse brain and in at least three regions of the human post-mortem brain at steady-state. Leveraging a comprehensive workflow that incorporates brain atlas registration and machine learning algorithms, we quantified tens of thousands of brain cells, enabling a profound analysis of cell proportions, cell identities, densities and ORF1p expression levels across the entire brain. Surprisingly, more than one-fifth of detected cells expressed ORF1p. Regional variations in ORF1p expression levels were observed, with each region exhibiting distinct proportions, cell density, and signal intensity of ORF1p+ cells. In a non-neurologically diseased human brain, ORF1p was expressed in all three regions examined, that is the cingulate gyrus, the frontal cortex and the cerebellum. This is in accordance to an earlier study using histological staining, which found ORF1p expression in the human frontal cortex, the hippocampus, in basal ganglia, thalamus, midbrain and the spinal cord 19. This suggests, similarly to the mouse brain, a generalized expression across the human brain. On the transcriptomic level using long-read sequencing of GTEx tissues, brain and liver were highlighted as the organs displaying the highest expression of putatively active, full-length LINE-1 elements 76. However, when the authors looked at sub-regions, they found transcript expression in cerebellar hemispheres and the putamen, but not in the caudate and the anterior cingulate gyrus and frontal cortex 76. This is in contrast to our data and the data from Sur et al, where ORF1p was found to be expressed in the latter two regions using two different antibodies. We used the anti-human LINE-1 ORF1p antibody clone 4H1, a well characterized antibody 74,77. While the sample size for the staining of human post-mortem tissues certainly needs to be increased in order to draw quantitative conclusions, the presence of the protein in two independent studies does point to a steady-state expression of ORF1p in the human brain.

It is interesting to note that ORF1p is expressed at steady-state in both, the mouse and human brain, despite the fact that evolutionary young and thus potentially ORF1-encoding LINE-1 elements in mice 78 (L1A, Tf/Gf) and in humans 4 (L1HS, L1PA2) differ significantly in number, sequence and regulation 79. Most differences lie in the 5’promoter region, but also ORF1 and ORF2 sequences are strikingly divergent between mouse and humans. For example, mouse LINE-1 promoters are composed of a varying number of monomers, a structure not found in human-specific LINE-1 promoters 80. This has obvious implications for LINE-1 expression regulation which can be very different, but examples of co-evolution of regulatory networks have been described 81 and might operate in the brain to regulate LINE-1 and thereby ORF1p expression.

In the mouse brain, we find ORF1p to be expressed predominantly if not exclusively in neurons using immunofluorescence and fluorescence-activated cell sorting (FACS) followed by Western blotting. This result is consistent with previous studies, such as the identification of ORF1p in excitatory neurons within the mouse frontal cortex 82, in parvalbumin neurons in the hippocampus 18, its presence in neurons in the ventral midbrain including in dopaminergic neurons 17 and the recognition of morphological similarities between stained neurons and ORF1p+ cells in a post-mortem hippocampus sample of a healthy individual 19. We also detected ORF1p in Purkinje cells in the mouse and human cerebellum. Neuronal specificity or preference of LINE-1 expression was also shown on the transcriptomic level in recent studies investigating LINE-1 expression in the mouse hippocampus, where neuronal LINE-1 expression exceeded that of astrocytes and microglia by approximately twofold 20, is abundant in parvalbumin interneurons 18 and single-nuclei RNA-seq data from the mouse hippocampus and frontal cortex which confirmed globally that repetitive elements including LINE-1 are more active in neurons than in glial cells 82. In the human brain, LINE-1 transcripts were found in greater quantities in neurons compared to non-neuronal cells by single-nucleus sequencing83. Furthermore, retrotransposition-competent LINE-1 elements were found expressed exclusively in neurons 84. While ORF1p expression is suggested to be expressed in microglia under experimental autoimmune encephalomyelitis conditions in the spinal cord 85, no evidence of such expression was observed in non-neuronal cells under non-pathological condition.

On average, throughout the mouse brain, the majority of neurons were positive for ORF1p and in some regions (i.e. the thalamus) around 80% of neurons expressed ORF1p. Comparing the results of both imaging approaches, the percentages of neurons expressing ORF1p in the ventral midbrain and frontal cortex were roughly similar (around 70% of neurons expressed ORF1p as quantified by confocal imaging and about 60% of neurons were identified as ORF1p+ using the slide scanner approach). In the human cingulate gyrus, we found that 37.2% of neurons express ORF1p and that 80% of cells expressing ORF1p were neurons, which are proportions similar to some regions of the mouse brain. It is however possible that these percentages are underestimated due to technical issues inherent to the machine-learning based algorithm for cell detection as our observations often indicated a positive signal in neurons which were classified as negative due to a particular shape or our stringent intensity threshold. A question which arises based on these findings is whether specific features distinguish ORF1p+ and ORF1p-neurons. One hint comes from a recent study suggesting that in the mouse hippocampus, it is the parvalbumin positive neurons that predominantly express ORF1p 18. However, as we show here, while PV-positive neurons often co-stain with ORF1p, not all ORF1p positive cells are PV-positive. In the mouse ventral midbrain, TH-positive dopaminergic neurons express higher levels of ORF1p compared to surrounding, non-dopaminergic neurons 17 (and this study Fig. 1B, panel 8). In the mouse cerebellum, ORF1p staining was detected in Purkinje cells and in the human post-mortem brain in Purkinje and possibly in Basket cells. Parvalbumin positive neurons are inhibitory neurons, so are Purkinje and Basket cells. However, dopaminergic neurons are modulatory neurons exerting excitatory and inhibitory effects depending on the brain region they act on. Specific neurons in the granular layer (i.e. Golgi and unipolar brush cells) of the cerebellum are inhibitory, but ORF1p negative, indicating that the decisive feature might not be the excitatory or inhibitory nature of a neuron. Further, ORF1p is expressed in excitatory (CamKIIa-positive) and CamKIIa-negative neurons in the mouse frontal cortex 82 and there is evidence of full-length L1 RNA expression in both excitatory and inhibitory neurons 83. While further studies are necessary to define the neuronal subtypes expressing ORF1p and their epigenetic make-up allowing this expression, it seems reasonable to conclude on the above-mentioned data that there is no neuronal sub-type specificity characterizing ORF1p expressing neurons. Another possibility is a cell-type specific chromatin organization permissive for the expression of LINE-1 and future single-cell studies in the mouse and human brain might reveal those differences.

Because transposable elements are known to become active in somatic tissues during aging 15,16,28,86,87, we aimed to investigate whether there was a corresponding increase at the protein level. In aged mice, ORF1p expression significantly increased throughout the mouse brain consistent with a previously documented increase in ORF1p outside the central nervous system in aged rats 75,88 and aged mice 86 and in neurons of layer 2/3 of the mouse frontal cortex82. By quantifying the mean intensity of ORF1p in over 70 000 cells identified as ORF1p+, we were able to characterize the extent of this increase in each anatomical sub-region. Remarkably, apart from the isocortex which did not show any change, ORF1p expression increased in all other brain regions by 7% to 27%, indicating a generalized increase of ORF1p expression in neurons throughout the brain (13%). We did not detect any change in cell identity of ORF1p expressing cells, that is, ORF1p expression remained predominantly if not exclusively neuronal. Globally, there was also no change in ORF1p-positive, neuronal or non-neuronal cell densities in the aged mouse brain. Further investigations are necessary to investigate the underlying mechanism of the loss of ORF1p+ cells in the dorsal striatum in the aged mouse brain and to examine a possible relationship to the change of proportions of cells in the ventral midbrain, a structure which contains the SNpc which projects to the dorsal striatum and which is prone to LINE-1 driven neuronal degeneration 17. Thus, while ORF1p intensities per cell increase significantly in older mice in several brain regions, here is no global change in ORF1p+ cell numbers.

An increase of ORF1p might have several direct or indirect consequences on a cell or here, on a neuron. As ORF1p is translated from a polycistronic LINE-1 RNA together with ORF2p, albeit in much higher amounts (the estimated ratio ORF1p to ORF2p is 240:1 in non-native conditions) 89, it can be expected that a LINE-1 ribonucleotide particles are formed and ORF2-dependent cell toxicity in form of genomic instability 17,21 and single-stranded cytoplasmic DNA triggered inflammation 27,28,86 might result. This has been shown in mouse dopaminergic neurons where oxidative stress induced LINE-1 causally contributed to neurodegeneration 17.

Neurodegeneration was partially prevented by anti-LINE-1 strategies among which NRTIs 17 and similar LINE-1 protein-dependent neuronal toxicity has been shown in drosophila31,32 and the mouse cerebellum 33.

In order to test whether an increase in LINE-1 is a feature of human brain aging, we turned to a unique RNA-seq dataset of human laser-captured dopaminergic neurons of 41 individuals ranging from 38 to 99 years 50. In accordance with our focus on LINE-1 sequences which are full-length and coding, we developed a rationale to interrogate LINE-1 families with representatives that are coding (L1HS, L1PA2, multimappers; RepeatMasker) and to specifically investigate full-length LINE-1 elements that have intact open reading frames for ORF1p and ORF2p (unique reads; L1Basev2 4). Indeed, we find an increase in L1HS and L1PA2 elements in individuals ≥65y as well as an increase in specific full-length LINE-1 elements but only a trend for increase of all full-length LINE-1 in sum in the elderly. This analysis has technical limitations inherent to transcriptomic analysis of repeat elements especially as it is based on short-read sequences and on a limited and disequilibrated number of individuals in both groups. Nevertheless, we tried to rule out several biases by demonstrating that mappability did not correlate with expression overall and used a combination of visualization, post-hoc power analysis and analysis of the mappability profile of each differentially expressed full-length LINE-1 locus. Interestingly, dysregulated full-length LINE-1 elements in aged dopaminergic neurons did not correspond to those identified in bladder cancer 90 indicating the intricate nature of this expression across tissues and pathological conditions. Overall, a slight net sum increase of UID transcripts/cell might be sufficient for the production of “above steady-state” levels ORF1p and ORF2p. Further, a dissociation of LINE-1 transcript and protein levels in aging has been observed recently in excitatory neurons of the mouse cortex. In the absence of transcriptional changes of LINE-1, protein levels of ORF1p were increased 82.

We can only speculate about the reason for an increase in ORF1p in the aged brain. A recent single-cell epigenome analysis of the mouse brain suggested a specific decay of heterochromatin in excitatory neurons of the mouse brain with age which was paralleled by an increase in ORF1p, albeit equally in excitatory and inhibitory neurons, again not indicating any dependency of ORF1p regulation on the excitatory or inhibitory nature of neurons 82. Chromatin and particularly heterochromatin disorganization are a primary hallmark of aging 87 but other repressive cellular pathways which control the LINE-1 life cycle might also fail with aging 13. Another possibility is a loss of accessibility of repressive factors to the LINE-1 promoter or an age-dependent decrease in their expression. Matrix correlation analysis of several known LINE-1 regulators, both positive and negative, revealed possible regulators of young LINE-1 sequences in human dopaminergic neurons. Despite known and most probable cell-type unspecific regulatory factors like the heterochromatin binding protein CBX5/HP1 54 or the DNA repair proteins XRCC5 and XRCC6 52, we identified the homeoprotein EN1 as negatively correlated with young LINE-1 elements including L1HS and L1PA2. EN1 is an essential protein for mouse dopaminergic neuronal survival 53 and binds, in its properties as a transcription factor, to the promoter of LINE-1 in mouse dopaminergic neurons 17. As EN1 is specifically expressed in dopaminergic neurons in the ventral midbrain, our findings suggest that EN1 controls LINE-1 expression in human dopaminergic neurons as well and serves as an example for a neuronal sub-type specific regulation of LINE-1. Although these proteins are known regulators of LINE-1, this correlative relationship awaits experimental validation.

The heterogenous, brain-wide presence of ORF1p expression at steady-state is intriguing. In cancer cell lines or mouse spermatocytes, ORF1p interacts with several “host” proteins, some if not most of which are related to the LINE-1 life cycle. However, a profile of endogenous ORF1p interactors in the mouse brain might inform on possible other and organ-specific functions besides its binding to the LINE-1 RNA in “cis”34. Among the total 424 potential interactors of endogenous ORF1p in the mouse brain, 38 partners had been previously identified by mass spectrometry in human cancers, cancerous cell lines and mouse spermatocytes 34,63,68–72 (Suppl_Table4). This supports the validity of the list of ORF1p partners identified, although we cannot rule out the possibility that unspecific protein partners might be pulled down due to colocalization in the same subcellular compartment. GO term analysis contained expected categories like “P-body”, mRNA metabolism related categories and “ribonucleoprotein granule”. We also identified NXF1 as a protein partner of ORF1p, a protein found to interact with LINE-1 RNA related to its nuclear export 91. This suggests the conservation of key interactors probably essential for completing or repressing the LINE-1 life cycle in both species, despite the divergence of mouse and human ORF1p protein sequences 92. Along these lines, several ORF1p protein partners we identified might complete the list of post-transcriptional regulators implicated in LINE-1 silencing. Recent work conducted on human cancerous cell lines has demonstrated that MOV10 orchestrates the recruitment of DCP2 for LINE-1 RNA decapping 73. In our analysis, we identified DCP2 along with DCP1A, known to enhance the decapping activity of DCP2 93, and DCP1b, a pivotal component of the mRNA decapping complex 94. Intriguingly, MOV10 was not detected in our mass spectrometry analysis, despite its established role in recruiting DCP2 and forming a complex with L1-RNP to mediate LINE-1 RNA decapping, as reported by Liu et al 73. However, we found two enhancers of mRNA decapping, EDC3 and EDC4, both core components of P-bodies, a membrane-less organelle known to contain L1-RNP 61. Multiple ubiquitin-ligase proteins were found although not appearing as a significantly enriched GO term. These results add to the picture of the post-transcriptional and translational control of ORF1p and suggest that these mechanisms, despite a steady-state expression, are operational in neurons. Further, several neuron-specific interactors were identified belonging to GO term categories “neuron projection” (75 proteins) and “neuronal cell body” (5 proteins), again pointing to the neuron-predominant expression of ORF1p in the mouse brain. Other interesting aspects were raised from this analysis. Among significantly enriched GO terms, several were related to the cytoskeleton, the functional consequences of which need to be determined in future studies. Our screen also identified PDE10A as an interactor of ORF1p in the mouse brain, a PDE almost exclusively expressed in medium spiny neurons of the striatum and a target for treatment of neurological diseases related to basal ganglia function like Huntington’s disease, schizophrenia and Tourette syndrome 95. Interestingly, PDE10A inhibition is related to beta-catenin signaling, another GO term which emerged from our screen 96. Finally, we found components of RNA polymerase II and the SWI/SNF complex as partners of ORF1p. This might further indicate that ORF1p has access to the nucleus in mouse brain neurons as described for other cells 97,98, however a bias due to a post-lysis effect cannot be excluded. These findings give rise to intriguing questions regarding the potential function of ORF1p in neuron in health and disease as (i) ORF1p is widely distributed throughout the brain under normal physiological conditions, (ii) ORF1p shows a wide range expression levels within and in between regions, (iii) ORF1p is expressed predominantly if not exclusively in neurons, (iv) but not in all neurons and (v) interacts with proteins that might not directly relate to the LINE-1 life cycle, some of which are neuron-specific. In addition, physicochemical properties of ORF1p to form compacted nucleic acid - bound complexes with sequestration potential were shown 92,99. Future loss-of-function studies should help to shed light on the necessity of ORF1p for neuronal functions if they exist. This data spurs the idea of a possible “physiological” function of ORF1p as an integrative protein with exapted function in neuronal homeostasis and a loss of restriction in the aged brain limiting LINE-1 expression to steady-state levels.

Materials & Methods

Animals

Swiss OF1 wild-type mice (Janvier) were housed on a 12h light/dark cycle with free access to water and food. Mice were sacrificed at 3-month or 16-month. Animal experiments were performed according to the EU directive 2010/63/EU.

Mouse tissue dissection and protein extraction

Tissues were extracted from 3-month-old and 16-month-old Swiss/OF1 mice. Briefly, the two hemispheres were separated in ice cold PBS -/-. For each mouse, one hemisphere was rinsed and fixed in 4% PFA for 1h followed by 24h of incubation in 30% sucrose. Hemispheres were kept at -20°C until being sliced on a freezing microtome (Epredia, HM 450) with a 20µm thickness. The other hemisphere was dissected in ice cold PBS -/- 1X and 6 brain regions were rinsed, cut in small pieces and dissociated separately using a large (21G) to small gauge (27G) needle in RIPA lysis buffer for 5 min. Lysates were kept on ice for 25 min, were sonicated for 15 min and supernatants were collected after a 30 min centrifugation at 4°C at 14 000 rpm. Proteins were quantified and Laemmli buffer was added before boiling for 10min at 95°C to be used for Western Blot.

Human Samples

Cerebellum, frontal cortex and cingulate gyrus human samples were provided by Neuro-CEB biobank from a 78-year-old brain-healthy male and conserved at -80°C.

Human Samples pulverization and protein extraction

We used the dry pulverizer Cryoprep (Covaris) for pulverization of tissue blocs. Each sample was disposed in a liquid-nitrogen precooled Tissue-tube bag and dry cryo-pulverized with one impact at the maximum level. The pulverized brain sample was then weighed and resuspended in lysis buffer (mg/v) (0.32M sucrose, 5mM CaCl2, 3mM Mg(CH3COOH)2, 0.1mM EDTA, 10mM Tris-HCL pH8, 1mM DTT, 0.1% TritonX-100 and Protease Inhibitors), kept on ice for 30 min with gentle up-and-down pipetting until homogenization. We added 2X RIPA buffer (v/v) to totals fractions for 30 min on ice. We then sonicated samples 2 times for 15 min. AtlasSupernatants were collected after a 30 min centrifugation at 14 000 rpm at 4°C, proteins were quantified and Laemmli buffer was added to be used for Western Blot. All samples were boiled 10 min at 95°C to be used for Western Blot.

Cell culture and siRNA delivery

LUHMES were purchased from ATCC (CRL2927). Cells were cultured on 50 μg/mL Poly-L-ornithine (Merck) and 1 μg/mL human plasma fibronectin (Sigma) coated flasks and cultured in Advanced DMEM/F12 (Gibco) added with 1% N-2 supplement (Gibco), GlutaMax (Gibco) and 40 ng/mL human recombinant FGF (Peprotech) at in 5 % CO2, 37C° incubator. Differentiation was initiated by adding to the media 1 µg/ml doxycycline (Sigma), 2 ng/ml recombinant human GDNF (Peprotech) and 1 mM cAMP (Sigma). Media was changed every two days. Experiences were performed on day 7 of differentiation. Cells were passaged less than 12 times.

MN9D were purchased from Merck (SCC281). Cells were cultured on 1 mg/mL poly-lysine (Merck) coated flasks and cultured in DMEM High Glucose (Sigma, D5796) with 10% FBS (Gibco) in 5 % CO2, 37C° incubator. Over a period of 10 days, cells were differentiated by adding to the media 1 mM n-butyrate (Sigma) and 1 mM dibutyryl cAMP (Sigma). Media was changed every two days. Experiences were performed on day 10 of differentiation. Cells were passaged less than 15 times.

LUHMES were lipofected using RNAiMAX (Invitrogen) at day 3 of differentiation with 100 nM of siRNA. MN9D were lipofected using RNAiMAX (Invitrogen) at day 7 of differentiation with 100 nM of siRNA. The siRNA sequences used were as follows:

siRNA control: TAATGTATTGGAACGCATA

siRNA ORF1 (LUHMES): AAGAAGGCTTCAGACGATCAA

siRNA ORF1 (MN9D): CTATTACTCTGATACCTAAAC

LUHMES were scrapped at day 7 of differentiation and MN9D at 10 days of differentiation, in RIPA buffer (10mM Tris-HCl, pH 8.0; 150mM NaCl; 1mM EDTA; 1% Triton X-100; 0.1% Sodium Deoxycholate; 0.1% SDS). Laemmli buffer was added and samples were boiled 10 min at 95C° before being loaded on a gel.

Western Blot

We used 1.5mm NuPAGE 4-12% Bis-Tris-Gel (Invitrogen™). Proteins samples (sorted mouse brain cells: 10000 cells/ µl -> 5 µl loaded; human brain lysates: 10µg; mouse brain lysates: 20µg) were loaded and gel migration was performed with NuPAGE™ MES SDS Running Buffer (Invitrogen™) for 45 minutes at 200mV. Gels were transferred onto a methanol activated PVDF membrane (Immobilon) in a buffer containing: Tris 25 mM, pH=8,3 and Glycine 192 mM, during 1h30 at 400 mA. Membranes were blocked 30 min with 5% milk in TBST (0,2% Tween 20, 150 mM NaCl, 10 mM Tris pH:8). The primary antibodies (mouse ORF1p antibody: abcam ab216324; human ORF1p antibody: abcam ab245249) were diluted in 5% milk in TBST, and membranes were incubated o/n at 4 C°. After 3 x 10 min washing in TBST, membranes were incubated for 1h30 with the respective secondary antibodies diluted at a concentration of 1/2000 in 5% milk TBST. Membranes were washed 3 x 10 min in TBST and were revealed by the LAS-4000 Fujifilm system using Clarity Western ECL Substrate (Bio Rad) or Maxi Clarity Western ECL Substrate (Bio Rad).

Immunostaining

Sagittal mouse brains slices were fixed for 10min in PFA 4% and rinsed 3 times for 10min in PBS -/-. Slices were then incubated 20 min in glycine 100mM, washed 3 times for 5min in PBS and immersed in 10mM citrate pH 6 at 62°C during 45min for antigen retrieval. Slices were then immersed 3 times in PBS with Triton X-100 0.2% and incubated in blocking buffer for 1,5 h (PBS with Triton X-100 0.2% and FBS (10%) previously inactivated 20min at 56°C (Gibco, 16141061). Primary antibodies (ORF1p antibody: abcam ab216324; NeuN antibody: GeneTex GTX00837) were diluted (1/200 and 1/500 respectively) in blocking buffer and incubated with slices overnight at 4°C and then washed 3 times for 10 min with PBS. For validation, an in-house ORF1p antibody was used (09) (guinea pig, 1/200). For non-neuronal marker (GFAP antibody: Millipore AB5541; Iba1 antibody: GeneTex GTX101495; Sox9 antibody: RnDsystems AF3075; S100β antibody: Sigma S2532), antibodies were diluted at 1/500. Additionally, WFA (Sigma L1516) and PV antibodies (Swant PV235) were used, diluted at 1/500 and 1/1000 respectively. Suitable secondary antibodies (Invitrogen) and Hoechst (Invitrogen, 15586276) were incubated for 1,5h at 1/2000 in PBS with inactivated FBS (10%) and washed 3 times 10 min in PBS. To quench tissue autofluorescence, especially lipofuscin, TrueBlack Plus (Biotium) in PBS was used during 10min. Slices were rinsed 3 times in PBS and mounted with Fluoromount (Invitrogen™).

For human cingulate gyrus stainings, the same protocol was performed, with the difference that a human ORF1p antibody (Abcam 245249) was used. Mouse and human brain slices were imaged by the Axioscan 7 Digital Slide Scanner (Zeiss) or a Spinning Disk W1 confocal microscope (Yogogawa).

Blocking peptide

The ORF1p antibody (abcam ab216324) was incubated 2h on a turning wheel with excess (4:1) of mouse ORF1p recombinant protein as in 17 before the blocked antibody was used in the above-described immunofluorescence protocol.

Quantification of confocal acquisitions

Analysis was conducted using a custom-written plugin developed for the Fiji software, incorporating Bio-Formats 100 and 3D ImageJ Suite 101 libraries. Code is freely available online at https://github.com/orion-cirb/DAPI_NEUN_ORF1P. Nuclei were detected in the Hoechst channel downscaled by a factor of 2 with the 2D-stitched version of Cellpose 102 (percentile normalization = [1-99], model = ‘cyto’, diameter = 30, flow threshold = 0.4, cell probability threshold = 0.0, stitching threshold = 0.75). The segmented image was then rescaled to its original size, and the obtained 3D nuclei were filtered by volume to reduce false positive detections. NeuN+ and ORF1p+ cells were detected in their respective channel using the same approach as for nuclei detection, but with adjusted Cellpose settings (model = ‘cyto2’, diameter = 40, flow threshold = 0.4, cell probability threshold = 0.0, stitching threshold = 0.75). Finally, each cell was associated with a nucleus having at least half of its volume in contact with. Cells without any associated nucleus were discarded. Each nucleus was thus labeled according to NeuN and/or ORF1p positivity.

Nuclei and cell detection using the respective Cellpose models and hyperparameters were evaluated on eight images per channel, capturing intensity variability across different mouse ages and brain regions. A total of approximately 2,000 nuclei and 1,000 NeuN and ORF1p cells were manually annotated. We evaluated model performance with the average precision (AP) metric, computed from the number of true positives (TP), false positives (FP), and false negatives (FN) as [inline] at the commonly used Intersection over Union (IoU) threshold of 0.5. The AP at an IoU threshold of 0.5 was 0.995 for nuclei, 0.960 for NeuN, and 0.974 for ORF1p cells. These results confirmed that the selected models and hyperparameters were well-suited for processing the entire dataset.

ABBA Registration and Qupath analysis

Each sagittal brain section was registered with the Allen Mouse Brain Atlas (CCFv3 103) using the Aligning Big Brains & Atlases plugin 104 in Fiji. Slices were first manually positioned and oriented along the Z-axis. Automated affine registration was then applied in the XY plane, followed by manual refinement. The final registration results were imported into QuPath software 105 for downstream processing. Cell analysis in each brain subregion was performed with custom Groovy scripts developed for QuPath. Code is freely available online at https://github.com/orion-cirb/QuPath_ORF1P. Hoechst nuclei were detected with StarDist 2D 106, applying the DSB 2018 pretrained model with the following parameters: percentile normalization = [1-99], probability threshold = 0.82, overlap threshold = 0.25. Cells in NeuN and ORF1p channels were detected with Cellpose 2D (percentile normalization = [1-99], model = ‘cyto2’, diameter = 30, flow threshold = 0.4, cell probability threshold = 0.0). Nuclei and cells were then filtered by area and intensity to minimize false positive detections. Minimum intensity threshold was based on the channel background noise, which was estimated for each subregion as the mean intensity of pixels not belonging to any detected nucleus or cell in the respective channel. Finally, each cell was associated with a nucleus having its centroid located inside the cell mask. Cells without an assigned nucleus were discarded, cells with associated nuclei were classified as NeuN+ or NeuN- and ORF1p+ or ORF1p-. Intensity values were normalized by subtracting the background noise computed in the corresponding channel and subregion. As a last step, subregional results were merged into regional ones and data were analyzed using the Pandas Python library 107.

Nuclei and cell detection were evaluated on 14 images per channel, corresponding to approximately 800 nuclei and 400 NeuN and ORF1p cells manually annotated. The average precision (AP) at an IoU threshold of 0.5 was lower than for confocal images: 0.806 for nuclei, 0.675 for NeuN, and 0.695 for ORF1p cells. This decline in performance was primarily due to a lower signal-to-noise ratio in slide scanner images, leading to an increased number of false positives and false negatives. While fine-tuning the models could enhance detection robustness, the selected models and hyperparameters were considered suitable for processing the entire dataset.

FACS

Mouse brains were dissociated with Adult Brain Dissociation kit (Miltenyi Biotec, 130-107-677) and incubated with the coupled antibody NeuN Alexa 647 (Abcam, ab190565) or the control isotype IgG Alexa 647 (Abcam, ab199093). Stained cells were filtered a last time with a 40µm filter before FACS sorting (FACS ARIA II). Neuronal and non-neuronal cells were separately collected in PBS -/- 2m EDTA and then centrifugate (5min at 700rpm). Pellets were resuspended in RIPA for protein extraction in an appropriate volume in order to achieve equal cell concentrations (10 000 cells/µl).

RNA-seq analysis

The RNA-seq dataset from Dong et al 50 was downloaded from dbGAP (phs001556.v1.p1) and contains unstranded paired-end 50bp and 75bp reads from pooled laser-capture micro-dissected dopaminergic neurons from human post-mortem brain (107 samples) from 93 individuals w/o brain disease. RNA-seq had been done on total and linearly amplified RNA. We focused our analysis on data obtained with 50bp reads, in order to avoid mappability bias, while still regrouping all age categories (n=41; with ages ranging from 38 to 97 (mean age: 79.88 (SD ±12.07); n=6 ≤65y; n=35 >65y; mean PMI: 7.07 (SD ±7.84), mean RIN: 7.09 (±0.94)). Sequencing reads were aligned on the Human reference genome (hg38) using the STAR mapper (v2.7.0a) 3 and two different sets of parameters. Genome-wide individual repeat quantification was performed using uniquely mapped reads and the following STAR parameters: --outFilter mapNmax 1–-alignEndsType EndToEnd–-outFilterMismatchNmax 999–- outFilterMismatchNoverLmax 0.06. Repeats class, family and name quantification was performed using a random mapping procedure and the following parameters : -- outFilterMultimapNmax 5000–-outSAMmultNmax 1–-alignEndsType EndToEnd–- outFilterMismatchNmax 999–-outFilterMismatchNoverLmax 0.06–-outSAMprimaryFlag OneBestScore–-outMultimapperOrder Random. Repeats annotations were downloaded from the UCSC Table Browser (repeatMasker database: https://genome.ucsc.edu/cgi-bin/hgTables) and coordinates of LINE-1 full length and coding elements were downloaded from the L1base database 2 (http://l1base.charite.de/l1base.php; 4) selecting LINE-1 full length elements containing two predicted complete open reading frames for ORF1 and ORF2 (UID= Unique IDentifier) from the LINE-1 database (http://l1base.charite.de/l1base.php) and corrected genomic intervals with the repeat masker annotation of the corresponding genomic locus. Repeat quantification from the aligned data was done using a gtf file composed of all genes (Gencode v29) and all individual repeat elements. This strategy was used to avoid overestimation of repeat elements due to overlaps with expressed genes. For individual repeat quantification of the full length L1 elements (L1base), we therefore used a gtf of all genes and all L1base entries, and ran the FeatureCounts tool 108 with the following parameters: -g gene_id -s 0 -p. In the context of the family-based analysis, we used a gtf with all genes and all annotated repeats elements and ran FeatureCounts with -g gene_family -s 0 -p -M. Before DeSeq2 analysis, we remove all genes and repeat elements with less than 10 reads in a minimum of n individuals, n being the number of individuals in the condition containing the fewest individuals (“young” condition: n=6, 38-65y, mean 57.5 years). We use the same conditions with genes and UIDs with less than 3 reads in a minimum of n individuals. Finally, we calculated the scaling factors using DeSeq2 on all genes + all repeat elements or all gene + UID according to the quantification method and then applied these scaling factors to the corresponding counts tables.

In order to test for the mappability of each UID (= full-length and coding LINE-1), we extracted the bed track « main on human:umap50 (genome hg38) from the UCSC genome browser (≈ 7Mio regions) directly into Galaxy (usegalaxy.org) and joined genomic intervals with a minimum overlap of 45bp of this dataset with a dataset containing the annotation of UIDs extracted from L1Basev2 4 corrected in length with repeat masker and completed with information on whether the UID is intra- or intergenic and, if intragenic, in which gene (NM_ID, chr, strand, start, end, gene length, number of exons, gene symbol) the fl-LINE-1 is located, which resulted in 1266 regions. We then used the “group on data and group by” function in Galaxy and counted the number of overlapping 50kmers with all 146 UIDs (=mappability score). Correlation analysis (non-parametric Spearman) was then done between the mappability score and the normalized read counts.

For visualization of expression, bigwig files were generated for each age group, that is </=65y and >65y, respectively. We used bamcoverage to obtain bigwig files (normalized by cpm). Then for each age group an average bigwig was generated using bigwigAverage from deeptools (galaxy version 3.5.4); Bigwigs were loaded into IGV alongside tracks showing mappability (Umap50) and selected tracks from repeat masker (LINE, SINE and LTR).

Post-hoc power analysis was performed using the “Post-hoc Power Calculator” (https://clincalc.com/stats/Power.aspx).

Immpunoprecipitation of ORF1p from the mouse brain

For immunoprecipitation, we used ORF1p (abcam, ab245122) and IgG rabbit (abcam, ab172730) antibodies. The antibodies were coupled to magnetic beads using the Dynabeads® Antibody Coupling Kit (Invitogen, 14311D) according to the manufacture’s recommendations. We used 5µg of antibody for 1 mg of beads and used 1.5mg of beads for IP. Individual mouse brain lysates (n=5), homogenized using dounce and sonicated, were incubated with ORF1p or IgG-control coupled beads and a small fraction was kept as input. Each of these two tubes containing coupled beads and brain lysates were diluted in 5 ml buffer (10 mM Tris HCl, 150 mM NaCl, protease inhibitor). The samples were then incubated overnight on a wheel at 4°C. Samples were then washed 3 times with 1 ml buffer (10 mM Tris HCl pH 8, 200 mM NaCl) using a magnet and then resuspended in the same buffer. The samples were boiled in Laemmli buffer (95°C, 10 min) and 20 µl of each sample were loaded on a 4-12% Nupage gel (Invitrogen, NP0336) to be revealed by WB. For samples used in Mass Spectrometry study, beads were washed with buffer (10 mM Tris HCl pH 8, 200 mM NaCl) using a magnet. After 3 washes with 1ml buffer the beads were washed twice with 100 µL of 25 mM NH4HCO3 (ABC buffer). Finally, beads were resuspended in 100 μl of 25mM ABC buffer and digested by adding 0.20 μg of trypsine/LysC (Promega) for 1 hour at 37 °C. A second round of digestion was applied simultaneously on the beads by adding 100 µL of 25 mM ABC buffer and to the previous digest by adding 0.20 µg of trypsin/LysC for 1 hour at 37 °C. Samples were then loaded into homemade C18 StageTips packed by stacking three AttractSPE® disk (#SPE-Disks-Bio-C18-100.47.20 Affinisep) into a 200 µL micropipette tip for desalting. Peptides were eluted using a ratio of 40:60 CH3CN:H2O + 0.1% formic acid and vacuum concentrated to dryness with a SpeedVac device. Peptides were reconstituted in 10 µL of injection buffer in 0.3% trifluoroacetic acid (TFA) before liquid chromatography-tandem mass spectrometry (LC-MS/MS) analysis.

Immunoprecipitation of ORF1p from LUHMES cells (human) differentiated into mature dopaminergic neurons

For immunoprecipitation, we used ORF1p (Millipore, MABC 1152) and IgG mouse (Thermofisher, #31903) antibodies. The antibodies were coupled to magnetic beads using the Dynabeads® Antibody Coupling Kit (Invitogen, 14311D) according to the manufacturer’s recommendations. We used 8 µg of antibody for 1 mg of beads. The appropriate volume of buffer was added to the coupled beads to achieve a final concentration of 10 mg/ml. Cells were washed with 1X PBS and harvested using 1 ml of lysis buffer (10 mM Tris HCl pH 8, 150 mM NaCl, NP40 0.5% v/v, protease inhibitor 10 µl/ml). Samples were sonicated for 15 min at 4°C, then centrifuged at 1200 rpm for 15 min at 4°C. The supernatants obtained were transferred to a new microcentrifuge tube and then separated into 3 tubes. One tube was used as the input and the two other two tubes were used for the control and ORF1p IP, respectively. Each of these two tubes were then diluted to 1.5 ml with buffer (10 mM Tris HCl pH8, 150 mM NaCl, 10 µl/ml protease inhibitor) to dilute the NP40. The samples were then incubated overnight on a wheel at 4°C, then washed 3 times (first wash corresponds to post-bead samples) with 1 ml buffer (10 mM Tris HCl pH 8, 150 mM NaCl, 10 µl/ml protease inhibitor) using a magnet and then resuspended in the same buffer. The samples were boiled in Laemmli buffer (95°C, 10 min) and 20 µl of each sample were deposited on a 4-12% Nupage gel (Invitrogen, NP0336).

Mass Spectrometry

Online chromatography was performed with an RSLCnano system (Ultimate 3000, Thermo Scientific) coupled to a Q Exactive HF-X with a Nanospay Flex ion source (Thermo Scientific). Peptides were first trapped on a C18 column (75 μm inner diameter × 2 cm; nanoViper Acclaim PepMapTM 100, Thermo Scientific) with buffer A (2/98 MeCN/H2O in 0.1% formic acid) at a flow rate of 2.5 µL/min over 4 min. Separation was then performed on a 50 cm x 75 μm C18 column (nanoViper Acclaim PepMapTM RSLC, 2 μm, 100Å, Thermo Scientific) regulated to a temperature of 50°C with a linear gradient of 2% to 30% buffer B (100% MeCN in 0.1% formic acid) at a flow rate of 300 nL/min over 91 min. MS full scans were performed in the ultrahigh-field Orbitrap mass analyzer in ranges m/z 375–1500 with a resolution of 120 000 at m/z 200. The top 20 intense ions were subjected to Orbitrap for further fragmentation via high energy collision dissociation (HCD) activation and a resolution of 15 000 with the intensity threshold kept at 1.3 x 105. We selected ions with charge state from 2+ to 6+ for screening. Normalized collision energy (NCE) was set at 27 and the dynamic exclusion of 40s. For identification, the data were searched against the Mus Musculus (UP000000589_10090 012019) Uniprot database using Sequest HT through proteome discoverer (version 2.4). Enzyme specificity was set to trypsin and a maximum of two-missed cleavage sites were allowed. Oxidized methionine, Met-loss, Met-loss-Acetyl and N-terminal acetylation were set as variable modifications. Maximum allowed mass deviation was set to 10 ppm for monoisotopic precursor ions and 0.02 Da for MS/MS peaks. The resulting files were further processed using myProMS 109 v3.10.0. FDR calculation used Percolator and was set to 1% at the peptide level for the whole study. The label free quantification was performed by peptide Extracted Ion Chromatograms (XICs), reextracted by conditions and computed with MassChroQ version 2.2.21 110. For protein quantification, XICs from proteotypic peptides shared between compared conditions (TopN matching) with missed cleavages were used. Median and scale normalization at peptide level was applied on the total signal to correct the XICs for each biological replicate (n=5). To estimate the significance of the change in protein abundance, a linear model (adjusted on peptides and biological replicates) was performed, and p-values were adjusted using the Benjamini–Hochberg FDR procedure. Proteins with at least three peptides, identified in each biological replicates of ORF1p condition, a 10-fold enrichment and an adjusted p-value ≤ 0.05 were considered significantly enriched in sample comparisons. Unique proteins were considered with at least three peptides in all replicates. Protein selected with these criteria were used for Gene Ontology enrichment analysis and string network analysis.

The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consorti-um (http://proteomecentral.proteomexchange.org) via the PRIDE partner repository 111 with the dataset identifier PXD047160.

GO term and STRING network analysis

Gene Ontology analysis was performed using GO PANTHER 112 and String network physical interactions were retrieved using the STRING database v11.5 (https://string-db.org/) and then implemented in Cytoscape software 113.

Statistical analyses

In column comparisons, data in each column were tested for normality using two normality and lognormality tests (Shapiro-Wilk test and Kolmogorov-Smirnov test). Data which passed the normality tests were analyzed subsequently by a parametric test, data which did not pass the normality tests were analyzed by a non-parametric statistical test as indicated in the figure legends. The significance threshold was defined as p<0.05 except stated otherwise. Statistical analyses were done with PRISM software (v10).

eLife Assessment

Bonnifet et al. present data on the expression and interacting partners of the transposable element L1 in the mammalian brain. The work includes important findings addressing the potential role of L1 in aging and neurodegenerative disease. The reviewers conclude that several aspects of the study are well done. However, the experimental evidence presented supporting the L1 increase with aging is not fully conclusive and this finding remains incomplete in its current form.

Reviewer #1 (Public review):

Summary:

In this study, Bonnifet et al. profile the presence of L1 ORF1p in the mouse and human brain and report that ORF1p is expressed in the human and mouse brain specifically in neurons at steady state and that there is an age-dependent increase in expression. This is a timely report as two recent papers have extensively documented the presence of full-length L1 transcripts in the mouse and human brain (PMID: 38773348 & PMID: 37910626). Thus, the finding that L1 ORF1p is consistently expressed in the brain is important to document and will be of value to the field.

Strengths:

Several parts of this manuscript appear to be well done and include the necessary controls. In particular, the documentation of neuron-specific expression of ORF1p in the mouse brain is an interesting finding with nice documentation. This will be very useful information for the field.

Weaknesses:

Several parts of the manuscript appear to be more preliminary and need further experiments to validate their claims. In particular, the data suggesting expression of L1 ORF1p in the human brain and the data suggesting increased expression in the aged brain need further validation. Detailed comments:

(1) The expression of ORF1p in the human brain shown in Fig. 1j is puzzling. Why are there two strong bands in the WB? How can the authors be sure that this signal represents ORF1p expression and not non-specific labelling? While the authors discuss that others have found double bands when examining human ORF1p, there are also several labs that report only one band. This discrepancy in the field should at least be discussed and the uncertainties with their findings should be acknowledged.

(2) The data showing a reduction in ORF1p expression in the aged mouse brain is an interesting observation, but the effect magnitude of effect is very limited and somewhat difficult to interpret. This finding should be supported by orthogonal methods to strengthen this conclusion. For example, by WB and by RNA-seq (to verify that the increase in protein is due to an increase in transcription).

(3) The transcriptomic data using human postmortem tissue presented in Figure 4 and Figure 5 are not convincing. Quantification of transposon expression on short read sequencing has important limitations. Longer reads and complementary approaches are needed to study the expression of evolutionarily young L1s (see PMID: 38773348 & PMID: 37910626 for examples of the current state of the art). As presented, the human RNA data is inconclusive due to the short read length and small sample size. The value of including an inconclusive analysis in the manuscript is difficult to understand. With this data set, the authors cannot investigate age-related changes in L1 expression in human neurons.

(4) In line with these comments, the title should be changed to better reflect the findings in the manuscript. A title that does not mention "L1 increase with aging" would be better.

Reviewer #2 (Public review):

Summary:

Bonnifet et al. sought to characterize the expression pattern of L1 ORF1p expression across the entire mouse brain, in young and aged animals and to corroborate their characterization with Western blotting for L1 ORF1p and L1 RNA expression data from human samples. They also queried L1 ORF1p interacting partners in the mouse brain by IP-MS.

Strengths:

A major strength of the study is the use of two approaches: a deep-learning detection method to distinguish neuronal vs. non-neuronal cells and ORF1p+ cells vs. ORF1p- cells across large-scale images encompassing multiple brain regions mapped by comparison to the Allen Brain Atlas, and confocal imaging to give higher resolution on specific brain regions. These results are also corroborated by Western blotting on six mouse brain regions. Extension of their analysis to post-mortem human samples, to the extent possible, is another strength of the paper. The identification of novel ORF1p interactors in brain is also a strength in that it provides a novel dataset for future studies.

Weaknesses:

The main weakness of the study is that cell type specificity of ORF1p expression was not examined beyond neuron (NeuN+) vs non-neuron (NeuN-). Indeed, a recent study (Bodea et al. 2024, Nature Neuroscience) found that ORF1p expression is characteristic of parvalbumin-positive interneurons, and it would be very interesting to query whether other neuronal subtypes in different brain regions are distinguished by ORF1p expression. The data suggesting that ORF1p expression is increased in aged mouse brains is intriguing, although it seems to be based upon modestly (up to 27%, dependent on brain region) higher intensity of ORF1p staining rather than a higher proportion of ORF1+ neurons. Indeed, the proportion of NeuN+/Orf1p+ cells actually decreased in aged animals. It is difficult to interpret the significance and validity of the increase in intensity, as Hoechst staining of DNA, rather than immunostaining for a protein known to be stably expressed in young and aged neurons, was used as a control for staining intensity. The main weakness of the IP-MS portion of the study is that none of the interactors were individually validated or subjected to follow-up analyses. The list of interactors was compared to previously published datasets, but not to ORF1p interactors in any other mouse tissue.

The authors achieved the goals of broadly characterizing ORF1p expression across different regions of the mouse brain, and identifying putative ORF1p interactors in the mouse brain. However, findings from both parts of the study are somewhat superficial in depth.

This provides a useful dataset to the field, which likely will be used to justify and support numerous future studies into L1 activity in the aging mammalian brain and in neurodegenerative disease. Similarly, the list of ORF1p interacting proteins in the brain will likely be taken up and studied in greater depth.

Comments on revisions:

The co-staining of Orf1p with Parvalbumin (PV) presented in Supplemental Figure S5 is a welcome addition exploring the cell type-specificity of Orf1p staining, and broadly corroborates the work of Bodea et al. while revealing that Orf1p also is expressed in non-PV+ cells, consistent with L1 activity across a range of neuronal subtypes. The authors also have strengthened their findings regarding the increased intensity of ORF1p staining in aged compared to young animals, and the newly presented results are indeed more convincing. The prospect of increased neuronal L1 activity with age is exciting, and the results in this paper have provided the groundwork for ongoing discoveries in this area. While it is disappointing that no Orf1p interactors were followed up, this is understandable and the data are nonetheless valuable and will likely prove useful to future studies.

Author response:

The following is the authors’ response to the previous reviews

Reviewer #1 (Public review):

Summary:

In this study, Bonnifet et al. profile the presence of L1 ORF1p in the mouse and human brain and report that ORF1p is expressed in the human and mouse brain specifically in neurons at steady state and that there is an age-dependent increase in expression. This is a timely report as two recent papers have extensively documented the presence of full-length L1 transcripts in the mouse and human brain (PMID: 38773348 & PMID: 37910626). Thus, the finding that L1 ORF1p is consistently expressed in the brain is important to document and will be of value to the field.

Strengths:

Several parts of this manuscript appear to be well done and include the necessary controls. In particular, the documentation of neuron-specific expression of ORF1p in the mouse brain is an interesting finding with nice documentation. This will be very useful information for the field.

We thank the reviewer for this positive comment.

Weaknesses:

Several parts of the manuscript appear to be more preliminary and need further experiments to validate their claims. In particular, the data suggesting expression of L1 ORF1p in the human brain and the data suggesting increased expression in the aged brain need further validation. Detailed comments:

(1) The expression of ORF1p in the human brain shown in Fig. 1j is puzzling. Why are there two strong bands in the WB? How can the authors be sure that this signal represents ORF1p expression and not non-specific labelling? While the authors discuss that others have found double bands when examining human ORF1p, there are also several labs that report only one band. This discrepancy in the field should at least be discussed and the uncertainties with their findings should be acknowledged.

Please see also our extensive response to this comment we made in round #1 of the revisions.

As a summary, in response to the initial review, we included several lines of additional evidence in the revised manuscript:

siRNA-mediated knockdown of ORF1p in human neurons, resulting in ≈50% signal reduction using the antibody in question (Suppl. Fig. 2C) immunoprecipitation using the human ORF1p antibody in question confirming signal specificity (Suppl. Fig. 2B) use of a second antibody in immunostainings, including a new control (Suppl. Fig. 2E) and a revised discussion acknowledging the uncertainty surrounding the lower band:

“The double band pattern in Western blots has been observed in other studies for human ORF1p outside of the brain as well as for mouse ORF1p. […] The nature of the lower band is unknown, but might be due to truncation, specific proteolysis or degradation.”

We have also now added more content to the paragraph starting from line 183 : "While there is some discrepancy in the field, the double band pattern in Western blots..."

To our understanding, this combination of independent methods using two antibodies and complementary validation strategies supports the presence of ORF1p in human brain tissue.

(2) The data showing a reduction in ORF1p expression in the aged mouse brain is an interesting observation, but the effect magnitude of effect is very limited and somewhat difficult to interpret. This finding should be supported by orthogonal methods to strengthen this conclusion. For example, by WB and by RNA-seq (to verify that the increase in protein is due to an increase in transcription).

This would indeed be valuable but at this point, we will not be able to perform these experiments at this point (please also see revision #1 for a more detailed answer)

(3) The transcriptomic data using human postmortem tissue presented in Figure 4 and Figure 5 are not convincing. Quantification of transposon expression on short read sequencing has important limitations. Longer reads and complementary approaches are needed to study the expression of evolutionarily young L1s (see PMID: 38773348 & PMID: 37910626 for examples of the current state of the art). As presented, the human RNA data is inconclusive due to the short read length and small sample size. The value of including an inconclusive analysis in the manuscript is difficult to understand. With this data set, the authors cannot investigate age-related changes in L1 expression in human neurons.

Please see also our extensive response to this comment we made in round #1 of the revisions.

In the revised version, we have added further statistical analyses, incorporated locus-specific mappability scores and provided an even more nuanced interpretation of our findings, as illustrated in lines 390 and 427.

We have acknowledged the limitations of short-read sequencing in this context, while referencing established methodologies (e.g., Teissandier et al., 2019) and recent benchmarking studies (e.g., Schwarz et al., 2022) that validate the use of such data under specific precautions—many of which we have implemented.

Given these considerations, and with the guidance of a co-author with specific expertise in TE bioinformatics, we believe our approach is justified and robust.

(4) In line with these comments, the title should be changed to better reflect the findings in the manuscript. A title that does not mention "L1 increase with aging" would be better.

In line with our response to Point (3), we prefer to retain the current analyses and discussion, which we believe strike an appropriate balance between caution and added scientific value.

Reviewer #2 (Public review):

Summary:

Bonnifet et al. sought to characterize the expression pattern of L1 ORF1p expression across the entire mouse brain, in young and aged animals and to corroborate their characterization with Western blotting for L1 ORF1p and L1 RNA expression data from human samples. They also queried L1 ORF1p interacting partners in the mouse brain by IP-MS.

Strengths:

A major strength of the study is the use of two approaches: a deep-learning detection method to distinguish neuronal vs. non-neuronal cells and ORF1p+ cells vs. ORF1p- cells across large-scale images encompassing multiple brain regions mapped by comparison to the Allen Brain Atlas, and confocal imaging to give higher resolution on specific brain regions. These results are also corroborated by Western blotting on six mouse brain regions. Extension of their analysis to post-mortem human samples, to the extent possible, is another strength of the paper. The identification of novel ORF1p interactors in brain is also a strength in that it provides a novel dataset for future studies.

We thank the reviewer for these positive comments.

Weaknesses:

The main weakness of the IP-MS portion of the study is that none of the interactors were individually validated or subjected to follow-up analyses. The list of interactors was compared to previously published datasets, but not to ORF1p interactors in any other mouse tissue.

As we had stated in the first round of revision, the list of previously published datasets does include a mouse dataset with ORF1p interacting proteins in mouse spermatocytes (please see line 478-4479: “ORF1p interactors found in mouse spermatocytes were also present in our analysis including CNOT10, CNOT11, PRKRA and FXR2 among others (Suppl_Table4).”) -> De Luca, C., Gupta, A. & Bortvin, A. Retrotransposon LINE-1 bodies in the cytoplasm of piRNA-deficient mouse spermatocytes: Ribonucleoproteins overcoming the integrated stress response. PLoS Genet 19, e1010797 (2023)). We agree that a validation of protein interactors of ORF1p in the mouse brain would have been valuable. However, the significant overlap with previously published interactors highlights the validity of our data. As reviewer #2 points out in the comments on revisions, we hope that follow-up studies will address these points and we anticipate that this list of ORF1p protein interactors in the mouse brain will be of further use for the community.

Comments on revisions:

The co-staining of Orf1p with Parvalbumin (PV) presented in Supplemental Figure S5 is a welcome addition exploring the cell type-specificity of Orf1p staining, and broadly corroborates the work of Bodea et al. while revealing that Orf1p also is expressed in non-PV+ cells, consistent with L1 activity across a range of neuronal subtypes. The authors also have strengthened their findings regarding the increased intensity of ORF1p staining in aged compared to young animals, and the newly presented results are indeed more convincing. The prospect of increased neuronal L1 activity with age is exciting, and the results in this paper have provided the groundwork for ongoing discoveries in this area. While it is disappointing that no Orf1p interactors were followed up, this is understandable and the data are nonetheless valuable and will likely prove useful to future studies.

Thank you for your time and constructive comments.

Reviewer #1 (Recommendations for the authors):

We would recommend that the human RNA-seq analysis is removed from the manuscript. The human RNA data is inconclusive due to the short read length and small sample size. The value of including an inconclusive analysis in the manuscript is difficult to understand. With this data set, the authors cannot investigate age-related changes in L1 expression in human neurons.

Reviewer #2 (Recommendations for the authors):

Thank you for addressing my suggestions. I have no further recommendations at this time.

Version 3

Abstract

Abstract

Recent studies have established a reciprocal causal link between aging and the activation of transposable elements, characterized in particular by a de-repression of LINE-1 retrotransposons. These LINE-1 elements represent 21% of the human genome, but only a minority of these sequences retain the coding potential essential for their mobility. LINE-1 encoded proteins can induce cell toxicity implicated in aging and neurodegenerative diseases. However, our knowledge of the expression and localization of LINE-1-encoded proteins in the central nervous system is limited. Using a novel approach combining atlas-based brain mapping with deep-learning algorithms on large-scale pyramidal brain images, we unveil a heterogeneous, neuron-predominant and widespread ORF1p expression throughout the murine brain at steady-state. In aged mice, ORF1p expression increases significantly which is corroborated in human post-mortem dopaminergic neurons by an increase in young LINE-1 elements including those with open reading frames. Mass spectrometry analysis of endogenous mouse ORF1p revealed novel, neuron-specific protein interactors. These findings contribute to a comprehensive description of the dynamics of LINE-1 and ORF1p expression in the brain at steady-state and in aging and provide insights on ORF1p protein interactions in the brain.

Introduction

Only about 2 % of the human genome are DNA sequences that will be translated into protein. The remaining 98 % are comprised of introns, regulatory elements, non-coding RNA, pseudogenes and repetitive elements including transposable elements. However, some sequences in what is generally considered “non-coding genome” do in fact contain sequences which encode proteins. This is true for specific lncRNAs which can encode peptides or functional proteins 1 but also for a few copies of two transposable element families, Long INterspersed Element-1 (LINE-1) and Human Endogenous RetroViruses (HERV). Non-functional copies of retrotransposons, to which LINE-1 and HERV belong, cover about 44% 2 of the human genome as remnants of an evolutionary ancient activity. Depending on the source, about 100 3 to 146 4 full-length LINE-1 elements with two open reading frames encoding ORF1 and ORF2 are present in the Human reference genome (GRCh38 Genome Assembly) and several incomplete HERV sequences encoding either or any combination of envelope (env), gag, pro or pol 5. The LINE-1 encoded protein ORF1p, an RNA binding protein with “cis” preference 6, 7, and ORF2p, an endonuclease and reverse transcriptase 8, 9 are required for the mobility of LINE-1 elements. As many other transposable elements (TEs) including HERVs, LINE-1 elements are repressed by multiple cellular pathways. It was thus generally thought that TEs are repressed in somatic cells with no expression at steady-state 10–12. However, the aging process reduces the reliability of these repressive mechanisms 13. It is now, 31 years after the initial proposition of the “transposon theory of aging” by Driver and McKechnie 14, still a matter of debate whether TE activation can be both, a cause and a consequence of aging 15, 16.

Sparse data has shown that the LINE-1 encoded protein ORF1p is expressed at steady-state in the mouse ventral midbrain 17, the mouse hippocampus 18 and in some regions of the human post-mortem brain 19 and recent data informed about the presence of full-length transcripts in cancer cells, human epithelial cells and mouse hippocampal neurons 20. Repression of LINE-1 might thus be incomplete and if so, it remains unclear how cells then prevent cell toxicity associated with LINE-1 encoded protein activity. Indeed, LINE-1 encoded proteins have been demonstrated to induce genomic instability (ORF2p endonuclease-mediated 17, 21–26) and inflammation (ORF2p reverse transcriptase-mediated 27–29) and these cellular activities might be causally related to organismal aging, cancer, autoimmune and neurological diseases 30. For instance, LINE-1 activation can drive neurodegeneration of mouse dopaminergic neurons 17, of drosophila neurons 31, 32 and of mouse Purkinje neurons 33 which can be at least partially rescued with nucleoside analogue reverse transcriptase inhibitors (NRTIs) or other anti-LINE-1 strategies. NRTIs are currently being tested in several clinical trials designed to target either the RT of HERVs or the RT encoded by the LINE-1 ORF2 protein. It is not known today, however, to which extend LINE-1 encoded proteins are expressed at steady-state throughout the mouse and human brain, whether there is cell-type specificity and whether activation of LINE-1 encoded proteins is associated with brain aging or human neurodegeneration. Here, using a deep-learning assisted cellular detection methodology applied to pyramidal large-scale images of the mouse brain mapped to the Allen mouse brain atlas combined with post-mortem human brain imaging, co-IP mass spectrometry and transcriptomic analysis of LINE-1 expression, we describe a brain-wide map of ORF1p expression and interacting proteins at steady-state and in the context of aging. We find a heterogenous but widespread expression of ORF1p in the mouse brain with predominant expression in neurons. In aged mice, neuronal ORF1p expression increases brain-wide and in some brain regions to up to 27%. In human dopaminergic neurons, young LINE-1 transcripts and specific full-length and coding LINE-1 copies are increased in aged individuals. We further describe endogenous mouse ORF1p interacting proteins revealing known interactors and unexpected interacting proteins belonging to GO categories related to RNA metabolism, chromatin remodeling, cytoskeleton and the synapse.

Results

Widespread and heterogenous expression of the LINE-1 encoded ORF1p protein in the wildtype mouse brain

To investigate the expression pattern and intensities of endogenous LINE-1 encoded ORF1p protein throughout the entire mouse brain, we devised a deep-learning assisted cellular detection methodology applied to pyramidal large-scale images using a comprehensive workflow complemented by an approach based on confocal imaging as schematized in Figure 1A. Briefly, starting from sagittal slide scanner images of the mouse brain, we defined anatomical brain regions by mapping the Allen Brain Atlas onto the slide scanner images using Aligning Big Brains & Atlases (ABBA) 34. We then employed a deep-learning detection method to identify all cell nuclei (Hoechst) and categorize all detected cells into neuronal cells (NeuN+) or non-neuronal cells (NeuN-) and ORF1p-expressing cells (ORF1p+) or cells that do not express ORF1p (ORF1p-). This workflow allowed us then to characterize the cell identity of ORF1p+ cells and ORF1p intensity throughout the whole brain but also in specific anatomical regions. In parallel, we completed the approach using confocal microscopy on selected anatomical regions allowing for comparison with higher resolution. Importantly, the specificity of the ORF1p antibody, a widely used, commercially available antibody 18, 35–39, was confirmed by blocking the ORF1p antibody with purified mouse ORF1p protein resulting in the complete absence of immunofluorescence staining (Suppl Fig. 1A), by using an in-house antibody against mouse ORF1p17 which colocalized with the anti-ORF1p antibody used (Suppl Fig. 1B, quantified in Suppl Fig. 1C), by immunoprecipitation and mass spectrometry used in this study (Fig 6A, Suppl Table 2) and by siRNA-mediated knock-down of ORF1 in a differentiated mouse dopaminergic cell line (MN9D; Suppl Fig. 1D). Unexpectedly, we found a generalized and widespread expression of ORF1p throughout the brain of wildtype mice (Fig. 1B; Swiss OF1 mice, three months-old; whole brain except regions with particularly high cellular density (cerebellum, hippocampus, olfactory bulb) which impedes nuclei detection by deep-learning. ORF1p is detectable in all regions and subregions analyzed with heterogenous expression patterns (density and intensity) per region/subregions. The ten regions shown in Figure 1B exemplify visible different densities of ORF1p+ cells with varying levels of expression. Notably, the expression pattern of ORF1p in the hippocampus is similar to what has recently been published 18 (Fig. 1B, panel 2). Throughout the entire brain, the mean density of ORF1p+ cells per mm² was ≈ 305 ±18 (mean ±SEM), representing up to 20% of all detected cells (Fig. 1C). ORF1p+ cells in each mouse brain analyzed showed up to eight-fold disparities in intensity between low- and high-expressed cells (Fig. 1C). We then quantified nine anatomical regions according to the Allen Brain Atlas on four brains of three-month old mice (Fig. 1D) using the automated workflow (Fig. 1A) with regard to cell density (Fig. 1D), cell proportions (Fig. 1E) and fluorescent intensity of ORF1p+ cells (Fig. 1F). This approach permitted the analysis of about 10 000 ORF1p+ cells per animal highlighting the power of our large-scale analysis. Densities of ORF1p+ cells ranged from the lowest density in the hindbrain with 154 ±19 cells per mm2 (mean ±SEM) to the highest density of ORF1p+ cells in the isocortex with 451 ±44 cells per mm2 (mean ±SEM) and the thalamus with 446 ±50 cells per mm2 (mean ±SEM). The proportion of ORF1p+ cells per anatomical region fluctuated between 10% ±2.1 (ventral striatum, mean ±SEM) and 31% ±1.6 (thalamus, mean ±SEM). The dorsal striatum (“striatal dorsal” in the Allen Brain Atlas denomination) exhibited the lowest ORF1p expression intensity (658 ±3 mean ±SEM) of all regions tested, the hindbrain the highest mean intensity of ORF1p per cell (mean ±SD 1221 ±548) as illustrated in Figure 1B and quantified in Figure 1D and 1F. Interestingly, cell density did not correlate with expression levels. Dorsal and ventral striatum for instance displayed similar ORF1p intensities per cell but exhibited significant differences in ORF1p cell density and proportion. The “midbrain motor” region as defined by the Allen Brain Atlas showed an intermediate cell density (mean ±SEM 265 ±16 cells per mm2) and a rather high ORF1p expression intensity (mean ±SEM 1006 ±533). Statistical analysis comparing mean density of ORF1p+cells per mm2 or mean intensity per ORF1p+ cells among regions confirmed the heterogeneity concerning ORF1p expression throughout the mouse brain (Fig. 1D, 1F). Slide scanner and confocal images revealed an exceptionally high ORF1p expression intensity in the ventral region of the midbrain which we identified as the Substantia nigra pars compacta (SNpc). This region displayed an important density of ORF1p+ cells and a comparatively high level of ORF1p expression as illustrated by confocal imaging (Fig. 1B, panel 8), but could not be quantified independently with our brain-wide approach due to the geometrically-complex anatomy of this region and its small size (subregion-level in the Allen Brain Atlas hierarchy). Another region which could not be included in our brain-wide analysis was the cerebellum due to its extremely high density of cell nuclei. However, slide scanner and confocal imaging (Fig. 1B, panel 10) revealed that ORF1p is expressed in Purkinje cells, while not detectable in the molecular or granular layers.

Widespread and heterogenous expression of ORF1p protein in the mouse brain

(A) Schematic representation of the unbiased cell detection pipeline on large-scale and confocal images. Immunofluorescent images on sagittal mouse brain slices were acquired on a digital pathology slide scanner or on a confocal microscope (DNA stain: Hoechst, neuronal marker; NeuN, protein of interest: ORF1p). Pyramidal images aquired with the slide scanner were then aligned with the hierarchical anatomical annotation of the Allen Brain Atlas using ABBA. Once the regions defined, a deep-learning based detection of cell nuclei (Hoechst staining, Stardist) and cell cytoplasm (NeuN staining, Cellpose) was performed on each sub-region of the atlas. Objects were filtered according to the background intensity measured in each sub-region for each channel (NeuN and ORF1p). The identity and intensity measures were analyzed at the regional and whole brain level. In parallel, confocal images (multiple z-stacks) of two selected regions (frontal cortex and ventral midbrain) were also acquired and identity and intensity were quantified using Cellpose. (B) Widespread and heterogenous expression of the LINE-1 encoded protein ORF1p in the mouse brain. Representative image of ORF1p immunostaining (orange) of a sagittal section of the brain of a young (three months-old) mouse acquired on a slide scanner. Scale bar = 1mm. (1-10) Representative images of confocal images of immunostainings showing ORF1p expression (orange) in 10 different regions of the mouse brain. Nuclei are represented in blue (Hoechst), scale bar = 50µm. (1) Isocortex, (2) Hippocampus, (3) Striatum dorsal, (4) Thalamus, (5) Midbrain motor, (6) Pallidum, (7) Hypothalamus, (8) Substantia nigra pars compacta, (9) Hindbrain, (10) Cerebellum. ORF1p expression profile in the mouse brain. The entire mouse brain with the exception of the olfactory bulb and the cerebellum were analyzed according to the pipeline on large-scale images described in (A). (C) Bar plot showing the total number of ORF1p+ cells per mm² in the mouse brain. Data is represented as mean ±SEM, n=4 mice (top). Bar plot indicating the proportion of ORF1p+ cells compared to all cells detected. Data is represented as mean ±SEM, n=4 mice (labeled M1 to M4), 202001 total cells analyzed (middle). Scatter plot showing the mean intensity of ORF1p per ORF1p+ cell. Data is represented as mean ±SD, n=4 mice, 40999 ORF1p+ cells analyzed (bottom). (D-F) ORF1p expression profile (density, proportion and expression) in defined anatomical regions of the mouse brain. Nine anatomical regions as defined by the Allen Brain Atlas and mapped onto sagittal brain slices (four three-month-old Swiss/ OF1) with ABBA were analyzed using the pipeline on large-scale images described in (A). (D) ORF1p+ cell density in 9 different regions. Bar plot showing the number of ORF1p+ cells per mm². Data is represented as mean ±SEM; *p<0.05; **p<0.01; adjusted p-value, one-way ANOVA followed by a Benjamin-Hochberg test (E) Proportion of ORF1p positive cells in 9 different regions. Bar plot showing the proportion of ORF1p+ cells among all cells detected per region. Data is represented as mean ±SEM. (F) Mean ORF1p expression per cell in 9 different regions. Dot plot showing the mean intensity of ORF1p signal per ORF1p+ cell in 9 different regions. Data is represented as mean ±SD. The number of analyzed cells per region is indicated in the figure. *p<0.05; **p<0.01; ***p<0.001; ****p<0.0001; adjusted p-value, nested one-way ANOVA followed by Sidak’ multiple comparison test. (G) ORF1p expression in the mouse frontal cortex and ventral midbrain. Confocal images with multiple z-stacks. Dot plot representing the mean intensity levels of ORF1p per ORF1p+ cells. Four three-month-old Swiss/ OF1 mice (labeled as M1 to M4) are represented each by a different color, the scattered line represents the median. ****p<0.0001, nested one-way ANOVA. Total cells analyzed = 4645. (H-I) ORF1p expression in the mouse frontal cortex and the ventral midbrain. (H) Western blots showing ORF1p (top) and actin expression (bottom) in four individual mice per region which were quantified in (I) using actin as a reference control. The signal intensity is plotted as the fold change of ORF1p expression in the ventral midbrain to ORF1p expression in the frontal cortex. *p<0.05; two-sided, unpaired student’s-test. (J) ORF1p expression in three regions of the human brain. Western blot showing human ORF1p expression in the cingulate gyrus (CG), frontal cortex (FC) and cerebellum (CB) of post-mortem tissues from a healthy individual. ORF1p (Top), Actin (bottom). For the full Western blot image please see Suppl Fig. 2A.

In order to confirm ORF1p expression by an independent method, we performed Western blot analysis on six micro-dissected regions from the mouse brain (Swiss/OF1 mouse, three-month old). As shown in Suppl Fig. 1E, ORF1p is expressed in all six regions with varying expression levels confirming the overall presence of ORF1p throughout the brain. We then chose two regions with significantly divergent ORF1p expression intensities as identified and quantified on pyramidal large-scale images: the frontal cortex (low) and the ventral midbrain (intermediate to high). We confirmed a significant higher expression of ORF1p in the ventral midbrain compared to the frontal cortex using an approach based on the unbiased, automated quantification of multiple z-stacks using a confocal microscope (Fig. 1G) and by Western-blotting on micro-dissected regions (Fig. 1H, 1I). In concordance with the findings stemming from the large-scale image quantification pipeline (Fig. 1F), the ventral midbrain showed ≈ 2-times higher expression of ORF1p than the frontal cortex as quantified in Figure 1G (1.8-fold) and Figure 1I (2.3-fold) validating our cellular detection methodology for pyramidal large-scale imaging and underscoring the heterogeneity of ORF1p expression levels in the mouse brain.

To investigate intra-individual expression patterns of ORF1p in the post-mortem human brain, we analyzed three brain regions of a neurologically-healthy individual (Fig. 1J, entire Western blot membrane in Suppl Fig. 2A) by Western blotting using a commercial and well characterized antibody which we further validated by several means. While there is some discrepancy in the field, the double band pattern in Western blots has been observed in other studies for human ORF1p outside of the brain 40, 41 as well as for mouse ORF1p 42. The nature of this lower band is unknown, but it might be due to truncation 43, specific proteolysis or degradation. Nevertheless, it appears that in cell culture models, a single ORF1p band is observed, whereas in murine and human samples, the ORF1p band is, to our knowledge, consistently associated with a lower molecular weight band 35–37, 40–42, 44, 45. We validated the antibody by immunoprecipitation and siRNA knock-down in human dopaminergic neurons in culture (differentiated LUHMES cells, Suppl Fig. 2B and 2C) where we detected in most cases the upper band only. ORF1p was expressed at different levels in the human post-mortem cingulate gyrus, the frontal cortex and the cerebellum underscoring a widespread expression of human ORF1p across the human brain. This was in accordance with ORF1p immunostainings of the human post mortem cingulate gyrus (Fig. 2H and Suppl Fig. 2E) and frontal cortex (Suppl Fig. 2E), with an absence of ORF1p staining when using the secondary antibody only (Suppl Fig. 2E).

ORF1p is predominantly expressed in neurons in the mouse brain

(A) ORF1p expression is absent in the white matter (corpus callosum) and predominantly expressed in neurons. Proportion of ORF1p+/NeuN+, ORF1p+/NeuN-, ORF1p-/NeuN+ and ORF1p-/NeuN- cells in the white matter (corpus callosum) and the grey matter (left) and in nine different regions (right) analyzed by the cell detection pipeline on large-scale images presented in Figure1A. Exact values can be found in Suppl_Table1. (B) Representative confocal microscopy images showing ORF1p (red) and NeuN expression (green) in two different regions of the mouse brain. The bottom images show the merge of the two stainings, an overlap of both markers is represented in orange. z-projections; scalebar = 25µm. (C) Proportion of neurons expressing ORF1p in the frontal cortex and ventral midbrain quantified on confocal images. ns: non-significant; chi-square test on the cell number of the different cell-types analyzed; n=4 mice, data is represented as mean ± SEM. (D) Proportion of ORF1p+ cells identified as NeuN+ or NeuN- in two different regions, analyzed by confocal microscopy on multiple z-stacks. ns: non-significant; chi-square test, n=4 mice. (E) ORF1p does not colocalize with glial or microglial cell markers. Representative confocal microscopy images showing ORF1p staining (red) and three different glial cell (GFAP, Sox9, S100β) or microglial (Iba1) markers (white). Note that Iba1 antibody (rabbit) was used with the ORF1p 09 antibody (guinea pig, in house) z-projections, scalebar = 50µm. (F-G) Separation of neuronal and non–neuronal - cells by FACS confirms predominant neuronal expression of ORF1p. (F) Neuronal (NeuN+) and non-neuronal (NeuN-) cells isolated by fluorescent activated cell sorting (FACS). Dot plots showing autofluorescence versus an appropriate control antibody (IgG rabbit 647; left) and an antibody against NeuN (AB 657, right). The P4 window represents isolated NeuN+ cells (pink) and the P5 fraction NeuN- cells (orange) containing the same number of cells as sorted in P4 for comparison, others NeuN- are represented in blue. (G) Western blot showing ORF1p expression in NeuN- and NeuN+ FACS- sorted cells stemming from Figure F (A and B representing two different FACS experiments). (H) Representative confocal microscopy images showing ORF1p (red), NeuN (green) and Hoechst (blue) in the cingulate gyrus of the human brain. z-projection; scalebar = 25µm (left). Examples of individuals neurons expressing ORF1p or not are shown on the right panel. z-projection; scalebar = 5 µm (right). (I) Proportion of ORF1p+ cells identified as NeuN+ or NeuN- in the human cingulate gyrus, analyzed by confocal microscopy on multiple z-stacks. (J) Proportion of neurons expressing ORF1p in the human cingulate gyrus, analyzed by confocal microscopy on multiple z-stacks.

In summary, our findings reveal the consistent presence of ORF1p expression throughout the mouse brain in all anatomical regions analyzed with high regional variability in terms of density of ORF1p+ cells and ORF1p+ cell intensity. ORF1p is also expressed in the human brain in at least three brain regions. This finding raises several questions concerning cell-type identity of ORF1p expressing cells and potential functions or consequences of ORF1p expression in the mouse and human brain at steady-state.

ORF1p is predominantly expressed in neurons

Following our observation of a wide-spread expression of endogenous ORF1p throughout the brain, we first addressed the question of the cellular identity of ORF1p+ cells. To this end, we used the neuron-specific marker NeuN, commonly used to identify post-mitotic neurons in the central nervous system 46. This allowed us to determine the proportion of neuronal (NeuN+) or non-neuronal cells (NeuN-) expressing ORF1p (ORF1p+) or not (ORF1p-). Making use of our large-scale imaging approach (Fig. 1A), we observed drastic dissimilarities in detected cellular proportions between the white and grey brain matter. As expected, we observed only 1% of NeuN+ cells in the white matter (corpus callosum; Fig. 2A) validating both, the neuronal marker NeuN as such and the ABBA superposition of the Allen Brain Atlas onto the sagittal brain slices. In the grey matter, our approach detected 30.5% NeuN+ cells (dark red and yellow bars in Fig. 2A) which, according to the literature, should include all post-mitotic neurons with only minor exceptions 46–49 and corresponds to the reported proportion of neurons present in the mouse brain 50. The nine identified grey matter regions in Fig. 2A display the proportions of the different cell types per region. The proportion of all cells in a given region which are positive for ORF1p (dark red bars) differed between regions (lowest proportion: hindbrain: 7%; highest proportion dorsal striatum: 26.6%). In the isocortex and the midbrain motor-related regions, the majority of neurons detected express ORF1p (54% and 59% by large-scale analysis, Suppl Fig. 3A; 68.7% and 68.8% by confocal imaging, Fig. 2B, quantified in C), while in the midbrain sensory related regions the proportion dropped to 25% whereas it reached 82% in the thalamus (Suppl Fig. 3A). Altogether, nearly half of all NeuN+ cells throughout the mouse brain expressed ORF1p (mean of all regions: 48.2%; Suppl Fig. 3A). Regarding the cell identity of ORF1p+ cells brain-wide, more than 70% were identified as neuronal by the large-scale approach (Suppl Fig. 3B). This contrasted somewhat with results obtained by the second approach using confocal imaging on multiple z-stacks which indicated that 91.3% (frontal cortex) and 88.5% (ventral midbrain) of ORF1p+ cells were neuronal (Fig. 2D). This difference in percentages of ORF1p+ expressing neurons among all neurons between the large-scale image cell detection methodology and the confocal workflow is most probably due to technical limitations inherent to the large-scale pipeline. Indeed, with the latter approach, region-dependent differences in cell density and signal intensity might be the cause for an underestimation of the proportion of ORF1p+ cells being neuronal due to difficulties in cell detection by StarDist/Cellpose (high cell density) on a single focal plan, technical difficulties which are widely reduced by the multiple z-stack based approach when using a confocal microscope. Moreover, the large-scale pipeline involves background measurements in each sub-region in order to apply stringent filtering. This, however, results in a loss of true positives cells, but avoids cells which are out-of-focus, the presence of which is inherent to the slide scanner microscope which lacks optical sectioning (see Materials & methods section for evaluated model performance). Notably, frontal cortex and ventral midbrain present similar proportion of neurons expressing ORF1p (Fig. 2C), although the percentage of NeuN+ cells between these two regions is significantly different (Suppl Fig. 3C). As we could not rule out that ORF1p might also be expressed in non-neuronal cells, we turned to non-neuronal markers specific for different glial cell populations using two different astrocytic markers (GFAP, Sox9), the astro-and oligodendrocytic marker S100β and the microglial marker Iba1 50, 51 and performed co-staining with ORF1p followed by confocal imaging as illustrated in Figure 2E. We screened multiple images of frontal cortex, ventral midbrain, hippocampus and striatum and did not find a single ORF1p+ cell which could unambiguously be defined as non-neuronal. This indicated that ORF1p is not or only very rarely expressed in non-neuronal cells. To further confirm the predominant presence of expression of ORF1p in neurons and the absence of ORF1p expression in non-neuronal cells, we used fluorescence-activated cell sorting (FACS) to isolate neurons (using a NeuN antibody) and non-neuronal cells (NeuN-) from the adult mouse brain followed by Western blotting with an antibody against ORF1p (Fig. 2F, 2G). As described above, this antibody is well characterized, extensively used and was validated further in this study. After FACS-sorting of neurons from the adult mouse brain using an antibody against NeuN (Fig. 2F, G), we detected ORF1p exclusively in the neuronal population (NeuN+, internal control = heavy chain), confirming the results based on two different imaging approaches. Finally, to assess whether predominant, if not exclusive ORF1p expression in neurons is mouse brain specific or a pattern also applicable to the human brain, we investigated the identity of ORF1p expressing cells in the post-mortem cingulate gyrus of a healthy human brain. Similar to what we found in the mouse brain, we observed sparse NeuN expression in the white and extensive NeuN staining in the grey matter corresponding to the cortical layers (Suppl Fig. 2D, grey and white matter separated by a dashed line) with ORF1p+ cells predominantly located in the grey matter (confocal images in Fig. 2H, Suppl Fig 2E are located in the grey matter). All cells stained by ORF1p were co-stained with NeuN indicating that ORF1p was expressed in neuronal cells in the human brain (Fig. 2H). However, due to the lower signal quality inherent to human post-mortem sections compared to mouse sections, the identity of ORF1p+ cells was estimated to be 80% neuronal by the automated image analysis pipeline of confocal images (Fig. 2I), although no ORF1p+ / NeuN- cells could be clearly identified. Of all neurons identified, 37.2% were ORF1p+ (Fig. 2J), indicating that, similar to the mouse brain, only a fraction of neurons express ORF1p (Fig. 2H, right).

Next, we asked the question of a neuron-subtype specific expression of ORF1p. Our previous study had revealed a higher expression of ORF1p in tyrosine-hydroxylase (TH) positive neurons compared to TH-negative neurons in the mouse ventral midbrain 17. A recent study reported that endogenous LINE-1 RNA and ORF1p expression were higher in parvalbumin (PV)-positive interneurons compared to PV-negative neurons in the mouse hippocampus 18. To address the question of a generalized co-expression of ORF1p and PV cells, we co-stained sagittal brain sections of young mice with antibodies against ORF1p and PV and, in some cases the lectin Wisteria floribunda agglutinin (WFA), which specifically stains glycoproteins surrounding PV+ neurons. Confocal imaging on several brain regions including the hippocampus, cortex, cerebellum, hindbrain, ventral midbrain and thalamus revealed ORF1p+ neurons co-expressing PV, but also many examples of equally intense ORF1p+ neurons that do not express PV (Suppl Fig. 4).

In summary, ORF1p expression in the mouse and human brain is widely restricted to neurons of which a proportion express ORF1p. This raises the question of the function and consequences of ORF1p expression specifically in neurons but also on the dynamic regulation of this expression upon exogenous (exposome) or endogenous (aging) challenges.

ORF1p expression is increased in the aged mouse brain

ORF1p is expressed at steady-state throughout the brain, but whether this expression is dynamically regulated remains unknown. Aging has been linked to LINE-1 regulation in some studies 16, 52 potentially as both, a trigger and as a consequence of LINE-1 activation, but whether this is true for the brain and if yes, whether this might be region-specific has not been investigated brain-wide. We therefore addressed the question whether advanced age was paralleled by a change of expression patterns or expression levels of ORF1p in the brain. We first analyzed ORF1p expression levels comparing young (3-month) to aged (16-month) mouse brains using the cell detection workflow applied to large-scale images described in Figure 1A. Interestingly, the mean intensity of ORF1p expression increased moderately but significantly with advanced age throughout the brain (13% increase brain-wide; n=4 young mice; n=4 aged mice; p=0.03; Fig. 3A). This was in contrast to another protein, NeuN, which we used as a control and whose intensity did not change between young and aged brains (n=4 young mice; n=4 aged mice; p=0.27; Fig. 3B). Frequency distribution analysis unveiled a shift in ORF1p mean expression per cell in aged mice (Fig. 3C). Importantly, the Hoechst mean intensity within nuclei of ORF1p+ cells, serving as an internal control, showed no significant change (Fig. 3D). Among nine analyzed regions, five demonstrated a general increase in ORF1p mean intensity per cell in aged mice (p≤0.05), a change independent from inter-individual variations in both young and aged mice (Fig. 3E). The increase of ORF1p expression (fold change intensity) throughout the brain, reaching nearly a 30% increase in some regions, is represented on the heatmap in Figure 1F. These results were confirmed by the confocal imaging approach; ORF1p expression in the frontal cortex remained unchanged but increased significantly in the ventral midbrain region in aged mice as shown in Figure 3G and quantified in Figure 3H.

ORF1p expression is increased in some regions of the aged mouse brain

(A) ORF1p mean intensity per ORF1p+ cell in the brain, analyzed on large-scale images. Dot plot showing the ORF1p mean expression per ORF1p+ cell in young (Y1-4) and aged (O1-4) mice in the whole brain (except cerebellum and olfactory bulb). 74985 total cells were analyzed; * p<0.05, nested two-way ANOVA; n=4 mice per condition, data is represented as mean ± SEM. (B) NeuN mean intensity per NeuN+ cell in the brain, analyzed on large-scale images. Dot plot showing the NeuN mean expression per NeuN+ cell in young (Y1-4) and aged (O1-4) mice in the whole brain (except cerebellum and olfactory bulb). Nested two-way ANOVA; n=4 mice per condition, data is represented as mean ± SEM. (C) Frequency distribution of ORF1p mean intensity in ORF1p+ cells of young (blue) and aged (orange) mice. ***p<0.001, Kolmogorov-Smirnov test. (D) Frequency distribution of Hoechst mean intensity in the nuclei of OrF1p+ cells of young (blue) and aged (orange) mice. ns: non-significant, Kolmogorov-Smirnov test. (E) Mean ORF1p expression per ORF1p+ cell in nine different anatomical regions. Dot plot showing the ORF1p mean expression per ORF1p positive cell (n=74985). Adjusted p-value are represented, two-tailed nested t-test followed by a Benjamin, Krieger and Yukutieli test; n=4 mice per region, data is represented as mean ± SEM. (F) Color-coded representation of fold-changes of ORF1p expression comparing young and aged brains. Represented is the fold-change in percent (aged vs young) of the “mean of the mean” ORF1p expression per ORF1p+ cell quantified and mapped onto the nine different regions analyzed as shown in (J). (G) Representative confocal microscopy images showing increased ORF1p expression (red) in the ventral midbrain region of aged mice (one z plan is shown). Cell nuclei are shown in blue (Hoechst staining). Scalebar = 50µm. (H) ORF1p expression is increased in the ventral midbrain of aged mice. Dot plot representing ORF1p expression in two different regions of young and aged mice analyzed on confocal images with multiple z-stacks; total cells analyzed = 8381 ns: non-significant *p<0.05, two-tailed one-way ANOVA; dashed lines represent the medians.

We then asked whether the increase in ORF1p expression levels observed in several brain areas in aged compared to young mice was also accompanied by a change in expression patterns. We therefore analyzed cell proportions and densities comparing young and aged mouse brains. Globally, we observed a reduction in the proportion of ORF1p+/NeuN+ cells in aged mouse brains using the cell detection workflow applied to large-scale images described in Figure 1A, a phenomenon mainly driven by the midbrain motor, the dorsal striatum, the pallidum and the thalamus regions (Suppl Fig. 5A, dark red bars, Suppl_Table1). The confocal approach applied to two regions, the frontal cortex and the ventral midbrain (Suppl Fig. 5B), confirmed this reduction in ORF1p+/NeuN+ cell proportions in favor of the ORF1-/NeuN- cell population in the ventral midbrain with no change in cell proportions in the frontal cortex in accordance with the large-scale imaging approach (Suppl Fig. 5A). The predominantly neuronal identity of ORF1p+ cells, however, was unchanged in the ventral midbrain (Suppl Fig. 5C) just as the proportion of neurons expressing ORF1p (Suppl Fig. 5D). We observed a significant shift in NeuN+/- cell proportions (Suppl Fig. 5E) which could either suggest a decrease in NeuN+ cells or a gain of NeuN- cells in this region with age. While proportions are less sensitive to technical variability and can identify cell population shifts, cell densities allow for absolute comparisons. When quantifying global cellular densities throughout the brain, we did not observe a significant reduction of ORF1p expressing cells (Suppl Fig. 5F), neurons (Suppl Fig. 5G) or non-neuronal cells (Suppl Fig. 5H). When analyzing cell densities in the nine brain regions separately, there were no significant changes in ORF1p positive (Suppl Fig. 5I), NeuN positive (Suppl Fig. 5J) or NeuN negative (Suppl Fig. 5K) cell densities in eight out of nine brain regions. The only exception was the dorsal striatum, but technical limitations applying to this particular brain region might account for these changes. Indeed, the dorsal striatum is different from the other brain regions as it represents the only region consisting of a single ABBA sub-region resulting in only one overall background measurement. Taken together, while there were no major changes in cell proportions, densities nor in ORF1p+ cell identities, we observed an age-dependent increase in ORF1p expression per cell of up to 27%.

Coding LINE-1 transcripts are increased in aged human dopaminergic neurons

Following the observation of increased ORF1p expression in the aged mouse brain, among which the ventral midbrain, and given the age-related susceptibility of dopaminergic neurons in the SNpc to cell death and to degeneration in PD 53, we turned to a previously published RNA-seq dataset of laser-captured micro-dissected post-mortem human dopaminergic neurons of brain-healthy individuals 54, in order to interrogate full-length LINE-1 mRNA expression profiles as a function of age. To avoid read-length bias to which TE analysis is particularly sensitive, we analyzed only the data derived from 50bp paired-end reads of linearly amplified total RNA as this dataset represented all age categories (n=41; with ages ranging from 38 to 97; mean age: 79.88 (SD ±12.07); n=6 ≤65y; n=35 >65y; mean PMI: 7.07 (SD ±7.84), mean RIN: 7.09 (±0.94), metadata available in Suppl_Table5). As age-related dysregulation of TEs might not be linear, we considered individuals with ages-at-death younger or equal to 65 years as “young” (n=6, 38-65 years, mean age 57.5 years (SD ±9.9)) and individuals older than 65 years as “aged” (n=35, 65-97 years, mean age 83 years (SD±7.8). The expression of the dopaminergic markers tyrosine hydroxylase (TH) and LMX1B were similar in both populations indicating no apparent change of dopaminergic identity of analyzed melanin-positive dopaminergic neurons (Suppl Fig. 6A, B). Next, we compared the expression of repeat elements at the class, family and name level based on the repeat masker annotation implemented in the UCSC genome browser using a commonly used mapping strategy for repeats consisting of randomly assigning multi-mapping reads 55. No overt dysregulation of repeat elements at either level of repeat element hierarchy was observed (Suppl Fig. 6C-F). There was a modest but significant increase in several younger LINE-1 elements including L1HS and L1PA2 at the “name” level (Fig. 4A, B), an analysis which was however underpowered (post-hoc power calculation; L1HS: 28.4%; L1PA2: 32.8%) and thus awaits further confirmation in independent studies. No expression changes were observed for HERVK-int, a human endogenous retrovirus family with some copies having retained coding potential (Fig. 4B) or other potentially active TEs like HERVH- int, HERV-Fc1, SVA-F or AluYa5 transcripts in the >65y group (Suppl Fig. 6G). Interestingly, L1HS expression was highly correlated with L1PA2 expression and this correlation extended to almost all younger LINE-1 subfamilies weaning down with evolutionary distance (Fig. 4C). This was not true for other active TEs as L1HS was negatively correlated with HERVK-int expression (Fig. 4C). Several regulators of LINE-1 activity have been identified 17, 56 and correlation of their expression with L1HS might allow to infer their relevance of interaction (activation or repression) with L1HS in human dopaminergic neurons. Spearman correlation analysis revealed three known repressors of LINE-1 activity whose expression was negatively correlated with LINE-1 expression; EN1 (Engrailed 1 17, Suppl Fig. 7A) with important functions for dopaminergic neuron homeostasis 57, CBX5/HP1a, a heterochromatin binding protein binding to the histone mark H3K9me3, thereby mediating epigenetic repression 58 (Suppl Fig. 7B) and XRCC5/6, also known as Ku86/Ku70, which are essential for DNA double-stranded break repair through the nonhomologous end joining (NHEJ) pathway and limit LINE-1 full-length insertions 59 (Suppl Fig. 7C). The transcripts of these genes showed, although not statistically significant, a trend for decreased expression in the elderly (Suppl Fig. 7D-G). Based on the increase of young LINE-1 families L1HS and L1PA2 in aged human dopaminergic neurons and the finding that ORF1p was increased in the aged mouse brain, we focused our attention on LINE-1 elements with coding potential for ORF1 and ORF2 according to the L1Basev2 annotation which are specific elements comprised in the L1HS and L1PA2 annotation at the “name” level. Most of the 146 full-length and coding LINE-1 termed UIDs (= Unique Identifier) in the L1Base are L1HS elements (76.03%), whereas the remaining 35 UIDs belong to the evolutionary older L1PA2 family (Suppl Fig. 7A). The L1Base annotation is based on the human reference genome (GRCh38) and annotates 146 human full-length (>6kB), intact LINE-1 elements (ORF1 and ORF2 intact) with a unique identifier from 1 to 146 4. Attribution of sequencing reads to a specific, individual TE copy is problematic 60 and several approaches have been proposed to circumvent this problem including the mapping of unique reads 55.While several tools using expectation maximization algorithms in assigning multi-mapping reads have been developed and successfully tested in simulations 55, 61, we used a different approach in mapping unique reads to the L1Base annotation of full-length LINE-1. Specific “hot” LINE-1 loci in a given cellular context have been identified 3, but usage of the L1Base annotation enabled an unbiased approach albeit ignoring polymorphic LINE-1 sequences. Unique read mapping strategies for repeat elements, especially young LINE-1 elements, will unavoidably underestimate LINE-1 locus-specific expression levels 55, but will be most accurate in assigning reads to a specific genomic location while allowing the comparison of two different conditions analyzed in parallel. Of the 146 full-length LINE-1 elements in the L1Base annotation, 111 were of the L1HS family and 35 belonged to the L1PA2 family (Suppl Fig. 8A). Assuming that expression of UIDs was correlated with mappability, we plotted a mappability count of each UID against its mean normalized read count expression of the six individuals ≤65y (Suppl Fig. 8B, C). Non-parametric Spearman correlation revealed no correlation between UID mappability and expression (Suppl Fig. 8C) indicating no apparent bias between the two parameters. However, individual UID dependency of mappability on expression cannot be excluded, especially for high expressing UIDs like UID-16 for example (Suppl Fig. 8B, C). Expression of LINE-1 at the locus-level has been attributed to artefacts not representing autonomous transcription including differential high intronic read counts 62, pervasive transcription or reads attributable to passive co-transcription with genes when the LINE-1 element is intronic 63. To evaluate the latter, we determined the number of intronic (46.58%) and intergenic UIDs (78/146; Suppl Fig. 8D) and identified the corresponding genes for intronic UIDs (Suppl Fig. 8E). Of the 146 UIDs, 140 passed the threshold of >3 reads in at least 6 individuals. Differential expression of UID between “young” and “aged” dopaminergic neurons revealed several significantly deregulated full-length LINE-1 loci (Fig. 5A). Paired analysis of the expression of all UIDs indicated a general increase (Fig. 5B), especially of low expressed UIDs. The comparative analysis of the sum expression of UIDs per individual comparing young (≤65y) with elderly human dopaminergic neurons, however, did not reach statistical significance (Fig. 5C). Several specific loci were dysregulated, in particular UID-68 (Fig. 5A and 5D), a L1HS element located on chromosome 7 (chr7: 141920659-141926712) in between two genes, OR9A4 (olfactory receptor family 9 subfamily A member 4) and CLEC5A (C-type lectin domain containing 5A; Fig. 5E and Suppl Fig.7E). This specific full-length LINE-1 element had a high mappability count of 16 (range of all UIDs: 1-30, mean 9.0 (SD ±6.05), Suppl Fig. 7C) and a post-hoc power analysis score of 96.6% (continuous endpoint, two independent samples, alpha 0.05). To rule out any influence of “hosting” gene transcription interference on measurable UID-68 expression differences, we performed Spearman correlation which did not indicate any correlation between OR9A4 (Fig. 5F) or CLEC5A (Fig. 5G) expression with UID-68. Together, this indicated that UID-68 might be a candidate for an age-dependent gain of activity. Other dysregulated UIDs (i.e. UID-129, UID-37, UID-127 and UID-137) had either a low mappability score, a low post-hoc power or did not pass the visualization check in IGV, reinforcing the notion that a combination of quality control criteria is crucial to retain a specific locus with confidence. In conclusion, TE expression analysis of this human dataset covering an age-span of 59 years (mean age difference between both groups 25.5 years) indicates an increase in the expression of young LINE-1 elements including those which have coding potential in elderly dopaminergic neurons, particularly a specific full-length LINE-1 element on chromosome 7 (UID-68). A slight net sum increase of UID transcripts/cell might be sufficient for the production of “above steady-state” levels of ORF1p and ORF2p. Other TEs with coding potential, namely members of the HERV family, were not increased. Further, correlation analyses suggest that L1HS expression might possibly be controlled by the homeoprotein EN1, a protein specifically expressed in dopaminergic neurons in the ventral midbrain 57, the heterochromatin binding protein HP1 and the DNA repair proteins XRCC5/6.

Young LINE-1 elements are increased in aged human dopaminergic neurons

TE transcript expression in RNA-seq data of laser-captured micro-dissected post-mortem human dopaminergic neurons of brain-healthy individuals was analyzed using RepeatMasker (multimappers) or the L1Base (unique reads). (A) Volcano plot of differential analysis of LINE-1 expression using DESeq2 comparing young (≤65y, n=6) or aged (>65y, n=35) human dopaminergic neurons at the “name” level of RepeatMasker. Young LINE-1 elements, including the two families L1HS and L1PA2 that have coding copies, are highlighted in red. (B) Scatter plots of normalized read counts (“name” level) of the young L1HS and L1PA2 families as well as the human endogenous virus family HERVK-int, another TE family with coding potential comparing young (≤65y, n=6) or aged (>65y, n=35) human dopaminergic neurons. Mann-Whitney test, p<0.05. (C) Correlation of the RNA expression levels of LINE-1 elements with known transposable element regulators in human dopaminergic neurons (all ages included). Spearman correlation of evolutionary close (L1HS, L1PA2-17) and distant LINE-1 (L1PB and L1MA5) as well as HERV elements with coding potential (HERV-Kint, HERV-Fc1, HERV-Fc2 and HERV-H-int) with known regulators of transposable elements for each individual sample, all ages included. HERV-W and TREX1 did not pass the normalized read count threshold of >3 reads in >6 individuals.

Dysregulation of locus-specific full-length LINE-1 elements in aged human dopaminergic neurons

(A) Volcano plot of differential expression analysis of TE expression using DEseq2 comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons at the locus-level of specific full-length LINE-1 elements (140 of 146 “UID’s” as annotated in L1Base; threshold >3 reads in at least 6 individuals). (B) Pairwise comparison of the expression of 140 out of 146 full-length LINE-1 elements comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons. Wilcoxon matched signed rank test, p<0.0001left panel). (C) The sum of read counts of all UIDs per individual were plotted comparing young (≤65y, n=6) and aged (>65y, n=35) human dopaminergic neurons. (D) UID-68 is dysregulated in aged human dopaminergic neurons. Normalized read counts of uniquely mapping reads mapping to the full-length LINE-1 element “UID-68” per individual were plotted comparing human post-mortem dopaminergic neurons from young (≤65y, n=6) and aged (>65y, n=35) individuals; two-tailed Mann-Whitney test (* p=0.046). (E) IGV window of the locus around the full-length LINE-1 UID-68 (chr7:141.918.055-141.932.948). UID-68 is located adjacent to the genes CLEC5A (right) and OR9A4 (left). Coverage and mappability of the locus including UID-68 is shown in tracks. Coverage is represented by average bigwig profiles (same scale, black: </=65y, red >65y); mappability of the genomic locus is depicted by Umap 50 tracks showing peaks overlapping with the peaks of reads (bigwig averages). (F) Spearman correlation analysis of the expression of UID-68 and OR9A4 in young (≤65y, n=6, black dots; Spearman r= 0.66, p=0.17)) or aged (>65y, n=35, red squares, Spearman r= 0.15, p=0.40) human dopaminergic neurons. (G) Spearman correlation analysis of the expression of UID-68 and CLEC5A in young (≤65y, n=6, black dots, Spearman r= 0.75, p=0.11) or aged (>65y, n=35, red squares; Spearman r= - 0.03, p=0.87) human dopaminergic neurons.

Endogenous ORF1p interactors in the mouse brain

Immunoprecipitation (IP) of endogenous ORF1p from the mouse brain. WB against ORF1p showing ORF1p enrichment after IP but no signal in the IgG control. Five independent samples were then prepared for proteomic analysis by mass spectrometry (LC-MS/MS). (B) GO slim enrichment analysis of proteins selected as endogenous ORF1p protein partners in the mouse brain after quantitative LC-MS/MS. ORF1p-immunoprecipitated proteins were categorized into GO slim terms. The nine GO slim term with the highest fold-change are plotted. Fold enrichment is depicted on the upper axis and displayed as bars, the FDR value appears on the lower axis and is represented by the black points. BP: Biological Process, CC: Cellular Component, MF: Molecular Function. (C) Venn diagram showing common interactors (purple) between interactors of endogenous ORF1p in the mouse brain identified in this study (red) and known (published) interactors of ORF1p (blue). Statistical significance of the overlap between the two groups of proteins was tested by an overrepresentation test (http://nemates.org/MA/progs/overlap_stats.html). (D) ORF1p associates with the SWI/SNF complex (red), RNA pol II complex (orange) and interactors belonging to GO terms related to neuronal cell body & neuron projection (green). Known interactors previously published 35, 67, 72–76 are indicated with a purple ring. STRING network of physical interactions where nodes represent proteins partners identified in (A) and edges thickness represents the strength of shared physical complexes. Only proteins sharing physical interactions were represented.

Endogenous ORF1p interactors in the mouse brain

To go further in our understanding of steady-state neuronal ORF1p expression across the mouse brain, we immunoprecipitated ORF1p and performed quantitative label-free LC-MS/MS to identify potential protein partners of ORF1p in the adult mouse brain. We successfully immunoprecipitated endogenous ORF1p from whole brain lysates (Fig. 6A), where we detect ORF1p exclusively in the 5 independent ORF1p-IP samples (and not at all in 5 independent IgG-IP control samples; Suppl_Table2) and identified a total of 424 potential protein interactors associated with ORF1p (Suppl_Table2; n=5 mice). Using Gene Ontology (GO) analysis, we identified several interacting proteins belonging to GO terms related to known functions of the ORF1p protein in RNA binding, preferentially 64 but not exclusively in cis 65, for instance RNA decapping and mRNA catabolic process, or related to the known presence of ORF1p in ribonucleoprotein particles 66, 67 (GO: cytoplasmic ribonucleotide granule) or the presence of ORF1p in p-bodies 65 as shown in Figure 6B and listed in Suppl_Table3. Other GO terms that emerged, to our knowledge not previously associated with ORF1p, were related to cGMP-mediated signaling (GO: cGMP-mediated signaling and 3’-5’phosphodiesterase activity: i.e. PDE4A, PDE4B, PDE4DIP) and the cytoskeleton (GO: microtubule depolymerization, cytoskeleton organization, microtubule and tubulin binding, cytoskeletal motor activity and protein binding). cGMP signaling is regulated by 3’-5’ phosphodiesterases (PDEs) which degrade 3’,5’-cyclic guanosine monophosphate (cGMP) and 3’,5’-cyclic adenosine monophosphate (cAMP), an activity essential for cell physiology for the integration of extra- and intracellular signals including neuronal excitability, synaptic transmission and neuroplasticity 68, 69. Further, several ORF1p interacting proteins were constituents of the mating-type switching/sucrose nonfermenting complex (SWI/SNF complex), i.e. ARID1A, ARID1B, SMARCA2, SMARCB1, SMARCC2), an ATP-dependent chromatin remodeler complex disrupting nucleosome/DNA contacts to facilitate DNA/chromatin accessibility by shifting, removing or exchanging nucleosomes along DNA 70, 71. Finally, we also observed proteins belonging to the GO term “neuronal cell body”, corroborating with the neuron-specific presence of ORF1p in the brain. A comparative analysis with previous mass spectrometry studies 35, 67, 72–76 aimed at identifying ORF1p interacting proteins unveiled significantly more common proteins than randomly expected (overrepresentation test; representation factor 2.6, p< 5.4e-08; Fig. 6C), including LARP1, STAU2, ATXN2, RALY, TARBP2 or DDX21 (for a full list see Suppl_Table4). The presence of a significant number of overlapping ORF1p interactors in different non-neuronal human cells (HEK 67, 72, 73, HeLa 74, human breast and ovarian tumors 76 and hESCs 75) and mouse brain cells (our study), suggest conserved key interactors between both species and between cell types, with a subset of these proteins regulating RNA degradation and translation potentially relevant for the LINE-1 lifecycle itself. However, differences in experimental conditions in between studies could also influence this overlap. ORF1p interactors found in mouse spermatocytes 35 were also present in our analysis including CNOT10, CNOT11, PRKRA and FXR2 among others (Suppl_Table4). To unravel the physical interactions between the identified interactors of endogenous ORF1p within the mouse brain, we used the STRING database (Search Tool for Recurring Instances of Neighboring Genes, https://string-db.org/). This analysis generated a network representation, where physical interactions are represented by edges (Fig. 6D). In analogy with the GO term analysis, ORF1p displayed interactions with various clusters, including well-known RNA decapping complexes directed against LINE-1 RNA, which also encompassed DCP2 and DCP1A which had not previously been identified as interacting with ORF1p 77. Furthermore, ORF1p exhibited interactions with the SWI/SNF complex (highlighted in red) as well as subunits of the RNA polymerase II complex suggesting a direct or indirect association with accessible chromatin, a hitherto unknown interaction of ORF1p with chromatin compartments within the nucleus. Notably, a multitude of novel interactors belonged to the “neuronal cell body” and “neuron projections” clusters, proposing potential neuron-specific partners of ORF1p such as GRM2/5, BAI1, EPHA4, KCNN2, GRIK2 and DMD among others. A last cluster, formed by NCOA5 (Nuclear Receptor Coactivator 5), NXF1 (Nuclear RNA Export Factor 1), RANBP2 and NU133 (both nucleoporins), might imply a role for these interactions in L1-RNA nuclear export and/or a mechanism for the LINE-1 RNP to gain access to the nucleus in post-mitotic neurons. Altogether, the identification of known and novel interactors of ORF1p in the mouse brain suggests roles of ORF1p in the LINE-1 life cycle (RNA binding and metabolism, RNP formation, nuclear access) but also suggests potential novel physiological roles of ORF1p in the brain related to cytoskeleton organization, cGMP signaling, neuron-specific functions (i.e. synaptic signaling, Suppl_Table3) and chromatin organization and/or transcription regulation.

Discussion

While LINE-1 derepression in aging has been extensively explored in peripheral tissues and various pathologies, including cancer, our understanding of LINE-1, particularly ORF1p, in the central nervous system remains limited 20, 41, 78. A recent search of ORF1p peptides in mass spectrometry data spanning 29 different healthy tissues did not reveal the presence of ORF1p in the brain, suggesting that its presence might lie below detection limits 20. Only a few studies explored and detected ORF1p expression in the brain, most focusing on a specific region (in mice 17, 33 in rats 79 and in human post-mortem brain 19), but it remained unclear if ORF1p is expressed throughout the entire brain, exhibits cell-type specificity, and most intriguingly, if its expression is influenced by the aging process. Here, using well characterized and validated antibodies, we demonstrate that ORF1p is expressed throughout the entire mouse brain and in at least three regions of the human post-mortem brain at steady-state. Leveraging a comprehensive workflow that incorporates brain atlas registration and machine learning algorithms, we quantified tens of thousands of brain cells, enabling a profound analysis of cell proportions, cell identities, densities and ORF1p expression levels across the entire brain. Surprisingly, more than one-fifth of detected cells expressed ORF1p. Regional variations in ORF1p expression levels were observed, with each region exhibiting distinct proportions, cell density, and signal intensity of ORF1p+ cells. In a non-neurologically diseased human brain, ORF1p was expressed in all three regions examined, that is the cingulate gyrus, the frontal cortex and the cerebellum. This is in accordance to an earlier study using histological staining, which found ORF1p expression in the human frontal cortex, the hippocampus, in basal ganglia, thalamus, midbrain and the spinal cord 19. This suggests, similarly to the mouse brain, a generalized expression across the human brain. On the transcriptomic level using long-read sequencing of GTEx tissues, brain and liver were highlighted as the organs displaying the highest expression of putatively active, full-length LINE-1 elements 80. However, when the authors looked at sub-regions, they found transcript expression in cerebellar hemispheres and the putamen, but not in the caudate and the anterior cingulate gyrus and frontal cortex 80. This is in contrast to our data and the data from Sur et al, where ORF1p was found to be expressed in the latter two regions using two different antibodies. We used the anti-human LINE-1 ORF1p antibody clone 4H1, a well characterized antibody 78, 81. While the sample size for the staining of human post-mortem tissues certainly needs to be increased in order to draw quantitative conclusions, the presence of the protein in two independent studies does point to a steady-state expression of ORF1p in the human brain.

It is interesting to note that ORF1p is expressed at steady-state in both, the mouse and human brain, despite the fact that evolutionary young and thus potentially ORF1-encoding LINE-1 elements in mice 82 (L1A, Tf/Gf) and in humans 4 (L1HS, L1PA2) differ significantly in number, sequence and regulation 83. Most differences lie in the 5’promoter region, but also ORF1 and ORF2 sequences are strikingly divergent between mouse and humans. For example, mouse LINE-1 promoters are composed of a varying number of monomers, a structure not found in human-specific LINE-1 promoters 84. This has obvious implications for LINE-1 expression regulation which can be very different, but examples of co-evolution of regulatory networks have been described 85 and might operate in the brain to regulate LINE-1 and thereby ORF1p expression.

In the mouse brain, we find ORF1p to be expressed predominantly if not exclusively in neurons using immunofluorescence and fluorescence-activated cell sorting (FACS) followed by Western blotting. This result is consistent with previous studies, such as the identification of ORF1p in excitatory neurons within the mouse frontal cortex 86, in parvalbumin neurons in the hippocampus 18, its presence in neurons in the ventral midbrain including in dopaminergic neurons 17 and the recognition of morphological similarities between stained neurons and ORF1p+ cells in a post-mortem hippocampus sample of a healthy individual 19. We also detected ORF1p in Purkinje cells in the mouse and human cerebellum. Neuronal specificity or preference of LINE-1 expression was also shown on the transcriptomic level in recent studies investigating LINE-1 expression in the mouse hippocampus, where neuronal LINE-1 expression exceeded that of astrocytes and microglia by approximately twofold 20, is abundant in parvalbumin interneurons 18 and single-nuclei RNA-seq data from the mouse hippocampus and frontal cortex which confirmed globally that repetitive elements including LINE-1 are more active in neurons than in glial cells 86. In the human brain, LINE-1 transcripts were found in greater quantities in neurons compared to non-neuronal cells by single-nucleus sequencing87. Furthermore, retrotransposition-competent LINE-1 elements were found expressed exclusively in neurons 88. While ORF1p expression is suggested to be expressed in microglia under experimental autoimmune encephalomyelitis conditions in the spinal cord 89, no evidence of such expression was observed in non-neuronal cells under non-pathological condition.

On average, throughout the mouse brain, the majority of neurons were positive for ORF1p and in some regions (i.e. the thalamus) around 80% of neurons expressed ORF1p. Comparing the results of both imaging approaches, the percentages of neurons expressing ORF1p in the ventral midbrain and frontal cortex were roughly similar (around 70% of neurons expressed ORF1p as quantified by confocal imaging and about 60% of neurons were identified as ORF1p+ using the slide scanner approach). In the human cingulate gyrus, we found that 37.2% of neurons express ORF1p and that 80% of cells expressing ORF1p were neurons, which are proportions similar to some regions of the mouse brain. It is however possible that these percentages are underestimated due to technical issues inherent to the machine-learning based algorithm for cell detection as our observations often indicated a positive signal in neurons which were classified as negative due to a particular shape or our stringent intensity threshold. A question which arises based on these findings is whether specific features distinguish ORF1p+ and ORF1p-neurons. One hint comes from a recent study suggesting that in the mouse hippocampus, it is the parvalbumin positive neurons that predominantly express ORF1p 18. However, as we show here, while PV-positive neurons often co-stain with ORF1p, not all ORF1p positive cells are PV-positive. In the mouse ventral midbrain, TH-positive dopaminergic neurons express higher levels of ORF1p compared to surrounding, non-dopaminergic neurons 17 (and this study Fig. 1B, panel 8). In the mouse cerebellum, ORF1p staining was detected in Purkinje cells and in the human post-mortem brain in Purkinje and possibly in Basket cells. Parvalbumin positive neurons are inhibitory neurons, so are Purkinje and Basket cells. However, dopaminergic neurons are modulatory neurons exerting excitatory and inhibitory effects depending on the brain region they act on. Specific neurons in the granular layer (i.e. Golgi and unipolar brush cells) of the cerebellum are inhibitory, but ORF1p negative, indicating that the decisive feature might not be the excitatory or inhibitory nature of a neuron. Further, ORF1p is expressed in excitatory (CamKIIa-positive) and CamKIIa-negative neurons in the mouse frontal cortex 86 and there is evidence of full-length L1 RNA expression in both excitatory and inhibitory neurons 87. While further studies are necessary to define the neuronal subtypes expressing ORF1p and their epigenetic make-up allowing this expression, it seems reasonable to conclude on the above-mentioned data that there is no neuronal sub-type specificity characterizing ORF1p expressing neurons. Another possibility is a cell-type specific chromatin organization permissive for the expression of LINE-1 and future single-cell studies in the mouse and human brain might reveal those differences.

Because transposable elements are known to become active in somatic tissues during aging 15, 16, 28, 90, 91, we aimed to investigate whether there was a corresponding increase at the protein level. In aged mice, ORF1p expression significantly increased throughout the mouse brain consistent with a previously documented increase in ORF1p outside the central nervous system in aged rats 79, 92 and aged mice 90 and in neurons of layer 2/3 of the mouse frontal cortex86. By quantifying the mean intensity of ORF1p in over 70 000 cells identified as ORF1p+, we were able to characterize the extent of this increase in each anatomical sub-region. Remarkably, apart from the isocortex which did not show any change, ORF1p expression increased in all other brain regions by 7% to 27%, indicating a generalized increase of ORF1p expression in neurons throughout the brain (13%). We did not detect any change in cell identity of ORF1p expressing cells, that is, ORF1p expression remained predominantly if not exclusively neuronal. Globally, there was also no change in ORF1p-positive, neuronal or non-neuronal cell densities in the aged mouse brain. Further investigations are necessary to investigate the underlying mechanism of the loss of ORF1p+ cells in the dorsal striatum in the aged mouse brain and to examine a possible relationship to the change of proportions of cells in the ventral midbrain, a structure which contains the SNpc which projects to the dorsal striatum and which is prone to LINE-1 driven neuronal degeneration 17. Thus, while ORF1p intensities per cell increase significantly in older mice in several brain regions, here is no global change in ORF1p+ cell numbers.

An increase of ORF1p might have several direct or indirect consequences on a cell or here, on a neuron. As ORF1p is translated from a polycistronic LINE-1 RNA together with ORF2p, albeit in much higher amounts (the estimated ratio ORF1p to ORF2p is 240:1 in non-native conditions) 93, it can be expected that a LINE-1 ribonucleotide particles are formed and ORF2-dependent cell toxicity in form of genomic instability 17, 21 and single-stranded cytoplasmic DNA triggered inflammation 27, 28, 90 might result. This has been shown in mouse dopaminergic neurons where oxidative stress induced LINE-1 causally contributed to neurodegeneration 17.

Neurodegeneration was partially prevented by anti-LINE-1 strategies among which NRTIs 17 and similar LINE-1 protein-dependent neuronal toxicity has been shown in drosophila31, 32 and the mouse cerebellum 33.

In order to test whether an increase in LINE-1 is a feature of human brain aging, we turned to a unique RNA-seq dataset of human laser-captured dopaminergic neurons of 41 individuals ranging from 38 to 99 years 54. In accordance with our focus on LINE-1 sequences which are full-length and coding, we developed a rationale to interrogate LINE-1 families with representatives that are coding (L1HS, L1PA2, multimappers; RepeatMasker) and to specifically investigate full-length LINE-1 elements that have intact open reading frames for ORF1p and ORF2p (unique reads; L1Basev2 4). Indeed, we find an increase in L1HS and L1PA2 elements in individuals ≥65y as well as an increase in specific full-length LINE-1 elements but only a trend for increase of all full-length LINE-1 in sum in the elderly. This analysis has technical limitations inherent to transcriptomic analysis of repeat elements especially as it is based on short-read sequences and on a limited and disequilibrated number of individuals in both groups. Nevertheless, we tried to rule out several biases by demonstrating that mappability did not correlate with expression overall and used a combination of visualization, post-hoc power analysis and analysis of the mappability profile of each differentially expressed full-length LINE-1 locus. Interestingly, dysregulated full-length LINE-1 elements in aged dopaminergic neurons did not correspond to those identified in bladder cancer 94 indicating the intricate nature of this expression across tissues and pathological conditions. Overall, a slight net sum increase of UID transcripts/cell might be sufficient for the production of “above steady-state” levels ORF1p and ORF2p. Further, a dissociation of LINE-1 transcript and protein levels in aging has been observed recently in excitatory neurons of the mouse cortex. In the absence of transcriptional changes of LINE-1, protein levels of ORF1p were increased 86.

We can only speculate about the reason for an increase in ORF1p in the aged brain. A recent single-cell epigenome analysis of the mouse brain suggested a specific decay of heterochromatin in excitatory neurons of the mouse brain with age which was paralleled by an increase in ORF1p, albeit equally in excitatory and inhibitory neurons, again not indicating any dependency of ORF1p regulation on the excitatory or inhibitory nature of neurons 86. Chromatin and particularly heterochromatin disorganization are a primary hallmark of aging 91 but other repressive cellular pathways which control the LINE-1 life cycle might also fail with aging 13. Another possibility is a loss of accessibility of repressive factors to the LINE-1 promoter or an age-dependent decrease in their expression. Matrix correlation analysis of several known LINE-1 regulators, both positive and negative, revealed possible regulators of young LINE-1 sequences in human dopaminergic neurons. Despite known and most probable cell-type unspecific regulatory factors like the heterochromatin binding protein CBX5/HP1 58 or the DNA repair proteins XRCC5 and XRCC6 56, we identified the homeoprotein EN1 as negatively correlated with young LINE-1 elements including L1HS and L1PA2. EN1 is an essential protein for mouse dopaminergic neuronal survival 57 and binds, in its properties as a transcription factor, to the promoter of LINE-1 in mouse dopaminergic neurons 17. As EN1 is specifically expressed in dopaminergic neurons in the ventral midbrain, our findings suggest that EN1 controls LINE-1 expression in human dopaminergic neurons as well and serves as an example for a neuronal sub-type specific regulation of LINE-1. Although these proteins are known regulators of LINE-1, this correlative relationship awaits experimental validation. The heterogenous, brain-wide presence of ORF1p expression at steady-state is intriguing. In cancer cell lines or mouse spermatocytes, ORF1p interacts with several “host” proteins, some if not most of which are related to the LINE-1 life cycle. However, a profile of endogenous ORF1p interactors in the mouse brain might inform on possible other and organ-specific functions besides its binding to the LINE-1 RNA in “cis”35. Among the total 424 potential interactors of endogenous ORF1p in the mouse brain, 38 partners had been previously identified by mass spectrometry in human cancers, cancerous cell lines and mouse spermatocytes 35, 67, 72–76 (Suppl_Table4). This supports the validity of the list of ORF1p partners identified, although we cannot rule out the possibility that unspecific protein partners might be pulled down due to colocalization in the same subcellular compartment. GO term analysis contained expected categories like “P-body”, mRNA metabolism related categories and “ribonucleoprotein granule”. We also identified NXF1 as a protein partner of ORF1p, a protein found to interact with LINE-1 RNA related to its nuclear export 95. This suggests the conservation of key interactors probably essential for completing or repressing the LINE-1 life cycle in both species, despite the divergence of mouse and human ORF1p protein sequences 96. Along these lines, several ORF1p protein partners we identified might complete the list of post-transcriptional regulators implicated in LINE-1 silencing. Recent work conducted on human cancerous cell lines has demonstrated that MOV10 orchestrates the recruitment of DCP2 for LINE-1 RNA decapping 77. In our analysis, we identified DCP2 along with DCP1A, known to enhance the decapping activity of DCP2 97, and DCP1b, a pivotal component of the mRNA decapping complex 98. Intriguingly, MOV10 was not detected in our mass spectrometry analysis, despite its established role in recruiting DCP2 and forming a complex with L1-RNP to mediate LINE-1 RNA decapping, as reported by Liu et al 77. However, we found two enhancers of mRNA decapping, EDC3 and EDC4, both core components of P-bodies, a membrane-less organelle known to contain L1-RNP 65. Multiple ubiquitin-ligase proteins were found although not appearing as a significantly enriched GO term. These results add to the picture of the post-transcriptional and translational control of ORF1p and suggest that these mechanisms, despite a steady-state expression, are operational in neurons. Further, several neuron-specific interactors were identified belonging to GO term categories “neuron projection” (75 proteins) and “neuronal cell body” (5 proteins), again pointing to the neuron-predominant expression of ORF1p in the mouse brain. Other interesting aspects were raised from this analysis. Among significantly enriched GO terms, several were related to the cytoskeleton, the functional consequences of which need to be determined in future studies. Our screen also identified PDE10A as an interactor of ORF1p in the mouse brain, a PDE almost exclusively expressed in medium spiny neurons of the striatum and a target for treatment of neurological diseases related to basal ganglia function like Huntington’s disease, schizophrenia and Tourette syndrome 99. Interestingly, PDE10A inhibition is related to beta-catenin signaling, another GO term which emerged from our screen 100. Finally, we found components of RNA polymerase II and the SWI/SNF complex as partners of ORF1p. This might further indicate that ORF1p has access to the nucleus in mouse brain neurons as described for other cells 101, 102, however a bias due to a post-lysis effect cannot be excluded. These findings give rise to intriguing questions regarding the potential function of ORF1p in neuron in health and disease as (i) ORF1p is widely distributed throughout the brain under normal physiological conditions, (ii) ORF1p shows a wide range expression levels within and in between regions, (iii) ORF1p is expressed predominantly if not exclusively in neurons, (iv) but not in all neurons and (v) interacts with proteins that might not directly relate to the LINE-1 life cycle, some of which are neuron-specific. In addition, physicochemical properties of ORF1p to form compacted nucleic acid - bound complexes with sequestration potential were shown 96, 103. Future loss-of-function studies should help to shed light on the necessity of ORF1p for neuronal functions if they exist. This data spurs the idea of a possible “physiological” function of ORF1p as an integrative protein with exapted function in neuronal homeostasis and a loss of restriction in the aged brain limiting LINE-1 expression to steady-state levels.

Materials & methods

Animals

Swiss OF1 wild-type mice (Janvier) were housed on a 12h light/dark cycle with free access to water and food. Mice were sacrificed at 3-month or 16-month. Animal experiments were performed according to the EU directive 2010/63/EU.

Mouse tissue dissection and protein extraction

Tissues were extracted from 3-month-old and 16-month-old Swiss/OF1 mice. Briefly, the two hemispheres were separated in ice cold PBS -/-. For each mouse, one hemisphere was rinsed and fixed in 4% PFA for 1h followed by 24h of incubation in 30% sucrose. Hemispheres were kept at -20°C until being sliced on a freezing microtome (Epredia, HM 450) with a 20µm thickness. The other hemisphere was dissected in ice cold PBS -/- 1X and 6 brain regions were rinsed, cut in small pieces and dissociated separately using a large (21G) to small gauge (27G) needle in RIPA lysis buffer for 5 min. Lysates were kept on ice for 25 min, were sonicated for 15 min and supernatants were collected after a 30 min centrifugation at 4°C at 14 000 rpm. Proteins were quantified and Laemmli buffer was added before boiling for 10min at 95°C to be used for Western Blot.

Human Samples

Cerebellum, frontal cortex and cingulate gyrus human samples were provided by Neuro-CEB biobank from a 78-year-old brain-healthy male and conserved at -80°C.

Human Samples pulverization and protein extraction

We used the dry pulverizer Cryoprep (Covaris) for pulverization of tissue blocs. Each sample was disposed in a liquid-nitrogen precooled Tissue-tube bag and dry cryo-pulverized with one impact at the maximum level. The pulverized brain sample was then weighed and resuspended in lysis buffer (mg/v) (0.32M sucrose, 5mM CaCl2, 3mM Mg(CH3COOH)2, 0.1mM EDTA, 10mM Tris-HCL pH8, 1mM DTT, 0.1% TritonX-100 and Protease Inhibitors), kept on ice for 30 min with gentle up-and-down pipetting until homogenization. We added 2X RIPA buffer (v/v) to totals fractions for 30 min on ice. We then sonicated samples 2 times for 15 min. AtlasSupernatants were collected after a 30 min centrifugation at 14 000 rpm at 4°C, proteins were quantified and Laemmli buffer was added to be used for Western Blot. All samples were boiled 10 min at 95°C to be used for Western Blot.

Cell culture and siRNA delivery

LUHMES were purchased from ATCC (CRL2927). Cells were cultured on 50 μg/mL Poly-L-ornithine (Merck) and 1 μg/mL human plasma fibronectin (Sigma) coated flasks and cultured in Advanced DMEM/F12 (Gibco) added with 1% N-2 supplement (Gibco), GlutaMax (Gibco) and 40 ng/mL human recombinant FGF (Peprotech) at in 5 % CO2, 37C° incubator. Differentiation was initiated by adding to the media 1 µg/ml doxycycline (Sigma), 2 ng/ml recombinant human GDNF (Peprotech) and 1 mM cAMP (Sigma). Media was changed every two days. Experiences were performed on day 7 of differentiation. Cells were passaged less than 12 times

MN9D were purchased from Merck (SCC281). Cells were cultured on 1 mg/mL poly-lysine (Merck) coated flasks and cultured in DMEM High Glucose (Sigma, D5796) with 10% FBS (Gibco) in 5 % CO2, 37C° incubator. Over a period of 10 days, cells were differentiated by adding to the media 1 mM n-butyrate (Sigma) and 1 mM dibutyryl cAMP (Sigma). Media was changed every two days. Experiences were performed on day 10 of differentiation. Cells were passaged less than 15 times.

LUHMES were lipofected using RNAiMAX (Invitrogen) at day 3 of differentiation with 100 nM of siRNA. MN9D were lipofected using RNAiMAX (Invitrogen) at day 7 of differentiation with 100 nM of siRNA. The siRNA sequences used were as follows:

siRNA control: TAATGTATTGGAACGCATA

siRNA ORF1 (LUHMES): AAGAAGGCTTCAGACGATCAA

siRNA ORF1 (MN9D): CTATTACTCTGATACCTAAAC

LUHMES were scrapped at day 7 of differentiation and MN9D at 10 days of differentiation, in RIPA buffer (10mM Tris-HCl, pH 8.0; 150mM NaCl; 1mM EDTA; 1% Triton X-100; 0.1% Sodium Deoxycholate; 0.1% SDS). Laemmli buffer was added and samples were boiled 10 min at 95C° before being loaded on a gel.

Western Blot

We used 1.5mm NuPAGE 4-12% Bis-Tris-Gel (Invitrogen™). Proteins samples (sorted mouse brain cells: 10000 cells/ µl -> 5 µl loaded; human brain lysates: 10µg; mouse brain lysates: 20µg) were loaded and gel migration was performed with NuPAGE™ MES SDS Running Buffer (Invitrogen™) for 45 minutes at 200mV. Gels were transferred onto a methanol activated PVDF membrane (Immobilon) in a buffer containing: Tris 25 mM, pH=8,3 and Glycine 192 mM, during 1h30 at 400 mA. Membranes were blocked 30 min with 5% milk in TBST (0,2% Tween 20, 150 mM NaCl, 10 mM Tris pH:8). The primary antibodies (mouse ORF1p antibody: abcam ab216324; human ORF1p antibody: abcam ab245249) were diluted in 5% milk in TBST, and membranes were incubated o/n at 4 C°. After 3 x 10 min washing in TBST, membranes were incubated for 1h30 with the respective secondary antibodies diluted at a concentration of 1/2000 in 5% milk TBST. Membranes were washed 3 x 10 min in TBST and were revealed by the LAS-4000 Fujifilm system using Clarity Western ECL Substrate (Bio Rad) or Maxi Clarity Western ECL Substrate (Bio Rad).

Immunostaining

Sagittal mouse brains slices were fixed for 10min in PFA 4% and rinsed 3 times for 10min in PBS -/-. Slices were then incubated 20 min in glycine 100mM, washed 3 times for 5min in PBS and immersed in 10mM citrate pH 6 at 62°C during 45min for antigen retrieval. Slices were then immersed 3 times in PBS with Triton X-100 0.2% and incubated in blocking buffer for 1,5 h (PBS with Triton X-100 0.2% and FBS (10%) previously inactivated 20min at 56°C (Gibco, 16141061). Primary antibodies (ORF1p antibody: abcam ab216324; NeuN antibody: GeneTex GTX00837) were diluted (1/200 and 1/500 respectively) in blocking buffer and incubated with slices overnight at 4°C and then washed 3 times for 10 min with PBS. For validation, an in-house ORF1p antibody was used (09) (guinea pig, 1/200). For non-neuronal marker (GFAP antibody: Millipore AB5541; Iba1 antibody: GeneTex GTX101495; Sox9 antibody: RnDsystems AF3075; S100β antibody: Sigma S2532), antibodies were diluted at 1/500. Additionally, WFA (Sigma L1516) and PV antibodies (Swant PV235) were used, diluted at 1/500 and 1/1000 respectively. Suitable secondary antibodies (Invitrogen) and Hoechst (Invitrogen, 15586276) were incubated for 1,5h at 1/2000 in PBS with inactivated FBS (10%) and washed 3 times 10 min in PBS. To quench tissue autofluorescence, especially lipofuscin, TrueBlack Plus (Biotium) in PBS was used during 10min. Slices were rinsed 3 times in PBS and mounted with Fluoromount (Invitrogen™).

For human cingulate gyrus stainings, the same protocol was performed, with the difference that a human ORF1p antibody (Abcam 245249) was used. Mouse and human brain slices were imaged by the Axioscan 7 Digital Slide Scanner (Zeiss) or a Spinning Disk W1 confocal microscope (Yogogawa).

Blocking peptide

The ORF1p antibody (abcam ab216324) was incubated 2h on a turning wheel with excess (4:1) of mouse ORF1p recombinant protein as in 17 before the blocked antibody was used in the above-described immunofluorescence protocol.

Quantification of confocal acquisitions

Analysis was conducted using a custom-written plugin developed for the Fiji software, incorporating Bio-Formats 104 and 3D ImageJ Suite 105 libraries. Code is freely available online at https://github.com/orion-cirb/DAPI_NEUN_ORF1P. Nuclei were detected in the Hoechst channel downscaled by a factor of 2 with the 2D-stitched version of Cellpose 106 (percentile normalization = [1-99], model = ‘cyto’, diameter = 30, flow threshold = 0.4, cell probability threshold = 0.0, stitching threshold = 0.75). The segmented image was then rescaled to its original size, and the obtained 3D nuclei were filtered by volume to reduce false positive detections. NeuN+ and ORF1p+ cells were detected in their respective channel using the same approach as for nuclei detection, but with adjusted Cellpose settings (model = ‘cyto2’, diameter = 40, flow threshold = 0.4, cell probability threshold = 0.0, stitching threshold = 0.75). Finally, each cell was associated with a nucleus having at least half of its volume in contact with. Cells without any associated nucleus were discarded. Each nucleus was thus labeled according to NeuN and/or ORF1p positivity.

Nuclei and cell detection using the respective Cellpose models and hyperparameters were evaluated on eight images per channel, capturing intensity variability across different mouse ages and brain regions. A total of approximately 2,000 nuclei and 1,000 NeuN and ORF1p cells were manually annotated. We evaluated model performance with the average precision (AP) metric, computed from the number of true positives (TP), false positives (FP), and false negatives (FN) as at the commonly used Intersection over Union (IoU) threshold of 0.5. The AP at an IoU threshold of 0.5 was 0.995 for nuclei, 0.960 for NeuN, and 0.974 for ORF1p cells. These results confirmed that the selected models and hyperparameters were well-suited for processing the entire dataset.

ABBA Registration and Qupath analysis

Each sagittal brain section was registered with the Allen Mouse Brain Atlas (CCFv3 107) using the Aligning Big Brains & Atlases plugin 34, 108 in Fiji. Slices were first manually positioned and oriented along the Z-axis. Automated affine registration was then applied in the XY plane, followed by manual refinement. The final registration results were imported into QuPath software 109 for downstream processing. Cell analysis in each brain subregion was performed with custom Groovy scripts developed for QuPath. Code is freely available online at https://github.com/orion-cirb/QuPath_ORF1P. Hoechst nuclei were detected with StarDist 2D 110, applying the DSB 2018 pretrained model with the following parameters: percentile normalization = [1-99], probability threshold = 0.82, overlap threshold = 0.25. Cells in NeuN and ORF1p channels were detected with Cellpose 2D (percentile normalization = [1-99], model = ‘cyto2’, diameter = 30, flow threshold = 0.4, cell probability threshold = 0.0). Nuclei and cells were then filtered by area and intensity to minimize false positive detections. Minimum intensity threshold was based on the channel background noise, which was estimated for each subregion as the mean intensity of pixels not belonging to any detected nucleus or cell in the respective channel. Finally, each cell was associated with a nucleus having its centroid located inside the cell mask. Cells without an assigned nucleus were discarded, cells with associated nuclei were classified as NeuN+ or NeuN- and ORF1p+ or ORF1p-. Intensity values were normalized by subtracting the background noise computed in the corresponding channel and subregion. As a last step, subregional results were merged into regional ones and data were analyzed using the Pandas Python library 111.

Nuclei and cell detection were evaluated on 14 images per channel, corresponding to approximately 800 nuclei and 400 NeuN and ORF1p cells manually annotated. The average precision (AP) at an IoU threshold of 0.5 was lower than for confocal images: 0.806 for nuclei, 0.675 for NeuN, and 0.695 for ORF1p cells. This decline in performance was primarily due to a lower signal-to-noise ratio in slide scanner images, leading to an increased number of false positives and false negatives. While fine-tuning the models could enhance detection robustness, the selected models and hyperparameters were considered suitable for processing the entire dataset.

FACS

Mouse brains were dissociated with Adult Brain Dissociation kit (Miltenyi Biotec, 130-107-677) and incubated with the coupled antibody NeuN Alexa 647 (Abcam, ab190565) or the control isotype IgG Alexa 647 (Abcam, ab199093). Stained cells were filtered a last time with a 40µm filter before FACS sorting (FACS ARIA II). Neuronal and non-neuronal cells were separately collected in PBS -/- 2m EDTA and then centrifugate (5min at 700rpm). Pellets were resuspended in RIPA for protein extraction in an appropriate volume in order to achieve equal cell concentrations (10 000 cells/µl).

RNA-seq analysis

The RNA-seq dataset from Dong et al 54 was downloaded from dbGAP (phs001556.v1.p1) and contains unstranded paired-end 50bp and 75bp reads from pooled laser-capture micro-dissected dopaminergic neurons from human post-mortem brain (107 samples) from 93 individuals w/o brain disease. RNA-seq had been done on total and linearly amplified RNA. We focused our analysis on data obtained with 50bp reads, in order to avoid mappability bias, while still regrouping all age categories (n=41; with ages ranging from 38 to 97 (mean age: 79.88 (SD ±12.07); n=6 ≤65y; n=35 >65y; mean PMI: 7.07 (SD ±7.84), mean RIN: 7.09 (±0.94)). Sequencing reads were aligned on the Human reference genome (hg38) using the STAR mapper (v2.7.0a) 3 and two different sets of parameters. Genome-wide individual repeat quantification was performed using uniquely mapped reads and the following STAR parameters: --outFilter mapNmax 1–-alignEndsType EndToEnd–-outFilterMismatchNmax 999–-outFilterMismatchNoverLmax 0.06. Repeats class, family and name quantification was performed using a random mapping procedure and the following parameters : -- outFilterMultimapNmax 5000–-outSAMmultNmax 1–-alignEndsType EndToEnd–-outFilterMismatchNmax 999–-outFilterMismatchNoverLmax 0.06–-outSAMprimaryFlag OneBestScore–-outMultimapperOrder Random. Repeats annotations were downloaded from the UCSC Table Browser (repeatMasker database: https://genome.ucsc.edu/cgi-bin/hgTables) and coordinates of LINE-1 full length and coding elements were downloaded from the L1base database 2 (http://l1base.charite.de/l1base.php; 4) selecting LINE-1 full length elements containing two predicted complete open reading frames for ORF1 and ORF2 (UID= Unique IDentifier) from the LINE-1 database (http://l1base.charite.de/l1base.php) and corrected genomic intervals with the repeat masker annotation of the corresponding genomic locus. Repeat quantification from the aligned data was done using a gtf file composed of all genes (Gencode v29) and all individual repeat elements. This strategy was used to avoid overestimation of repeat elements due to overlaps with expressed genes. For individual repeat quantification of the full length L1 elements (L1base), we therefore used a gtf of all genes and all L1base entries, and ran the FeatureCounts tool 112 with the following parameters: -g gene_id -s 0 -p. In the context of the family-based analysis, we used a gtf with all genes and all annotated repeats elements and ran FeatureCounts with -g gene_family -s 0 -p -M. Before DeSeq2 analysis, we remove all genes and repeat elements with less than 10 reads in a minimum of n individuals, n being the number of individuals in the condition containing the fewest individuals (“young” condition: n=6, 38-65y, mean 57.5 years). We use the same conditions with genes and UIDs with less than 3 reads in a minimum of n individuals. Finally, we calculated the scaling factors using DeSeq2 on all genes + all repeat elements or all gene + UID according to the quantification method and then applied these scaling factors to the corresponding counts tables.

In order to test for the mappability of each UID (= full-length and coding LINE-1), we extracted the bed track « main on human:umap50 (genome hg38) from the UCSC genome browser (≈ 7Mio regions) directly into Galaxy (usegalaxy.org) and joined genomic intervals with a minimum overlap of 45bp of this dataset with a dataset containing the annotation of UIDs extracted from L1Basev2 4 corrected in length with repeat masker and completed with information on whether the UID is intra- or intergenic and, if intragenic, in which gene (NM_ID, chr, strand, start, end, gene length, number of exons, gene symbol) the fl-LINE-1 is located, which resulted in 1266 regions. We then used the “group on data and group by” function in Galaxy and counted the number of overlapping 50kmers with all 146 UIDs (=mappability score). Correlation analysis (non-parametric Spearman) was then done between the mappability score and the normalized read counts.

For visualization of expression, bigwig files were generated for each age group, that is </=65y and >65y, respectively. We used bamcoverage to obtain bigwig files (normalized by cpm). Then for each age group an average bigwig was generated using bigwigAverage from deeptools (galaxy version 3.5.4); Bigwigs were loaded into IGV alongside tracks showing mappability (Umap50) and selected tracks from repeat masker (LINE, SINE and LTR).

Post-hoc power analysis was performed using the “Post-hoc Power Calculator” (https://clincalc.com/stats/Power.aspx).

Immpunoprecipitation of ORF1p from the mouse brain

For immunoprecipitation, we used ORF1p (abcam, ab245122) and IgG rabbit (abcam, ab172730) antibodies. The antibodies were coupled to magnetic beads using the Dynabeads® Antibody Coupling Kit (Invitogen, 14311D) according to the manufacture’s recommendations. We used 5µg of antibody for 1 mg of beads and used 1.5mg of beads for IP. Individual mouse brain lysates (n=5), homogenized using dounce and sonicated, were incubated with ORF1p or IgG-control coupled beads and a small fraction was kept as input. Each of these two tubes containing coupled beads and brain lysates were diluted in 5 ml buffer (10 mM Tris HCl, 150 mM NaCl, protease inhibitor). The samples were then incubated overnight on a wheel at 4°C. Samples were then washed 3 times with 1 ml buffer (10 mM Tris HCl pH 8, 200 mM NaCl) using a magnet and then resuspended in the same buffer. The samples were boiled in Laemmli buffer (95°C, 10 min) and 20 µl of each sample were loaded on a 4-12% Nupage gel (Invitrogen, NP0336) to be revealed by WB. For samples used in Mass Spectrometry study, beads were washed with buffer (10 mM Tris HCl pH 8, 200 mM NaCl) using a magnet. After 3 washes with 1ml buffer the beads were washed twice with 100 µL of 25 mM NH4HCO3 (ABC buffer). Finally, beads were resuspended in 100 μl of 25mM ABC buffer and digested by adding 0.20 μg of trypsine/LysC (Promega) for 1 hour at 37 °C. A second round of digestion was applied simultaneously on the beads by adding 100 µL of 25 mM ABC buffer and to the previous digest by adding 0.20 µg of trypsin/LysC for 1 hour at 37 °C. Samples were then loaded into homemade C18 StageTips packed by stacking three AttractSPE® disk (#SPE-Disks-Bio-C18-100.47.20 Affinisep) into a 200 µL micropipette tip for desalting. Peptides were eluted using a ratio of 40:60 CH3CN:H2O + 0.1% formic acid and vacuum concentrated to dryness with a SpeedVac device. Peptides were reconstituted in 10 µL of injection buffer in 0.3% trifluoroacetic acid (TFA) before liquid chromatography-tandem mass spectrometry (LC-MS/MS) analysis.

Immunoprecipitation of ORF1p from LUHMES cells (human) differentiated into mature dopaminergic neurons

For immunoprecipitation, we used ORF1p (Millipore, MABC 1152) and IgG mouse (Thermofisher, #31903) antibodies. The antibodies were coupled to magnetic beads using the Dynabeads® Antibody Coupling Kit (Invitogen, 14311D) according to the manufacturer’s recommendations. We used 8 µg of antibody for 1 mg of beads. The appropriate volume of buffer was added to the coupled beads to achieve a final concentration of 10 mg/ml. Cells were washed with 1X PBS and harvested using 1 ml of lysis buffer (10 mM Tris HCl pH 8, 150 mM NaCl, NP40 0.5% v/v, protease inhibitor 10 µl/ml). Samples were sonicated for 15 min at 4°C, then centrifuged at 1200 rpm for 15 min at 4°C. The supernatants obtained were transferred to a new microcentrifuge tube and then separated into 3 tubes. One tube was used as the input and the two other two tubes were used for the control and ORF1p IP, respectively. Each of these two tubes were then diluted to 1.5 ml with buffer (10 mM Tris HCl pH8, 150 mM NaCl, 10 µl/ml protease inhibitor) to dilute the NP40. The samples were then incubated overnight on a wheel at 4°C, then washed 3 times (first wash corresponds to post-bead samples) with 1 ml buffer (10 mM Tris HCl pH 8, 150 mM NaCl, 10 µl/ml protease inhibitor) using a magnet and then resuspended in the same buffer. The samples were boiled in Laemmli buffer (95°C, 10 min) and 20 µl of each sample were deposited on a 4-12% Nupage gel (Invitrogen, NP0336).

Mass Spectrometry

Online chromatography was performed with an RSLCnano system (Ultimate 3000, Thermo Scientific) coupled to a Q Exactive HF-X with a Nanospay Flex ion source (Thermo Scientific). Peptides were first trapped on a C18 column (75 μm inner diameter × 2 cm; nanoViper Acclaim PepMapTM 100, Thermo Scientific) with buffer A (2/98 MeCN/H2O in 0.1% formic acid) at a flow rate of 2.5 µL/min over 4 min. Separation was then performed on a 50 cm x 75 μm C18 column (nanoViper Acclaim PepMapTM RSLC, 2 μm, 100Å, Thermo Scientific) regulated to a temperature of 50°C with a linear gradient of 2% to 30% buffer B (100% MeCN in 0.1% formic acid) at a flow rate of 300 nL/min over 91 min. MS full scans were performed in the ultrahigh-field Orbitrap mass analyzer in ranges m/z 375–1500 with a resolution of 120 000 at m/z 200. The top 20 intense ions were subjected to Orbitrap for further fragmentation via high energy collision dissociation (HCD) activation and a resolution of 15 000 with the intensity threshold kept at 1.3 x 105. We selected ions with charge state from 2+ to 6+ for screening. Normalized collision energy (NCE) was set at 27 and the dynamic exclusion of 40s. For identification, the data were searched against the Mus Musculus (UP000000589_10090 012019) Uniprot database using Sequest HT through proteome discoverer (version 2.4). Enzyme specificity was set to trypsin and a maximum of two-missed cleavage sites were allowed. Oxidized methionine, Met-loss, Met-loss-Acetyl and N-terminal acetylation were set as variable modifications. Maximum allowed mass deviation was set to 10 ppm for monoisotopic precursor ions and 0.02 Da for

MS/MS peaks. The resulting files were further processed using myProMS 113 v3.10.0. FDR calculation used Percolator and was set to 1% at the peptide level for the whole study. The label free quantification was performed by peptide Extracted Ion Chromatograms (XICs), reextracted by conditions and computed with MassChroQ version 2.2.21 114. For protein quantification, XICs from proteotypic peptides shared between compared conditions (TopN matching) with missed cleavages were used. Median and scale normalization at peptide level was applied on the total signal to correct the XICs for each biological replicate (n=5). To estimate the significance of the change in protein abundance, a linear model (adjusted on peptides and biological replicates) was performed, and p-values were adjusted using the Benjamini– Hochberg FDR procedure. Proteins with at least three peptides, identified in each biological replicates of ORF1p condition, a 10-fold enrichment and an adjusted p-value ≤ 0.05 were considered significantly enriched in sample comparisons. Unique proteins were considered with at least three peptides in all replicates. Protein selected with these criteria were used for Gene Ontology enrichment analysis and string network analysis.

The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consorti-um (http://proteomecentral.proteomexchange.org) via the PRIDE partner repository 115 with the dataset identifier PXD047160.

GO term and STRING network analysis

Gene Ontology analysis was performed using GO PANTHER 116 and String network physical interactions were retrieved using the STRING database v11.5 (https://string-db.org/) and then implemented in Cytoscape software 117.

Statistical analyses

In column comparisons, data in each column were tested for normality using two normality and lognormality tests (Shapiro-Wilk test and Kolmogorov-Smirnov test). Data which passed the normality tests were analyzed subsequently by a parametric test, data which did not pass the normality tests were analyzed by a non-parametric statistical test as indicated in the figure legends. The significance threshold was defined as p<0.05 except stated otherwise. Statistical analyses were done with PRISM software (v10).